### 2024-07-04 reports/paper/vspyct.tex
first complete version of the paper

%%\documentclass[12pt,oneside,a4paper]{article}
%\documentclass[11pt,oneside]{amsart}
%\usepackage{amsmath,amssymb,amsfonts}
%%\usepackage{algorithmic}
%\usepackage{graphicx}
%%\usepackage{textcomp}
%%\usepackage{xcolor}
%%\usepackage{pdfpages}
%%\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
%%    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
%
%\usepackage{amscd}
%%\usepackage[section]{placeins}
%%\usepackage{float}
%%\usepackage[T2A,T1]{fontenc}
%%\usepackage{array}
%\usepackage[utf8]{inputenc}
%\usepackage[english]{babel}
%\usepackage{enumitem}
%\usepackage{graphicx}
%\usepackage[hidelinks]{hyperref}
%\usepackage{caption}
%%\usepackage{epstopdf}
%\usepackage{url}
%\usepackage[numbers,sort&compress]{natbib}
%
%
%\usepackage{mathtools}
%%\usepackage{cuted}
%%\usepackage{flushend}\usepackage{amsmath}
%%\usepackage{amsfonts}
%%\usepackage{amssymb}
%%\hypersetup{draft}

%\usepackage{todonotes}
%
%%\setlength{\topmargin}{-1.5cm}
%%\setlength{\textheight}{24cm}
%%\setlength{\oddsidemargin}{0cm}
%%\setlength{\textwidth}{16cm}
%
%%\usepackage{showframe}% <-- only for show page layout, in real document had to be deleted
%%\renewcommand*\ShowFrameColor{\color{red}}
%
%%\usepackage{tcolorbox}
%%\newtcolorbox{mybox}{colback=red!5!white,colframe=red!75!black}
%
%
%%
%
%%\usepackage[version=4]{mhchem}
%
%%\renewcommand{\figurename}{Fig.}
%%\usepackage[figurename=Fig.]{caption}
%%\newcommand{\eqnref}[1]{(\ref{#1})}
%
%\usepackage{subfig}
%
%\usepackage{setspace}
%\usepackage{todonotes}
%
%
%\usepackage[normalem]{ulem}
%\usepackage[nomargin,inline,marginclue,draft]{fixme}
% \fxsetup{theme=color, mode=multiuser} 
% \FXRegisterAuthor{pb}{apb}{\color{blue}PB} 
% \FXRegisterAuthor{va}{ava}{\color{red}VA} 
% \FXRegisterAuthor{bmb}{abmb}{\color{yellow}BMB} 
% \newcommand{\FXUser}[2]{\todo[inline,size=\small]{{\bfseries #1:} #2}}

%\documentclass[11pt]{amsart}
\documentclass[3p,review,authoryear]{elsarticle}
\DeclareUnicodeCharacter{0308}{\forall}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{a4paper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{adjustbox}
\usepackage{amssymb}
\usepackage{rotating}
\usepackage{hyperref}
\usepackage{booktabs}
\usepackage{subfig}
\usepackage{engord}
%\usepackage{amsaddr}
\usepackage{url}
\usepackage{mathtools}
\usepackage{amsmath}
\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator*{\argmin}{arg\,min}
%\usepackage[sorting=nyt,firstinits=true]{biblatex}
%\usepackage[backend=biber,sorting=none,style=numeric-comp,urldate=long,maxcitepnames=10,maxbibnames=9]{biblatex}
%\addbibresource{bibliography.bib}
\usepackage{todonotes}
\usepackage[normalem]{ulem}
\newcommand{\mpop}[2]{{\color{orange}\sout{#1}\color{purple}{#2}}}
\usepackage[nomargin,inline,marginclue,draft]{fixme}
 \fxsetup{theme=color, mode=multiuser} 
 \FXRegisterAuthor{pb}{apb}{\color{blue}PB} 
 \FXRegisterAuthor{va}{avs}{\color{red}VA} 
 \FXRegisterAuthor{bmb}{abmb}{\color{blue}BMB}
 %\FXRegisterAuthor{ro}{aro}{\color{red}R1}
 %\FXRegisterAuthor{rt}{art}{\color{blue}R2}
 \newcommand{\FXUser}[2]{\todo[inline,size=\small]{{\bfseries #1:} #2}}

%
\newcommand{\mkom}[1]{{\color{purple} #1}}
%\newcommand{\mnew}[1]{{\color{Tan} #1}}
%\newcommand{\mpop}[2]{{\color{orange}\sout{#1}\color{purple}{#2}}}

\usepackage{todonotes}
\usepackage{xargs} 
\usepackage{tabularx}
\newcommandx{\pbtodo}[2][1=]{\todo[linecolor=red,backgroundcolor=red!25,bordercolor=red,#1]{\tiny PB:  #2}}

\newcommandx{\vatodo}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!25,bordercolor=blue,#1]{\tiny VA: #2}}

\newcommandx{\bmbtodo}[2][1=]{\todo[linecolor=blue,backgroundcolor=blue!55,bordercolor=blue,#1]{\tiny BMB: #2}}

\usepackage[commandnameprefix=ifneeded]{changes}

\setcommentmarkup{\linepenalty=1000\todo[color=authorcolor!10, bordercolor=authorcolor, linecolor=authorcolor!70, nolist]%
	{\textbf{[#3\penalty1000]} \hskip 0pt plus1fill\relax\penalty50\hskip 0pt plus-1fill\relax#1}%
}

\definechangesauthor[name=r1, color=brown]{r1}
\definechangesauthor[name=r2, color=orange]{r2}

\usepackage{verbatim}
\usepackage{arydshln}

\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{pythontex}

\usepackage[acronym]{glossaries}
\newacronym{ml}{ML}{Machine Learning}
\newacronym{bs}{BS}{Brier Score}
\newacronym{pes}{PES}{Public Employment Services}
\newacronym{pex}{PEX}{probability of exit}
\newacronym{spyct}{SPYCT}{oblique predictive clustering tree}
\newacronym{vspyct}{VSPYCT}{variational oblique predictive clustering tree}
\newacronym{sslmtr}{SSL-MTR}{semi-supervised multi-target regression}
\newacronym{pct}{PCT}{predictive clustering tree}
\newacronym{sop}{SOP}{structured output prediction}
\newacronym{fp}{FP}{False Positive}
\newacronym{fn}{FN}{False Negative}
\newacronym{tpr}{TPR}{True Positive Rate}
\newacronym{fpr}{FPR}{False Positive Rate}
\newacronym{auc}{AUC}{Area under the ROC curve}
\newacronym{tp}{TP}{True Positives}
\newacronym{tn}{TN}{True Negatives}
\newacronym{ecdf}{ECDF}{Empirical Cumulative Distribution Function}
\newacronym{isce}{ISCED}{International Standard Classification of Education}

\graphicspath{ {./}{../figures/} }

\begin{document}
\begin{frontmatter}
	\title{Variational oblique predictive clustering trees for multi-target regression}
	\author[1,2]{Viktor Andonovikj\corref{cor1}}
	\ead{viktor.andonovikj@ijs.si}
	\author[1]{Sašo Džeroski}
	\ead{saso.dzeroski@ijs.si}
	\author[1,3]{Biljana Mileva Boshkoska}
	\ead{biljana.mileva@ijs.si}
	\author[1]{Pavle Boškoski}
	\ead{pavle.boskoski@ijs.si}
	\address[1]{Jožef Stefan Institute, Jamova cesta 39, 1000 Ljubljana, Slovenia}
	\address[2]{Jožef Stefan International Postgraduate School, Jamova cesta 39, 1000 Ljubljana, Slovenia}
	\address[3]{Faculty of Information Studies in Novo mesto, Ljubljanska cesta 31b, 8000 Novo mesto, Slovenia}
	\cortext[cor1]{Corresponding author.} 
\begin{abstract}

\Glspl{spyct} are semi-supervised multi-target prediction models mainly used for \gls{sop} problems.
They are computationally efficient and when combined in ensembles they achieve state-of-the-art results.
However, one major issue is that it's challenging to interpret an ensemble of \glspl{spyct} without the use of a model-agnostic method.
We propose variational oblique predictive clustering trees, which address this challenge.
The parameters of each split node are treated as distributions and they are learned through the Variational Bayes method. We evaluate the model on several benchmark datasets of different sizes.
The experimental analyses show that a single \gls{vspyct} achieves state-of-the-art or on-par performance with ensemble of standard \glspl{spyct}. We also present a method for extracting feature importance scores from the model.
Finally, we present a method to visually interpret the model's decision making process through analysis of the relative feature importance in each split node.
\end{abstract}
%\maketitle
\end{frontmatter}
\glsresetall

\section{Introduction}
%PARAGRAF ZA ENSEMBLE MODEL VS BAYESIAN MODEL

In the evolving field of machine learning, the fusion of semi-supervised learning and decision tree models has led to the development of models like \glspl{pct}~\citep{Kocev_2013} and their advanced variant, \glspl{spyct}~\citep{Stepi_nik_2021}.
These models leverage both labeled and unlabelled data to enhance prediction accuracy and efficiency.
\Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal performance.
While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, crucial for informed decision-making across various applications.

To address these limitations, we introduce the \gls{vspyct} model.
\gls{vspyct} integrates the variational Bayes method~\citep{Blei_2017} to facilitate complex decision-making within a singular model framework, thus maintaining the interpretability inherent to decision trees without necessitating ensembles.
Furthermore, by embedding model-specific uncertainty quantification directly into the decision tree structure through Bayesian inference, \gls{vspyct} enhances the depth of insight into the model's decision process and confidence levels.
%This approach not only elevates the explainability of semi-supervised decision trees by consolidating predictive capabilities into a single model but also marks a significant advancement in machine learning, particularly within semi-supervised learning domains.
By combining the robustness of \gls{spyct} ensembles with the clarity and interpretability of a single tree model, alongside introducing uncertainty quantification, our research provides a comprehensive tool aimed to overcome current limitations of predictive clustering methodologies.

\begin{figure}[h!]
    \centering
    \includegraphics[width=1.0\textwidth]{main_flowchart.pdf}
    \caption{Ensemble of trees vs. Variational tree}
    \label{fig:main_flowchart}
\end{figure}

Our proposed \gls{vspyct} framework diverges from traditional ensemble approaches by directly incorporating Bayesian principles into oblique decision trees' architecture.
This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.
The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, modelling these parameters as random variables with Gaussian distributions.
This represents a paradigm shift toward a probabilistic understanding of decision-making processes within tree-based models, significantly enhancing interpretability, especially in uncertainty estimation domains.
The whole framework is given in Figure~\ref{fig:flowchart}

\begin{figure}[h!]
    \centering
    \includegraphics[width=1.0\textwidth]{methodology_flowchart.pdf}
    \caption{Flowchart of the methodology}
    \label{fig:flowchart}
\end{figure}

It sets a new benchmark for interpretable, efficient, and reliable machine learning models.

%citiraj community detection (nasiot trud), rise of gen-ai/black-box ai/llms (llama, bert, gpt), post-hoc explainability (shap),  vb metodot,

\section{Related work}

%references for VB applications in ML and Survival analysis models. Cite pavles paper.

\Glspl{pct} have been an essential development in extending decision tree capabilities to a variety of predictive modelling tasks, including structured output prediction.
\Glspl{pct} are versatile and can be combined into ensembles to achieve state-of-the-art performance~\citep{Kocev_2013}.
However, as the dimensionality of the output space increases, the learning time of \glspl{pct} scales poorly, posing challenges in tasks like hierarchical multi-label classification where outputs can consist of hundreds of potential labels.

\Glspl{pct} are celebrated for their interpretability and efficiency, yet they often fall short in predictive performance due to their myopic, greedy nature of selecting splits.
The introduction of option predictive clustering trees (OPCTs)~\citep{Stepisnik_2020} aimed to address this limitation by incorporating multiple alternative splits within option nodes.
This approach mitigates the inherent myopia and achieves competitive performance with ensemble methods like bagging and random forests, while still maintaining a degree of interpretability.
OPCTs, despite mitigating myopia and enhancing performance, still face challenges in maintaining interpretability when extended to complex tasks.

On the other hand, \glspl{spyct}~\citep{Stepi_nik_2021} utilise oblique splits that incorporate linear combinations of features, allowing splits to correspond to arbitrary hyperplanes in the input space.
This makes \glspl{spyct} highly efficient for high-dimensional data and capable of leveraging data sparsity.
Experimental evaluations on numerous benchmark datasets have shown that \glspl{spyct} achieve performance on par with state-of-the-art methods while being significantly faster than standard \glspl{pct}~\citep{Andonovikj_2024}.
%Moreover, meaningful feature importance scores can be extracted from SPYCT models, enhancing their interpretability.

Despite these advancements, ensemble models such as random forests and gradient boosting machines remain popular due to their superior predictive accuracy.
However, their complexity often obscures interpretability, making it challenging for users to understand the model’s decision-making process.
Efforts to bridge this gap include the iForest visual analytics system~\citep{Zhao_2019}, which summarises decision paths and tree structures within random forests, thereby making the ensemble's decisions more transparent.
Additionally, the Tree Space Prototypes approach~\citep{Tan_2020} uses representative points (prototypes) for each class, offering a more intuitive understanding of ensemble classifiers compared to traditional feature-based explanations.
Visual analytics systems like iForest and prototype-based approaches offer some clarity but do not fully resolve the complexity issue.

In critical applications, such as medical diagnosis, the need for transparent and trustworthy predictions is paramount.
Methods for explaining classifier predictions and estimating the reliability of regression predictions, as demonstrated in breast cancer recurrence prediction, provide users with additional insights and build trust by clarifying the decision-making process.
Similarly, the MAPLE~\citep{plumb2018model} model combines local linear modelling with a dual interpretation of random forests, offering faithful self-explanations and maintaining high predictive accuracy.
MAPLE addresses the accuracy-interpretability trade-off and provides both example-based and local explanations, making it a comprehensive tool for understanding model behaviour.
Methods like MAPLE and other interpretability frameworks strike a balance between accuracy and transparency, but they often fall short in handling high-dimensional, sparse data efficiently.

While significant progress has been made in improving decision tree models and their ensembles, several weaknesses remain.
Many of these approaches lack inherent mechanisms for uncertainty quantification, which is crucial for applications requiring high reliability and decision certainty.

Our proposed \glspl{vspyct} model addresses these critical challenges by integrating Bayesian principles directly into the decision tree framework.
%Unlike traditional ensemble strategies, \glspl{vspyct} employs a probabilistic treatment of model parameters through the variational Bayes method, which enhances interpretability and provides inherent uncertainty quantification.
This approach maintains the clarity and simplicity of a single tree model while achieving state-of-the-art performance, thus offering a significant improvement over existing methods.
By introducing uncertainty quantification directly into the decision-making process, \glspl{vspyct} enhances the reliability and applicability of the model across various domains where decision certainty is crucial.
%This advancement bridges the gap between theoretical developments and practical applications, marking a significant step forward in making complex decision trees more accessible, interpretable, and effective.

\section{Preliminaries}

\subsection{Variational Bayes method}

% optional slika - Optimisation process of finding the closest variational distribution qω(θ) over the set of latent variables ω. (da se generira vo python)

In conducting a stochastic analysis, like survival analysis, the process of inferring model parameters hinges on applying Bayes' theorem.
For observations $x$ produced by a system defined by parameters $\theta$, Bayes' theorem is given as:

\begin{equation}
p(\theta|x)=\frac{p(x|\theta)p(\theta)}{p(x)}
\label{eq:bayes_rule}
\end{equation}

Here, the model-prescribed likelihood $p(x|\theta)$ and the selected prior $p(\theta)$ are generally well-defined.
The primary computational challenge arises in determining the posterior distribution $p(\theta|x)$ due to the denominator $p(x)$, which is derived as:

\begin{equation}
p(x)=\int_{\theta} p(x|\theta)p(\theta) \,dx,
\label{eq:evidence}
\end{equation}

This integral rarely yields a closed-form solution.
For cases with multiple dimensions, the computational burden often renders Monte Carlo methods infeasible.
The VB approach offers an approximation to this issue by closely mimicking the true posterior $p(\theta|x)$ with an approximate distribution $q_{\omega^*}(\theta)$. This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterized by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
Typically, the mean-field variational family is utilized, assuming independence among the parameters $\omega$. To find the optimal $\omega^*$, one minimises the Kullback–Leibler (KL) divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:

\begin{equation}
\omega^* = \argmin_{\omega \in \Omega}KL(q_\omega(\theta) \parallel p(\theta|x))
\label{eq:optim}
\end{equation}

Given the unknown nature of the true posterior, a rearrangement is needed to compute the KL divergence, simplifying to:

\begin{align*}
KL(q_\omega(\theta) \parallel p(\theta|x)) &= \mathbb{E}_q\left[\log \frac{q_\omega (\theta)}{p(\theta|x)}\right] \\
&=\mathbb{E}_q\left[\log q_\omega(\theta)\right]-\mathbb{E}_q\left[\log p(\theta|x)\right]\\
&=\mathbb{E}_q\left[\log q_\omega(\theta)\right]-\mathbb{E}_q\left[\log p(x,\theta)\right]\\
&=\mathbb{E}_q[\log q_\omega(\theta) - \log p(x,\theta)+\log p(x)]\\
&=-\mathbb{E}_q\left[\log q_\omega(\theta) - \log q_\omega(\theta)\right] + \log p(x)\\
&\quad \quad \underbrace{\phantom{\mathbb{E}_q\left[\log q_\omega(\theta) - \log q_\omega(\theta)\right]}}_{\text{ELBO}}
\label{eq:kl_divergence}
\end{align*}

Maximising the ELBO inversely minimises the KL divergence, with $\log p(x)$ remaining constant and excluded from the optimisation.
Typically, the ELBO maximisation criterion is non-convex, with no assurance of global extremum convergence by optimisation algorithms.
Various algorithms have been employed for this challenge, including coordinate ascent variational inference and stochastic variational inference~\cite{hoffman2013stochastic}, with the latter's gradient calculated as per the black-box variational inference~\cite{sashank2018convergence}.
The adoption of an approximate variational distribution introduces bias contingent on the chosen variational family $\mathcal{Q}$, necessitating a decision grounded in empirical evidence or expert knowledge.
Despite this bias, VB's computational efficiency significantly surpasses traditional methods like Monte Carlo integration.
In our work, the ADAM optimiser~\citep{kingma2014adam} is used to facilitate the ELBO optimisation.

\subsection{Semi-supervised Multi-target Regression in Survival Analysis}

Survival analysis primarily focuses on time-to-event data, where not all events are fully observed due to some form of censoring that halts observation before the event transpires.
The most prevalent form of censoring in this context is right-censoring, which occurs when there is incomplete knowledge about the survival time at the end of the follow-up period.
As a result, each survival record is characterised by three elements: the actual survival time $T$, its observed component $t$, and a binary indicator $\delta$, which is defined as:
\begin{equation}
\delta = \begin{cases}
0, & \text{if } T>t \text{ indicating a right-censored record},\\
1, & \text{if } T=t \text{, denoting that the event has been observed and the record is not censored}.
\end{cases}
\label{eq:delta}
\end{equation}

Moreover, assume that each observation is described by a set of covariates or features $\mathbf{X}_i=\{x_{i_1},\ldots,x_{i_N}\}$.
Thus, each record $i$ comprises the variables $\mathbf{X}_i$, $t_i$, and $\delta_i$ as specified in~\eqref{eq:delta}.
If $\delta_k=1$, the event is confirmed to have occurred at time $t_k$ and $y_{k_j} = 0$ for all $j \geq{k}$.
Conversely, if $\delta_k=0$, it implies that the fate of the instance beyond time $t_k$ is unknown, making $y_{k_j}$ missing for $j \geq{t_k}$.
This scenario of censored examples introduces missing data, which makes such datasets apt for semi-supervised learning methodologies."

\section{Methodology}

As a main part of the \gls{vspyct} model lies the oblique tree structure, which is a form of a \gls{pct} where splits are made using linear combinations of the input features rather than single features.
This allows for more complex decision boundaries, capturing intricate patterns in the data with fewer nodes compared to axis-aligned trees.
However, unlike traditional oblique trees that fix the parameters of these linear combinations, \gls{vspyct} introduces a probabilistic twist: the parameters of the splits are not single fixed values but are instead modelled as random variables with Gaussian distributions.
This probabilistic representation captures the uncertainty in the model parameters, a feature that is particularly useful in scenarios with noisy data.
The construction of the \gls{vspyct} begins with a root node and iteratively expands by adding internal nodes or leaves through a learning process based on the variational Bayes method.
Each node represents a binary decision point where instances are split into two child nodes.
At each internal node, a variational Bayes approach is used to learn the parameters of an oblique split.
This involves optimising the parameters of a linear model that best divides the data to minimise the ELBO.
The linear model is defined as:

\begin{equation}
f(\mathbf{x})=\sigma \left(\mathbf{w}^T \mathbf{x} + b \right)
\label{eq:linear_model}
\end{equation}

where $\mathbf{x} \in \mathbb{R}^N$ is the feature vector, $\mathbf{w} \in \mathbb{R}^N$ represents the weights, $b \in \mathbb{R}$ is the bias, and $\sigma(\cdot)$ denotes the sigmoid function ensuring the output lies in $(0, 1)$, indicating the probability of branching to the right child node.
The learning process involves defining a prior distribution over the weights and biases, typically Gaussian, and updating this to a posterior distribution given the observed data through variational inference.
The ELBO is optimised using stochastic gradient descent, with the learning rate $\alpha$ and potentially variable batch sizes $B$.
The split at each node is determined by evaluating the learned linear model on the descriptive data.
Instances are split based on the probability threshold of $0.5$, with those having a probability greater than $0.5$ moving to the right child node and the rest to the left.
For leaf nodes, a prototype is computed as the mean of the target variables for the instances assigned to the leaf.
This prototype represents the prediction for instances that fall into the leaf during the prediction phase.

\subsection{Learning splits using Variational Bayes}

% optional: computation graph diagram

The learning of these Gaussian-distributed parameters is achieved through a Variational Bayes (VB) method.
In the context of \gls{vspyct}, the VB method is employed to learn the Gaussian distributions representing the parameters of the oblique splits.
The process involves defining a prior distribution over the parameters and then updating this prior to a posterior distribution given the observed data.
This is done by optimising the ELBO, which balances the fit of the model to the data with the complexity of the model.
The procedure for learning a split using the VB method is given in Algorithm~\ref{alg:learn_split}.

\begin{algorithm}[h!]
\caption{Variational Learning of Split Parameters in \gls{vspyct}}
\label{alg:learn_split}
\begin{algorithmic}[1]

\State \textbf{Input:} Dataset $\mathcal{D}$, features $\mathbf{X}$, targets $\mathbf{Y}$, parameters $\theta$, epochs $E$, batch size $\beta$, learning rate $\lambda$, selection probability $\sigma$
\State \textbf{Output:} Parameters $\Theta$, guide $\mathcal{G}$

\Procedure{LearnSplit}{$\mathcal{D}$, $\mathbf{X}$, $\mathbf{Y}$, $\theta$, $E$, $\lambda$, $\beta$, $\sigma$}
    \State $\Phi \sim \text{Bernoulli}(\sigma)$ \Comment{Select attributes based on probability $\sigma$}
    \State $\mathbf{X}_\Phi \leftarrow$ subset of $\mathbf{X}$ using $\Phi$
    \State Model $\mathcal{M} \leftarrow \text{Define Impurity}(\dim(\mathbf{X}_\Phi), 1, \theta)$
    \State Guide $\mathcal{G} \leftarrow \text{Define AutoDiagonalNormal}(\mathcal{M})$
    \State Define optimization parameters $\alpha \leftarrow \{\text{"lr"}: \lambda\}$
    \State Initialise optimizer $\text{Optimizer} \leftarrow \text{Adam}(\alpha)$
    \State Early stopping mechanism $\text{EarlyStopping} \leftarrow \text{Initialize}(\beta)$

    \State Initialise loss list $L$
    \State Calculate number of batches $B \leftarrow \lceil \frac{|\mathcal{D}|}{\beta} \rceil$

    \For{$e \in \{1, \dots, E\}$}
        \State Initialise training loss $\mathcal{L}_{\text{train}} \leftarrow 0$
        \For{$b \in \{1, \dots, B\}$}
            \State $(\mathbf{X}_b, \mathbf{Y}_b) \leftarrow \text{Get Batch}(\mathbf{X}_\Phi, \mathbf{Y}, b, \beta)$
            \State Update training loss $\mathcal{L}_{\text{train}} \leftarrow \mathcal{L}_{\text{train}} + \text{SVI Step}(\mathcal{M}, \mathcal{G}, \mathbf{X}_b, \mathbf{Y}_b)$
        \EndFor
        \State Append $\mathcal{L}_{\text{train}}$ to $L$
        \If{$\text{EarlyStopping}(\mathcal{L}_{\text{train}})$}
            \State \textbf{break}
        \EndIf
    \EndFor
    \State \Return $\mathcal{M}$, $\mathcal{G}$
\EndProcedure

\end{algorithmic}
\end{algorithm}

\begin{algorithm}[h!]
\caption{Impurity}
\begin{algorithmic}[1]

\Procedure{Impurity}{$\xi$, $\zeta$, $\theta$}
    \State $\mathbf{w}, b \sim \mathcal{N}(0, I)$ \Comment{Weight and bias priors}
    \State $\rho(\mathbf{x}) \gets \sigma(\mathbf{w}^\top \mathbf{x} + b)$ \Comment{Right selection probability}
    \State $\lambda(\mathbf{x}) \gets 1 - \rho(\mathbf{x})$ \Comment{Left selection probability}
    \State Compute $\text{Var}_\rho(\mathbf{Y})$ and $\text{Var}_\lambda(\mathbf{Y})$ \Comment{Weighted variances}
    \State $\Omega \gets \lambda \cdot \text{Var}_\lambda(\mathbf{Y}) + \rho \cdot \text{Var}_\rho(\mathbf{Y}) + \|\mathbf{w}\|_{0.5}$
%    \State Observe $\Omega$ with likelihood $\mathcal{N}(0, 1)$
    \State \textbf{return} $\Omega$
\EndProcedure

\end{algorithmic}
\end{algorithm}

\subsection{Making a prediction}

The procedure for making a prediction for a new instance using the \gls{vspyct} model is given in Algorithm~\ref{alg:make_pred}.
It involves traversing the tree from the root to a leaf, at each internal node using the variational posterior to sample parameters and compute the probability of branching right or left.
The prediction of the reached leaf's prototype is returned as the output.
Operationally, when a new data point is to be classified or a prediction is to be made, it traverses the tree from the root to a leaf.
At each node, the oblique split is determined by evaluating the linear combination of the input features, with parameters sampled from the learned Gaussian distributions.
This probabilistic approach to decision-making within the tree allows for a more nuanced interpretation of the data, accounting for uncertainty in the model parameters.

\begin{algorithm}
\caption{\gls{vspyct} Prediction Process}
\label{alg:make_pred}
\begin{algorithmic}[1]

\State \textbf{Input:} Feature vector $\mathbf{x}$, decision tree $T$ rooted at node $\Omega$
\State \textbf{Output:} Prediction $\hat{y}$

\Function{Predict}{$\Omega$, $\mathbf{x}$}
    \If{$\Omega$ is a leaf}
        \State \Return $\Omega.prototype$
    \Else
        \State $\Theta \gets$ \Call{SampleParameters}{$\Omega.split\_model$, $\Omega.guide$}
        \State Define sigmoid function $\rho(\Theta, \mathbf{x}) = \frac{1}{1 + \exp(-\Theta^\top \mathbf{x})}$
        \If{$\rho(\Theta, \mathbf{x}) \leq 0.5$}
            \State \Return \Call{Predict}{$\Omega.left$, $\mathbf{x}$}
        \Else
            \State \Return \Call{Predict}{$\Omega.right$, $\mathbf{x}$}
        \EndIf
    \EndIf
\EndFunction

\Function{SampleParameters}{model, guide}
    \State Let $\Theta = (\mathbf{w}, b)$ where:
    \State $\mathbf{w}$ are weights sampled from variational posterior $q_{\omega}(\mathbf{w}|\mathcal{D})$
    \State $b$ is bias sampled from variational posterior $q_{\omega}(b|\mathcal{D})$
    \State \Return $\Theta$
\EndFunction

\State Compute prediction $\hat{y}$ by invoking \Call{Predict}{$\Omega$, $\mathbf{x}$}
\State \textbf{Return} $\hat{y}$

\end{algorithmic}
\end{algorithm}

\subsection{Time complexity analysis}

In this section, we analyse the time complexity of learning a split in \gls{vspyct} and compare it with \gls{spyct}.
Our analysis focuses on the computational expense associated with determining a split, as this represents the principal distinction between the two approaches.
In the predictive clustering framework, the set of clustering attributes $K$, which contributes to the variance calculation, encompasses both the features and the target variables.
As a result, $K$ equals $N + T$, with $N$ representing the feature count and $T$ denoting the target count.
The time complexity for learning in \glspl{spyct} can be expressed as $\mathcal{O}(NI_o(D+K))$, wherein $I_o$ stands for the iteration count needed for optimisation.
Notably, the computational demand for the \glspl{spyct}' gradient-based version increases linearly with the attribute count.

The introduction of variational Bayes method in the architecture of \gls{spyct} to create \gls{vspyct} adds a new layer of computational complexity.
Specifically, the variational approach requires iterating over the dataset multiple times to optimise the ELBO for each split.
This iterative process involves sampling from the posterior distributions of split parameters, evaluating the ELBO, and updating the parameters accordingly.
Consequently, the time complexity for learning a split in \gls{vspyct} can be described as $\mathcal{O}(MI_oNI_{vb})$, where $M$ is the number of Monte Carlo samples used to approximate the posterior distributions, $I_{vb}$ represents the number of iterations required for variational inference to converge, and the other variables retain their meanings from the \gls{spyct} context.

Comparing \glspl{vspyct} to \glspl{spyct}, the introduction of variational Bayes method increases the computational load due to the necessity of Monte Carlo sampling and iterative ELBO optimisation.
However, this additional complexity is counterbalanced by the enhanced predictive performance, model interpretability, and uncertainty quantification offered by \glspl{vspyct}.
%While \glspl{spyct} are computationally efficient due to their linear scaling with the number of attributes and utilisation of gradient-based optimisation, \glspl{vspyct} leverage the power of variational inference to provide a deeper understanding of the data and model behaviour, even if at the cost of increased time complexity.

The choice between using \glspl{vspyct} over \glspl{spyct} hinges on the specific requirements of the application scenario.
For tasks where model interpretability and uncertainty quantification are paramount, and computational resources are sufficient, \glspl{vspyct} offer a compelling advantage.
Conversely, for applications where computational efficiency is critical, and the interpretability requirements are less stringent, \glspl{spyct} may still be favourable.

\subsection{Feature importance}

Feature importance is derived from the aggregated impact of features across all splits, quantified by examining the variational posterior distributions of the weights.
The importance of each feature is proportional to its contribution to reducing impurity across splits.

Given a \gls{vspyct} model, the feature importances are derived from the variational parameters defining the split decisions across the tree.
The essence of the computation lies in evaluating the expected influence of each feature, taking into account the probabilistic nature of the model parameters.
This is encapsulated by the following formula:

\begin{equation}
\text{Imp}(T) = \sum_{s \in T} \left( \frac{s_n}{N} \right) \left( \frac{\mathbb{E}[\mathbf{w}_s]}{|\mathbb{E}[\mathbf{w}_s]|_1} \right)
\end{equation}

where $T$ represents the variational oblique predictive clustering tree, $s$ iterates over the set of split nodes within $T$, $s_n$ is the number of training examples that passed through node $s$, $N$ denotes the total number of training examples used in the tree, $\mathbf{w}_s$ symbolizes the weight vector defining the oblique split at node $s$, treated as a random variable with a Gaussian distribution, $\mathbb{E}[\mathbf{w}_s]$ signifies the expected value (mean) of the weight vector $\mathbf{w}_s$, calculated from its posterior distribution as inferred by the variational Bayesian method.
The normalization factor $|\mathbb{E}[\mathbf{w}_s]|_1$ ensures the comparability of feature importances by scaling them relative to the aggregate magnitude of expected weights.
This formulation accounts for the stochastic nature of each parameter, reflecting the central tendency of the distributions defining the split hyperplanes.

For practical computation, the expected weights $\mathbb{E}[\mathbf{w}_s]$ are obtained by sampling from the posterior distributions inferred during the learning process.
Each weight vector $\mathbf{w}_s$ is drawn such that $\mathbf{w}_s \sim q(\mathbf{w}_s|\mathcal{D})$, where $q(\mathbf{w}_s|\mathcal{D})$ represents the variational approximation to the posterior distribution of weights given the data $\mathcal{D}$.
The importance of each feature is then a function of these expected weights, modulated by the frequency of training instances influencing each split, thereby emphasising the splits that impact a larger portion of the data.
This mathematical framework for computing feature importances in \gls{vspyct} not only highlights the contribution of individual features to the model's decision-making process but also integrates the inherent uncertainty of these contributions, offering a richer, more nuanced understanding of feature relevance within a probabilistic modelling context.

\section{Data and evaluation metric}

\subsection{Data}

\subsection{R Survival datasets}

Our evaluation encompasses four benchmark survival analysis datasets, each serving as a standard reference within the domain and accessible via the survival R package~\citep{survival-package}.

\begin{enumerate}
\item \textit{Retinopathy Dataset}: Investigates the efficacy of laser coagulation in delaying vision loss due to diabetic retinopathy among 197 patients, with survival times adjusted for a minimum event horizon of 6.5 months.
\item \textit{Lung Dataset}: Analyses survival data for patients with advanced lung cancer, focusing on their lifespan post-diagnosis and daily activity performance levels.
\item \textit{Primary Biliary Cirrhosis (PBC) Dataset}: Examines the progression of primary biliary cholangitis in 424 patients, with data drawn from a clinical trial conducted between 1974 and 1984, exploring disease progression towards cirrhosis.
\item \textit{Heart Dataset}: Studies the survival of patients on the waiting list for heart transplants at Stanford, evaluating the length of time patients remain on the waiting list.
\end{enumerate}

\begin{table}[h!]
\centering
\caption{Main properties of the datasets. Censoring shows the percentage of censored observations.}
\adjustbox{max width=\textwidth}{
\begin{tabular}{llll}
\toprule
Dataset& \# of examples& \# of features& Censoring\\
\midrule
Mayo Clinic Primary Biliary Cholangitis&312&17&60\%\\
NCCTG Lung Cancer&228&8&\\
Stanford Heart Transplant&172&4&\\
Retinopathy&394&9&\\
\bottomrule
\end{tabular}
}
\label{table:dataset}
\end{table}%

\subsection{Unemployment dataset}
The dataset we consider consists of $74,086$ instances and contains anonymised personal and professional characteristics of jobseekers that have entered the Slovenian \gls{pes}.
The dataset is complex because its attributes come in different forms (categorical, numerical, date and time), and most have to undergo transformation.
An overview of the dataset and its features is given in \tablename~\ref{table:dataset}.

\begin{table}[h!]
\centering
\caption{Features of the dataset and their cardinalities \citep{Andonovikj_2024}}
\begin{tabular}{ll}
\toprule
Continuous features& Discrete features\\
\midrule
Day of PES entry&Education direction (30)\\
Months of work experience&Occupation (ISCO) (123)\\
Age&Reason for PES entry (5)\\
Month of PES entry&Reason for termination (43)\\
&Social benefits (2)\\
&Gender (2)\\
&eApplication (2)\\
&Unemployment benefits (2)\\
\bottomrule
\end{tabular}
\label{table:dataset}
\end{table}%

The general structure of the jobseekers' attributes is described by dividing the attributes into several prominent groups: socioeconomic variables (gender, age), information on job readiness (education, health limitations), and opportunities (regional labour market development), and all available labour market history information, such as prior work experience.
The target variable is in numeric form.
It is a counter of days till a jobseeker gets employed or gets out of the study and leaves no further information about the jobseeker's employment status.

\figurename~\ref{fig:grid_prob} shows the distributions of some features in the dataset.
The sample of jobseekers is balanced in terms of gender.
Most jobseekers enter the \gls{pes} without any working experience. 
As expected, the age distribution is between 18 and 64 since this is the legal age interval of the labour-active population.
It should be noted that the age distribution is bi-modal.
The first peak is around 23-26 years since this is the expected graduation age for a university degree.
The second peak is just before 60 years of age, i.e. jobseekers who became unemployed just before retirement due to various reasons.
The final histogram shows that the dataset contains 30\% of right censored data.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.95\textwidth]{feature_distributions.pdf}
    \caption{Distributions of some of the dataset features \citep{Andonovikj_2024}}
    \label{fig:grid_prob}
\end{figure}

The categorical features are diverse with respect to their cardinality.
The categories of the features with high cardinality ($>1000$ different class values per feature) are transformed using domain knowledge and the number of class values is significantly reduced.
Because the \emph{ISCO} standard and \emph{Education direction} provide a hierarchical classification of occupations and education directions, we took the higher-level classifications and thus significantly reduced the number of categories.
The remaining categorical features were encoded using one-hot encoding.
The temporal (cyclical) features, such as ``Day of PES entry" and ``Month of PES entry" were transformed by calculating the feature's $\sin$ and $\cos$ components.
In this way, the artificially significant difference between, for example, the last day of the previous month and the first day of the current month is removed, and the value of the feature can be viewed as a point $\left(\sin(x), \cos(x)\right)$ of a circle.

\subsection{Evaluation metric}

The efficacy of survival models is predominantly assessed through the Concordance Index~\citep{brentnall2018use} and the Brier score~\citep{gerds2006consistent}.
While the Concordance Index offers valuable insights, its inability to account for the error magnitude in predicted probabilities limits its utility.
The Brier score addresses this gap by factoring in the error magnitude, making it a more comprehensive metric for our experimental evaluation.
The Brier score assesses the precision of a forecasted survival function at a specific time $t$, quantifying the mean squared differences between observed survival statuses and forecasted survival probabilities.
It ranges from $0$ to $1$, where a score of $0$ indicates perfect accuracy.
This metric not only assesses the precision of time-to-event forecasts, akin to a squared error loss, but also necessitates adjustments for right-censored samples in the dataset.
This adjustment employs the inverse probability of censoring weights method to weight the squared differences.

The conditional survival function estimator for censoring times, $\hat{G}(t)=P(C>t)$, utilises the Kaplan-Meier method, with $C$ denoting the censoring time.
The Brier score is thus calculated as~\citep{trickbrier}:

\begin{equation}
BS(t) = \frac{1}{M}\sum\limits_{i=1}^N \left(\frac{\hat{S}(t_i,\mathbf{X}i)\cdot1{T_i\geq{t},\delta_i=1}}{\hat{G}({T_i}^-)}+\frac{(1-\hat{S}(t_i,\mathbf{X}i))\cdot1{T_i>t, \delta_i=0}}{\hat{G}(T_i)}\right)
\label{eq:bs}
\end{equation}

Here, $M$ signifies the count of test instances, $\mathbf{X}_i$ the feature vector for each instance, $\hat{S}(t, \mathbf{X}_i)$ the predicted survival probability at time $t_i$, $\delta_i$ a boolean indicator, and $T_i$ and $T_i^-$ the survival times for $\delta_i$ being 0 and 1, respectively. The Brier score is computed at various timestamps to evaluate how the method's performance varies with predictions extended into the future.

The performance was estimated by using $k$-fold cross-validation with the parameter $k$ set to 5~\citep{elstatl}.
The prediction model yields an estimated survival curve for each individual in the study.

\section{Results}

The resulting BS curves are shown in Figure~\ref{fig:brier_scores}.
It should be noted that lower BS indicates a more accurate forecast.
The analysis revealed distinct performance patterns between the two models across different datasets.
Specifically, in the PBC and lung cancer datasets, \gls{vspyct} outperformed \gls{spyct}, evidenced by consistently lower Brier scores throughout the observation period.
This superior performance of \gls{vspyct} could be attributed to its robust handling of parameter uncertainty through the VB method, which likely improved its predictive accuracy in these contexts.
The \gls{vspyct} model demonstrated a particularly notable advantage in maintaining lower peak Brier scores, suggesting enhanced reliability in predicting survival outcomes for patients with PBC and lung cancer.

Conversely, in the heart transplant and retinopathy datasets, \gls{spyct} showed better performance compared to \gls{vspyct}, as indicated by lower Brier scores over time.
This suggests that the simpler, more direct approach of \gls{spyct} may be more effective in certain contexts, possibly due to less complex data structures or the nature of the underlying survival processes in these specific conditions.
The lower Brier scores achieved by \gls{spyct} in these cases highlight its efficiency in accurately predicting the survival outcomes of heart transplant patients and those undergoing treatment for diabetic retinopathy.

% average rank diagram - zemi go kodot od Matej Petkovic

\begin{figure}[h!]
\centering
\subfloat[Lung]{%
\includegraphics{bs_r_cancer.pdf}
\label{fig:bs_cancer}}
\quad % Adds some space between the figures in the same row
\subfloat[Retinopathy]{%
\includegraphics{bs_r_retinopathy.pdf}
\label{fig:bs_retinopathy}}\\ % Ends the first row and starts a new one
\subfloat[Heart]{%
\includegraphics{bs_r_heart_truncated.pdf}
\label{fig:bs_heart}}
\quad % Adds some space between the figures in the same row
\subfloat[PBC]{%
\includegraphics{bs_r_pbc.pdf}
\label{fig:bs_pbc}}

\caption{Performance evaluation of \gls{vspyct} and \gls{spyct} models measured through the Brier score on the R survival datasets}
\label{fig:brier_scores}
\end{figure}

\subsection{Interpretability}

%slika od drvoto so oboeni edges

Figure~\ref{fig:inter_tree} presents the visual representation of the \gls{vspyct} applied to the Unemployment dataset.
Based on that, one can distill a concise interpretation enriched with practical insights.

\begin{figure}[h!]
    \centering
    \includegraphics{hecat_tree_old_temp.pdf}
    \caption{The \gls{vspyct} model applied to the Unemployment dataset. The colouring distinguishes the different feature groups of the data set}
    \label{fig:inter_tree}
\end{figure}

At the heart of this \gls{vspyct} model lies the ability to visually demarcate the influence of various feature categories on unemployment duration predictions.
The tree's root node is predominantly influenced by features categorised under ``other" (represented in green), encompassing age, months of work experience, and various demographic characteristics.
This indicates that such features play a pivotal role in initial split decisions, underscoring the broad impact of demographic and experiential factors on unemployment trends.

As we navigate through the tree, a dichotomy emerges based on the expected survival times—essentially, the projected duration an individual is likely to remain unemployed.
On branches leading to higher expected survival times, features categorised under ``p16 level of education" (coloured in blue) become increasingly dominant.
This suggests that higher levels of education are associated with longer periods of unemployment, potentially reflecting challenges in finding employment that matches an individual's qualifications or aspirations.

Conversely, on branches associated with lower expected survival times, the dominance shifts towards features represented by the ``ISCO classification of professional occupation" (coloured in red).
This dominance implies that certain professional occupations, as defined by the ISCO classification, may expedite reemployment, possibly due to higher demand or greater flexibility in job roles within these sectors.

This visual breakdown not only highlights the model's interpretability but also offers tangible insights into how different factors—ranging from personal demographics and work experience to education and professional classifications—impact unemployment durations.
Such insights are invaluable for policymakers and stakeholders aiming to devise targeted interventions that address specific barriers to employment faced by diverse demographic groups.

In essence, the \gls{vspyct} model's tree structure, with its intuitive colour-coded feature categories and quantified predictions at the leaf nodes, offers a powerful tool for understanding the multifaceted nature of unemployment.
It bridges the gap between complex machine learning models and practical socio-economic applications, providing a clear, actionable framework for addressing unemployment through informed, data-driven strategies.

\section{Conclusion}

We introduced the \gls{vspyct} model, an approach that strategically combines the benefits of oblique splits, variational Bayes, and Bayesian inference within a single, interpretable framework.
The \gls{vspyct} model addresses the trade-off between the performance of ensemble models and the interpretability of single decision trees, and the absence of inherent uncertainty quantification in predictions.
Our results demonstrate that the \gls{vspyct} model is competitive to, and on some data sets exceeds the performance of ensemble of \glspl{spyct}, without sacrificing the model's interpretability.
By employing variational Bayes method, we introduce a mechanism for uncertainty quantification directly into the decision-making process, enhancing the reliability and applicability of the model across various domains where decision certainty is paramount.
%The \gls{vspyct} model marks a significant step forward in making complex decision trees more accessible, interpretable, and effective.
The model opens new possibilities for research in semi-supervised learning and Bayesian inference, promising to bridge the gap between theoretical advancements and practical applications.
Future work will focus on further refining the model's performance, and extending its applications in diverse \gls{sop} settings.

\section*{Acknowledgement}
The authors acknowledge the research core funding No.\ P2-0001,  P2-0103 and P2-0383 financially supported by the Slovenian Research Agency.
The authors also acknowledge the funding received from the European Union's Horizon 2020 research and innovation programme project HECAT under grant agreement No.\ 870702 .

%\printbibliography

\bibliographystyle{elsarticle-harv}
\bibliography{bibliography}

\end{document}

### 2024-07-04 src/data/dataset.py

@@ -8,6 +8,7 @@ from sklearn.preprocessing import MinMaxScaler, OneHotEncoder
 from sklearn.model_selection import train_test_split
 import pickle
 from torch.utils.data import Dataset, DataLoader
+from tqdm import tqdm
 
 DIR_PATH = os.path.abspath(os.path.dirname(__file__))
 
@@ -41,30 +42,30 @@ class SurvivalDataset:
     self.preprocessed = None
     self.random_state = random_state
 
-  def preprocess(self):
+  def preprocess(self, time = 'time', status = 'status'):
     if self.preprocessed is not None: return self.preprocessed
     df = self.dataset.copy()
     df = df.drop(columns=['id']) if 'id' in df.columns else df
     object_cols = df.select_dtypes(include=['object', 'category']).columns
     df = pd.get_dummies(df, columns=object_cols)
     df = df.fillna(df.mean())
-    numeric_cols = df.select_dtypes(include=['number']).drop(columns=['time', 'status']).columns
+    numeric_cols = df.select_dtypes(include=['number']).drop(columns=[time, status]).columns
     from sklearn.preprocessing import StandardScaler
     scaler = StandardScaler()
     df[numeric_cols] = scaler.fit_transform(df[numeric_cols])
-    df['status'] = df['status'].map({1: 0, 2:1, 0:0})
+    if 2 in df[status].tolist(): df[status] = df[status].map({1: 0, 2:1, 0:0})
     df[df.select_dtypes(include=['bool']).columns] = df.select_dtypes(include=['bool']).astype('int')
     self.preprocessed = df
     return df
   
-  def get_survival_vector(self):
-    if self.preprocessed is None: self.preprocess()
-    survival_time = self.preprocessed.time.values
-    survival_status = self.preprocessed.status.values
+  def get_survival_vector(self, time = 'time', status = 'status'):
+    if self.preprocessed is None: self.preprocess(time, status)
+    survival_time = self.preprocessed[time].values
+    survival_status = self.preprocessed[status].values
     time_upper_limit = int(survival_time.max())
     time_tensor = torch.empty((survival_time.shape[0], time_upper_limit))
     print(time_tensor.shape[1])
-    for i,el in enumerate(survival_time):
+    for i,el in tqdm(enumerate(survival_time)):
       for j in range(time_tensor.shape[1]):
         if (survival_status[i] == 1) and (j>=el): time_tensor[i][j] = np.nan
[…]

### 2024-07-04 src/eval.py

@@ -8,18 +8,40 @@ import torch
 class IPCWBrier:
   def __init__(self, times, event_observed):
     from lifelines import KaplanMeierFitter
+    from scipy.interpolate import interp1d
     self.times = times
     self.event_observed = event_observed
     self.num_events = sum(event_observed)
     self.kmf = KaplanMeierFitter()
     self.kmf.fit(self.times, self.event_observed)
-    cumulative_censoring_probs = 1 - self.kmf.survival_function_.values
-    self.ipcw_coeffs = (1 / cumulative_censoring_probs)[1:]
+    
+    # Calculate cumulative censoring probabilities
+    cumulative_censoring_probs = 1 - self.kmf.survival_function_.values.flatten()
+    
+    # Get the time points corresponding to the survival function estimates
+    kmf_times = self.kmf.survival_function_.index.values
+    
+    # Interpolate to match the original number of time points
+    # Create an interpolation function based on the kmf times and the cumulative censoring probabilities
+    interp_func = interp1d(kmf_times, cumulative_censoring_probs, fill_value="extrapolate")
+    
+    # Use the interpolation function to estimate ipcw_coeffs for each original time point
+    self.ipcw_coeffs = interp_func(self.times)
+    
+    # Add a small epsilon value to avoid division by zero
+    epsilon = 1e-10
+    self.ipcw_coeffs = np.clip(self.ipcw_coeffs, epsilon, 1)
+    
+    # Inverse to get IPCW coefficients
+    self.ipcw_coeffs = 1 / self.ipcw_coeffs
 
   def evaluate(self, times_arr, predicted_probs):
     if isinstance(times_arr, torch.Tensor): times_arr = times_arr.numpy()
     if isinstance(predicted_probs, torch.Tensor): predicted_probs = predicted_probs.numpy()
-    brier_score = torch.nanmean(torch.tensor(self.ipcw_coeffs) * torch.tensor(np.power(times_arr - predicted_probs, 2)), axis=0) / self.num_events
+    diff = torch.tensor(np.power(times_arr - predicted_probs, 2))
+    ipcw_coeffs = torch.tensor(self.ipcw_coeffs, dtype=torch.float32)  # Ensure it's a PyTorch tensor
+    ipcw_coeffs = ipcw_coeffs.unsqueeze(-1)
+    brier_score = torch.nanmean(ipcw_coeffs * diff, axis=0) / self.num_events
     return brier_score
 
 

### 2024-07-04 src/models/node.py

@@ -68,7 +68,7 @@ class VNode:
                 if splits <= 0.5: return self.left.predict(x)
                 else: return self.right.predict(x)
 
-    def mc_predict(self, x, num_samples=50):
+    def mc_predict(self, x, num_samples=10):
         pred_lst = []
         for i in range(num_samples):
             pred_lst.append(self.predict(x))

### 2024-07-04 src/models/split.py

@@ -183,6 +183,6 @@ def learn_split_vb(rows, descriptive_data, clustering_data, device, epochs, bs, 
             print(f"Early stopping at epoch {epoch}.")
             break
         losses.append(train_loss)
-        print("[iteration %04d] loss: %.4f" % (epoch+1, train_loss))
+        # print("[iteration %04d] loss: %.4f" % (epoch+1, train_loss))
     print(model)
     return (model, guide, param_store)

### 2024-07-08 reports/paper/vspyct.tex
My comments

@@
 This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterized by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
-Typically, the mean-field variational family is utilized, assuming independence among the parameters $\omega$.
-To find the optimal $\omega^*$, one minimises the Kullback–Leibler (KL) divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:
+Typically, the mean-field variational family is utilized, assuming independence among the parameters $\omega$\pbnote{Add citation}. 
+To find the optimal $\omega^*$, one minimises the Kullback–Leibler (KL)\pbnote{Use glossaries} divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:
 
@@
 
-Given the unknown nature of the true posterior, a rearrangement is needed to compute the KL divergence, simplifying to:
+Given the unknown nature of the true posterior, a rearrangement is needed to compute the KL divergence, simplifying to \pbnote{mislam deka ne ti treba celiot derivation.
+Može da napraviš referenca do PML2 od Murphy in da go dadeš samo poslediont red}:
 
@@
 
-Maximising the ELBO inversely minimises the KL divergence, with $\log p(x)$ remaining constant and excluded from the optimisation.
-Typically, the ELBO maximisation criterion is non-convex, with no assurance of global extremum convergence by optimisation algorithms.
-Various algorithms have been employed for this challenge, including coordinate ascent variational inference and stochastic variational inference~\cite{hoffman2013stochastic}, with the latter's gradient calculated as per the black-box variational inference~\cite{sashank2018convergence}.
-The adoption of an approximate variational distribution introduces bias contingent on the chosen variational family $\mathcal{Q}$, necessitating a decision grounded in empirical evidence or expert knowledge.
+Maximising the ELBO\pbnote{Use package glossaries for abbreviations} inversely minimises the KL divergence, with $\log p(x)$ remaining constant and excluded from the optimisation \pbnote{why is it excluded, explain}.
+Typically, the ELBO maximisation criterion is non-convex, with no assurance of global extremum convergence by optimisation algorithms \pbnote{True, add citation}.
+Various algorithms have been employed for this challenge, including coordinate ascent variational inference and stochastic variational inference~\citep{hoffman2013stochastic}, with the latter's gradient calculated as per the black-box variational inference~\citep{sashank2018convergence}.
+\pbnote{Ne se samo tie. Vidi tuka \url{https://probml.github.io/pml-book/book2.html} chapter 10}
+The adoption of an approximate variational distribution introduces bias \pbnote{True, but explain where the bias is} contingent on the chosen variational family $\mathcal{Q}$, necessitating a decision grounded in empirical evidence or expert knowledge.
 Despite this bias, VB's computational efficiency significantly surpasses traditional methods like Monte Carlo integration.
@@
 \subsection{Semi-supervised Multi-target Regression in Survival Analysis}
-
+\pbnote{Do we need this subsection here?
+Maybe put the name to methodology, 3.1.
+is VB and then continue directly to 4.
+This subsection can go into results or application and you can actually reference it from ESWA paper}
 Survival analysis primarily focuses on time-to-event data, where not all events are fully observed due to some form of censoring that halts observation before the event transpires.
@@
 This allows for more complex decision boundaries, capturing intricate patterns in the data with fewer nodes compared to axis-aligned trees.
-However, unlike traditional oblique trees that fix the parameters of these linear combinations, \gls{vspyct} introduces a probabilistic twist: the parameters of the splits are not single fixed values but are instead modelled as random variables with Gaussian distributions.
-This probabilistic representation captures the uncertainty in the model parameters, a feature that is particularly useful in scenarios with noisy data.
-The construction of the \gls{vspyct} begins with a root node and iteratively expands by adding internal nodes or leaves through a learning process based on the variational Bayes method.
+However, unlike traditional oblique trees that fix the parameters of these linear combinations, \gls{vspyct} introduces a probabilistic twist: the parameters of the splits are \mpop{not single fixed values but are instead}{} modelled as random variables \pbnote*{It does not need to be Gaussian}{with Gaussian distributions}.
+This probabilistic representation captures the uncertainty in the model parameters \pbnote*{Why noisy data, what if we have perfect data but wrong model}{, a feature that is particularly useful in scenarios with noisy data}.
+The construction of the \gls{vspyct} \pbnote{here you can say that the consutrction follows SPYCT, because the process is the same.
+Only the optimisation step is different.
+Explain this better here}begins with a root node and iteratively expands by adding internal nodes or leaves through a learning process based on the variational Bayes method.
 Each node represents a binary decision point where instances are split into two child nodes.
 At each internal node, a variational Bayes approach is used to learn the parameters of an oblique split.
-This involves optimising the parameters of a linear model that best divides the data to minimise the ELBO.
+\pbnote*{Keep ELBO until later.
+Here explain the basic SPYCT, there are no stochastic parts here yet} {This involves optimising the parameters of a linear model that best divides the data to minimise the ELBO.}
 The linear model is defined as:
@@
 where $\mathbf{x} \in \mathbb{R}^N$ is the feature vector, $\mathbf{w} \in \mathbb{R}^N$ represents the weights, $b \in \mathbb{R}$ is the bias, and $\sigma(\cdot)$ denotes the sigmoid function ensuring the output lies in $(0, 1)$, indicating the probability of branching to the right child node.
+
+\pbnote{The part below should be separate.
+Above explain how the mechanism of SPYCT works.
+Below you can point out the main difference.
+May be merge this with the next subsection, but it shoud le rewritten carefully.
+Here you are missing your key idea. How to minimise the impurity function. You got an idea that you should put the target value at 0 and run ELBO on it. This is missing here.}
 The learning process involves defining a prior distribution over the weights and biases, typically Gaussian, and updating this to a posterior distribution given the observed data through variational inference.
@@
 
+\pbnote{Algorithm Impurity is not referenced anywhere in the text. Do you need it at all?}
 \begin{algorithm}[h!]
@@
 The procedure for making a prediction for a new instance using the \gls{vspyct} model is given in Algorithm~\ref{alg:make_pred}.
+\pbnote{How? What do you get from predictive? Why do you need predictive, why not just using the estimated model parameters directly? This subsection is not written OK!} 
+\pbnote{What is prototype in Algorithm~\ref{alg:make_pred}?} 
+\pbnote{Can you write this in a mathematical form}
 It involves traversing the tree from the root to a leaf, at each internal node using the variational posterior to sample parameters and compute the probability of branching right or left.
@@
 As a result, $K$ equals $N + T$, with $N$ representing the feature count and $T$ denoting the target count.
-The time complexity for learning in \glspl{spyct} can be expressed as $\mathcal{O}(NI_o(D+K))$, wherein $I_o$ stands for the iteration count needed for optimisation.
+The time complexity for learning in \glspl{spyct} can be expressed as $\mathcal{O}(NI_o(D+K))$, wherein $I_o$ stands for the iteration count needed for optimisation\pbnote{Citation}.
 Notably, the computational demand for the \glspl{spyct}' gradient-based version increases linearly with the attribute count.
@@
 This iterative process involves sampling from the posterior distributions of split parameters, evaluating the ELBO, and updating the parameters accordingly.
-Consequently, the time complexity for learning a split in \gls{vspyct} can be described as $\mathcal{O}(MI_oNI_{vb})$, where $M$ is the number of Monte Carlo samples used to approximate the posterior distributions, $I_{vb}$ represents the number of iterations required for variational inference to converge, and the other variables retain their meanings from the \gls{spyct} context.
+Consequently, the time complexity for learning a split in \gls{vspyct} can be described as $\mathcal{O}(MI_oNI_{vb})$, where $M$ is the number of Monte Carlo samples used to approximate the posterior distributions, $I_{vb}$ represents the number of iterations required for variational inference to converge, and the other variables retain their meanings from the \gls{spyct} context \pbnote{This is great but you have to explain how did you get to this complexity.
+Can you write the derivation}.
 
@@
 
+\pbnote{Maybe this paragraph can go to conclusion or discussion}
 The choice between using \glspl{vspyct} over \glspl{spyct} hinges on the specific requirements of the application scenario.
@@
 
-Feature importance is derived from the aggregated impact of features across all splits, quantified by examining the variational posterior distributions of the weights.
+\pbnote*{Complex sentence}{Feature importance is derived from the aggregated impact of features across all splits, quantified by examining the variational posterior distributions of the weights.}
 The importance of each feature is proportional to its contribution to reducing impurity across splits.
@@
 \end{equation}
-
-where $T$ represents the variational oblique predictive clustering tree, $s$ iterates over the set of split nodes within $T$, $s_n$ is the number of training examples that passed through node $s$, $N$ denotes the total number of training examples used in the tree, $\mathbf{w}_s$ symbolizes the weight vector defining the oblique split at node $s$, treated as a random variable with a Gaussian distribution, $\mathbb{E}[\mathbf{w}_s]$ signifies the expected value (mean) of the weight vector $\mathbf{w}_s$, calculated from its posterior distribution as inferred by the variational Bayesian method.
-The normalization factor $|\mathbb{E}[\mathbf{w}_s]|_1$ ensures the comparability of feature importances by scaling them relative to the aggregate magnitude of expected weights.
+where $s$ is a split in the variational oblique predictive clustering tree $T$, $s_n$ is the number of training examples that passed through node $s$, \pbnote*{Shoudn't this be $N_s$}{$N$} denotes the total number of training examples used in the tree, $\mathbf{w}_s$ symbolizes the weight vector defining the oblique split at node $s$\mpop{, treated as a random variable with a Gaussian distribution}{}, $\mathbb{E}[\mathbf{w}_s]$ signifies the expected value \mpop{(mean)}{ }of the weight vector $\mathbf{w}_s$, calculated from its \pbnote*{Careful, posterior or vairational or predictive}{posterior} distribution as inferred by the variational Bayesian method.
+The normalization factor \pbnote*{What does $_1$ mean?
+Can you give a complete definition}{$|\mathbb{E}[\mathbf{w}_s]|_1$} ensures the comparability of feature importances by scaling them relative to the aggregate magnitude of expected weights.
 This formulation accounts for the stochastic nature of each parameter, reflecting the central tendency of the distributions defining the split hyperplanes.
 
-For practical computation, the expected weights $\mathbb{E}[\mathbf{w}_s]$ are obtained by sampling from the posterior distributions inferred during the learning process.
-Each weight vector $\mathbf{w}_s$ is drawn such that $\mathbf{w}_s \sim q(\mathbf{w}_s|\mathcal{D})$, where $q(\mathbf{w}_s|\mathcal{D})$ represents the variational approximation to the posterior distribution of weights given the data $\mathcal{D}$.
+For practical computation, the expected weights $\mathbb{E}[\mathbf{w}_s]$ are obtained by sampling from the \pbnote*{Again which?
+Where in the code is this?}{posterior} distributions inferred during the learning process.
+Each weight vector $\mathbf{w}_s$ is drawn such that $\mathbf{w}_s \sim q(\mathbf{w}_s|\mathcal{D})$, where $q(\mathbf{w}_s|\mathcal{D})$ represents the \pbnote*{so, variational approximation not the posterior}{variational approximation} to the posterior distribution of weights given the data $\mathcal{D}$.
 The importance of each feature is then a function of these expected weights, modulated by the frequency of training instances influencing each split, thereby emphasising the splits that impact a larger portion of the data.
@@
 
+\pbnote{This is rather complex subsection it needs more bachground. How did you come up with the idea of this relation, provide simple toy example with 2 or 3 weights to explain how this works}
+
 \section{Data and evaluation metric}
+\pbnote{Now I understand.
+Why did you focus only on survival data.
+If I remember correctly you calculated the performance on different data sets not just survival.
+If we leave it only on this it will not go OK with the review.}
 

### 2024-12-13 reports/paper/vspyct.tex
Extra comment
new version

--- 2024-07-04 reports/paper/vspyct.tex
@@
 \usepackage{pythontex}
+\usepackage{graphicx}
 
@@
 \newacronym{vspyct}{VSPYCT}{variational oblique predictive clustering tree}
+\newacronym{opct}{OPCT}{option predictive clustering tree}
 \newacronym{sslmtr}{SSL-MTR}{semi-supervised multi-target regression}
@@
 \newacronym{isce}{ISCED}{International Standard Classification of Education}
+\newacronym{kl}{KL}{Kullback–Leibler}
+\newacronym{elbo}{ELBO}{Evidence Lower Bound}
 
@@
 \begin{frontmatter}
-	\title{Variational oblique predictive clustering trees for multi-target regression}
+	\title{Variational oblique predictive clustering trees}
 	\author[1,2]{Viktor Andonovikj\corref{cor1}}
@@
 We propose variational oblique predictive clustering trees, which address this challenge.
-The parameters of each split node are treated as distributions and they are learned through the Variational Bayes method. We evaluate the model on several benchmark datasets of different sizes.
-The experimental analyses show that a single \gls{vspyct} achieves state-of-the-art or on-par performance with ensemble of standard \glspl{spyct}.
+The parameters of each split node are treated as random variables, described with a probability distribution, and they are learned through the Variational Bayes method.
+We evaluate the model on several benchmark datasets of different sizes.
+The experimental analyses show that a single \gls{vspyct} achieves competitive, and sometimes better predictive performance than the ensemble of standard \glspl{spyct}.
 We also present a method for extracting feature importance scores from the model.
@@
 This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.
-The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, modelling these parameters as random variables with Gaussian distributions.
+The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, modelling these parameters as random variables with probability distributions.
 This represents a paradigm shift toward a probabilistic understanding of decision-making processes within tree-based models, significantly enhancing interpretability, especially in uncertainty estimation domains.
-The whole framework is given in Figure~\ref{fig:flowchart}
-
-\begin{figure}[h!]
-    \centering
-    \includegraphics[width=1.0\textwidth]{methodology_flowchart.pdf}
-    \caption{Flowchart of the methodology}
-    \label{fig:flowchart}
-\end{figure}
-
 It sets a new benchmark for interpretable, efficient, and reliable machine learning models.
@@
 
-Our proposed \glspl{vspyct} model addresses these critical challenges by integrating Bayesian principles directly into the decision tree framework.
+Our proposed \gls{vspyct} model addresses these critical challenges by integrating Bayesian principles directly into the predictive clustering tree framework.
 %Unlike traditional ensemble strategies, \glspl{vspyct} employs a probabilistic treatment of model parameters through the variational Bayes method, which enhances interpretability and provides inherent uncertainty quantification.
 This approach maintains the clarity and simplicity of a single tree model while achieving state-of-the-art performance, thus offering a significant improvement over existing methods.
-By introducing uncertainty quantification directly into the decision-making process, \glspl{vspyct} enhances the reliability and applicability of the model across various domains where decision certainty is crucial.
+By introducing uncertainty quantification directly into the decision-making process, \gls{vspyct} enhances the reliability and applicability of the model across various domains where decision certainty is crucial.
 %This advancement bridges the gap between theoretical developments and practical applications, marking a significant step forward in making complex decision trees more accessible, interpretable, and effective.
 
-\section{Preliminaries}
-
-\subsection{Variational Bayes method}
+\section{The variational Bayes method}
 
@@
 
-In conducting a stochastic analysis, like survival analysis, the process of inferring model parameters hinges on applying Bayes' theorem.
+In conducting a stochastic analysis, the process of inferring model parameters hinges on applying Bayes' theorem.
 For observations $x$ produced by a system defined by parameters $\theta$, Bayes' theorem is given as:
@@
 This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterized by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
-Typically, the mean-field variational family is utilized, assuming independence among the parameters $\omega$.
-To find the optimal $\omega^*$, one minimises the Kullback–Leibler (KL) divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:
+Typically, the mean-field variational family is utilised, assuming independence among the parameters $\omega$~\citep{Blei_2017}. 
+To find the optimal $\omega^*$, one minimises the \gls{kl} divergence $KL(q_\omega(\theta) \parallel p(\theta|x))$:
 
@@
 
-Given the unknown nature of the true posterior, a rearrangement is needed to compute the KL divergence, simplifying to:
+Given the unknown nature of the true posterior, a rearrangement is needed to compute the \gls{kl} divergence, simplifying to~\citep[Chapter~10]{murphy2023probabilistic}:
 
 \begin{align*}
-KL(q_\omega(\theta) \parallel p(\theta|x)) &= \mathbb{E}_q\left[\log \frac{q_\omega (\theta)}{p(\theta|x)}\right] \\
-&=\mathbb{E}_q\left[\log q_\omega(\theta)\right]-\mathbb{E}_q\left[\log p(\theta|x)\right]\\
-&=\mathbb{E}_q\left[\log q_\omega(\theta)\right]-\mathbb{E}_q\left[\log p(x,\theta)\right]\\
-&=\mathbb{E}_q[\log q_\omega(\theta) - \log p(x,\theta)+\log p(x)]\\
-&=-\mathbb{E}_q\left[\log q_\omega(\theta) - \log q_\omega(\theta)\right] + \log p(x)\\
-&\quad \quad \underbrace{\phantom{\mathbb{E}_q\left[\log q_\omega(\theta) - \log q_\omega(\theta)\right]}}_{\text{ELBO}}
+KL(q_\omega(\theta) \parallel p(\theta|x)) = - & \mathbb{E}_q\left[\log p(x,\theta) - \log q_\omega(\theta)\right] + \log p(x)\\
+& \underbrace{\phantom{\mathbb{E}_q\left[\log p(x,\theta) - \log q_\omega(\theta)\right]}}_{\text{ELBO}}
 \label{eq:kl_divergence}
@@
 
-Maximising the ELBO inversely minimises the KL divergence, with $\log p(x)$ remaining constant and excluded from the optimisation.
-Typically, the ELBO maximisation criterion is non-convex, with no assurance of global extremum convergence by optimisation algorithms.
-Various algorithms have been employed for this challenge, including coordinate ascent variational inference and stochastic variational inference~\cite{hoffman2013stochastic}, with the latter's gradient calculated as per the black-box variational inference~\cite{sashank2018convergence}.
+Maximising the \gls{elbo} serves to inversely minimise the \gls{kl} divergence between the approximate posterior \(q_\omega(\theta)\) and the true posterior \(p_\omega(\theta)\).
+In the decomposition of the KL divergence, the term \(\log p(x)\), known as the marginal likelihood or evidence, remains constant with respect to the variational parameters \(\theta\).
+Since this term does not involve \(\theta\), it does not influence the optimisation of the variational distribution and is thus excluded from the \gls{elbo} maximisation process.
+By maximising the \gls{elbo}, we aim to tighten the bound provided by the \gls{kl} divergence, effectively making the variational posterior a better approximation of the true posterior.
+Typically, the \gls{elbo} maximisation criterion is non-convex, with no assurance of global extremum convergence by optimisation algorithms~\citep[Chapter~10]{murphy2023probabilistic}.
+Various algorithms have been proposed for this challenge, including variational EM \citep{bernardo2003variational}, stochastic variational inference~\citep{hoffman2013stochastic, sashank2018convergence}, amortised variational inference~\citep{gershman2014amortized, le2017inference}, and semi-amortised inference~\citep{kim2018semi}.
+
 The adoption of an approximate variational distribution introduces bias contingent on the chosen variational family $\mathcal{Q}$, necessitating a decision grounded in empirical evidence or expert knowledge.
-Despite this bias, VB's computational efficiency significantly surpasses traditional methods like Monte Carlo integration.
-In our work, the ADAM optimiser~\citep{kingma2014adam} is used to facilitate the ELBO optimisation.
-
-\subsection{Semi-supervised Multi-target Regression in Survival Analysis}
-
-Survival analysis primarily focuses on time-to-event data, where not all events are fully observed due to some form of censoring that halts observation before the event transpires.
-The most prevalent form of censoring in this context is right-censoring, which occurs when there is incomplete knowledge about the survival time at the end of the follow-up period.
-As a result, each survival record is characterised by three elements: the actual survival time $T$, its observed component $t$, and a binary indicator $\delta$, which is defined as:
-\begin{equation}
-\delta = \begin{cases}
-0, & \text{if } T>t \text{ indicating a right-censored record},\\
-1, & \text{if } T=t \text{, denoting that the event has been observed and the record is not censored}.
-\end{cases}
-\label{eq:delta}
-\end{equation}
-
-Moreover, assume that each observation is described by a set of covariates or features $\mathbf{X}_i=\{x_{i_1},\ldots,x_{i_N}\}$.
-Thus, each record $i$ comprises the variables $\mathbf{X}_i$, $t_i$, and $\delta_i$ as specified in~\eqref{eq:delta}.
-If $\delta_k=1$, the event is confirmed to have occurred at time $t_k$ and $y_{k_j} = 0$ for all $j \geq{k}$.
-Conversely, if $\delta_k=0$, it implies that the fate of the instance beyond time $t_k$ is unknown, making $y_{k_j}$ missing for $j \geq{t_k}$.
-This scenario of censored examples introduces missing data, which makes such datasets apt for semi-supervised learning methodologies."
+This bias arises because the family \(\mathcal{Q}\) may not be capable of capturing the true complexity of the posterior distribution \(p_\omega(\theta)\).
+As a result, the accuracy of the approximation depends significantly on how well \(\mathcal{Q}\) aligns with the true underlying distribution.
+The choice of \(\mathcal{Q}\) thus not only affects the efficiency of the inference but also the quality and reliability of the model.
+Despite this bias, VB's computational efficiency significantly surpasses traditional methods like Markov Chain Monte Carlo (MCMC).
+In our case, the ADAM optimiser~\citep{kingma2014adam} is used to facilitate the \gls{elbo} optimisation.
 
@@
 
-As a main part of the \gls{vspyct} model lies the oblique tree structure, which is a form of a \gls{pct} where splits are made using linear combinations of the input features rather than single features.
-This allows for more complex decision boundaries, capturing intricate patterns in the data with fewer nodes compared to axis-aligned trees.
-However, unlike traditional oblique trees that fix the parameters of these linear combinations, \gls{vspyct} introduces a probabilistic twist: the parameters of the splits are not single fixed values but are instead modelled as random variables with Gaussian distributions.
-This probabilistic representation captures the uncertainty in the model parameters, a feature that is particularly useful in scenarios with noisy data.
-The construction of the \gls{vspyct} begins with a root node and iteratively expands by adding internal nodes or leaves through a learning process based on the variational Bayes method.
-Each node represents a binary decision point where instances are split into two child nodes.
-At each internal node, a variational Bayes approach is used to learn the parameters of an oblique split.
-This involves optimising the parameters of a linear model that best divides the data to minimise the ELBO.
-The linear model is defined as:
+As the foundation of the \gls{vspyct} model, the underlying architecture closely follows that of the \gls{spyct} model, particularly in its tree construction process.
+Similar to \gls{spyct}, \gls{vspyct} constructs a decision tree with an oblique structure, where each split is defined by a linear combination of input features, rather than relying on single-feature thresholds as in axis-aligned trees.
+
+The key distinction between \gls{vspyct} and \gls{spyct} lies in the optimisation step used to determine the parameters of these oblique splits.
+While \gls{spyct} utilises a deterministic approach to optimise split parameters, \gls{vspyct} introduces a probabilistic framework through variational Bayes (VB) optimisation.
+In \gls{vspyct}, the parameters of the linear combination at each split—specifically, the weight vector \(\mathbf{w}\) and the bias term \(b\)—are modelled as random variables with Gaussian distributions.
+This probabilistic representation of split parameters not only enhances the model's ability to handle uncertainty and noise in the data but also provides a measure of predictive uncertainty.
+
+The construction of the \gls{vspyct} begins with a root node and proceeds iteratively, expanding the tree by adding internal nodes or leaf nodes in a manner identical to \gls{spyct}.
+Each node in the tree represents a binary decision point, where instances are split based on the oblique hyperplane defined by the linear model:
 
 \begin{equation}
-f(\mathbf{x})=\sigma \left(\mathbf{w}^T \mathbf{x} + b \right)
+f(\mathbf{x}) = \sigma \left(\mathbf{w}^\top \mathbf{x} + b \right)
 \label{eq:linear_model}
 \end{equation}
+where \(\mathbf{x} \in \mathbb{R}^N\) is the feature vector, \(\mathbf{w} \in \mathbb{R}^N\) represents the weight vector, \(b \in \mathbb{R}\) is the bias term, and \(\sigma(\cdot)\) denotes the sigmoid function, which ensures the output is a probability, determining whether an instance is sent to the right or left child node.
 
-where $\mathbf{x} \in \mathbb{R}^N$ is the feature vector, $\mathbf{w} \in \mathbb{R}^N$ represents the weights, $b \in \mathbb{R}$ is the bias, and $\sigma(\cdot)$ denotes the sigmoid function ensuring the output lies in $(0, 1)$, indicating the probability of branching to the right child node.
-The learning process involves defining a prior distribution over the weights and biases, typically Gaussian, and updating this to a posterior distribution given the observed data through variational inference.
-The ELBO is optimised using stochastic gradient descent, with the learning rate $\alpha$ and potentially variable batch sizes $B$.
-The split at each node is determined by evaluating the learned linear model on the descriptive data.
-Instances are split based on the probability threshold of $0.5$, with those having a probability greater than $0.5$ moving to the right child node and the rest to the left.
-For leaf nodes, a prototype is computed as the mean of the target variables for the instances assigned to the leaf.
-This prototype represents the prediction for instances that fall into the leaf during the prediction phase.
+While the tree structure and the decision-making process follow the same principles as in \gls{spyct}, the optimisation of \(\mathbf{w}\) and \(b\) in \gls{vspyct} differs fundamentally.
+In \gls{vspyct}, these parameters are not fixed but are instead inferred through variational Bayes optimisation.
+The introduction of this probabilistic element distinguishes \gls{vspyct} from \gls{spyct}, enabling it to maintain the interpretability of a single decision tree while incorporating the robustness typically associated with ensemble methods.
+
+\begin{comment}
+\pbnote{Why noisy data, what if we have perfect data but wrong model}
+\vanote{If we have perfect data - then even a deterministic model will do well for the prediction.
+We choose the type of model before we do the optimisation.
+If we choose a wrong model - e.g.
+a decision tree where it is simply a mismatch for the type of problem at hand - we will have bad predictive performance, regardless of the modelling framework (deterministic or probabilistic).
+But if we choose the right type of model, and face noisy data at test time - then a probabilistic model would capture the predictive uncertainty in its parameter distributions.}
+\end{comment}
 
@@
 
-The learning of these Gaussian-distributed parameters is achieved through a Variational Bayes (VB) method.
-In the context of \gls{vspyct}, the VB method is employed to learn the Gaussian distributions representing the parameters of the oblique splits.
-The process involves defining a prior distribution over the parameters and then updating this prior to a posterior distribution given the observed data.
-This is done by optimising the ELBO, which balances the fit of the model to the data with the complexity of the model.
-The procedure for learning a split using the VB method is given in Algorithm~\ref{alg:learn_split}.
+The goal is to update these prior distributions to approximate the posterior distributions given the observed data, thereby capturing the uncertainty in the model parameters.
+The weights \(\mathbf{w}\) and bias \(b\) in the linear model, as in~\eqref{eq:linear_model}, are initially modelled with Gaussian prior distributions:
+
+\begin{equation}
+\mathbf{w} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}), \quad b \sim \mathcal{N}(0, 1)
+\end{equation}
+
+The VB method aims to approximate the true posterior distributions of these parameters with a variational distribution \(q(\mathbf{w}, b|\mathcal{D})\).
+We assume a simpler distributional form - a diagonal Gaussian:
+
+\begin{equation}
+q(\mathbf{w}, b|\mathcal{D}) = \mathcal{N}(\mathbf{w}|\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w) \mathcal{N}(b|\mu_b, \sigma_b^2)
+\end{equation}
+
+where \(\boldsymbol{\mu}_w\), \(\boldsymbol{\Sigma}_w\), \(\mu_b\), and \(\sigma_b^2\) are the parameters of the variational distribution, learned through the optimisation process.
+
+The optimisation of the variational parameters is performed by maximising the \gls{elbo}, which serves as a lower bound to the log marginal likelihood \(\log p(\mathcal{D})\).
+The \gls{elbo} is given by:
+
+\begin{equation}
+\mathcal{L}(\boldsymbol{\mu}, \boldsymbol{\Sigma}) = \mathbb{E}_{q(\mathbf{w}, b|\mathcal{D})}\left[\log p(\mathcal{D}|\mathbf{w}, b)\right] - KL\left(q(\mathbf{w}, b|\mathcal{D}) \parallel p(\mathbf{w}, b)\right)
+\label{eq:elbo}
+\end{equation}
+
+The first term represents the expected log-likelihood of the data under the variational distribution, while the second term is the Kullback-Leibler (KL) divergence between the variational distribution and the prior.
+The \gls{elbo} is maximised using the ADAM optimiser, with the goal of refining the variational parameters \(\boldsymbol{\mu}\) and \(\boldsymbol{\Sigma}\) to closely approximate the true posterior.
+
+The quality of a split is assessed using an impurity function, which is defined as:
+
+\begin{equation}
+\text{Impurity} = \sum_{i=1}^{N} \left(\rho_i \cdot \text{Var}_{\rho}(\mathbf{Y}) + \lambda_i \cdot \text{Var}_{\lambda}(\mathbf{Y})\right)
+\end{equation}
+
+where \(\rho_i = \sigma(\mathbf{w}^\top \mathbf{x}_i + b)\) is the probability of an instance being routed to the right child node, and \(\lambda_i = 1 - \rho_i\) is the probability of routing to the left child node.
+The variances \(\text{Var}_{\rho}(\mathbf{Y})\) and \(\text{Var}_{\lambda}(\mathbf{Y})\) represent the variance of the target variables in the right and left child nodes, respectively, weighted by their selection probabilities.
+
+The objective is to minimise the impurity by optimising the parameters \(\mathbf{w}\) and \(b\) using the VB method, thereby ensuring that the split at each node effectively separates the data into homogeneous subsets.
+
+The overall procedure for learning a split in \gls{vspyct} is detailed in Algorithm~\ref{alg:learn_split_vb}.
+The algorithm involves initialising the variational parameters, performing Monte Carlo sampling to estimate the \gls{elbo}, and iteratively updating the parameters until convergence.
 
 \begin{algorithm}[h!]
-\caption{Variational Learning of Split Parameters in \gls{vspyct}}
-\label{alg:learn_split}
+\caption{Variational Learning of Split Parameters in \gls{vspyct} with Impurity Minimization}
+\label{alg:learn_split_vb}
 \begin{algorithmic}[1]
+    \State \textbf{Input:} $\mathcal{D} = \{\mathbf{X}, \mathbf{Y}\}$ (dataset), $\theta = \{\mathbf{w}_0, b_0\}$ (initial parameters), $E$ (epochs), $\lambda$ (learning rate), $\beta$ (batch size), $\sigma$ (selection probability)
+    \State \textbf{Output:} $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ (variational parameters)
 
-\State \textbf{Input:} Dataset $\mathcal{D}$, features $\mathbf{X}$, targets $\mathbf{Y}$, parameters $\theta$, epochs $E$, batch size $\beta$, learning rate $\lambda$, selection probability $\sigma$
-\State \textbf{Output:} Parameters $\Theta$, guide $\mathcal{G}$
-
-\Procedure{LearnSplit}{$\mathcal{D}$, $\mathbf{X}$, $\mathbf{Y}$, $\theta$, $E$, $\lambda$, $\beta$, $\sigma$}
-    \State $\Phi \sim \text{Bernoulli}(\sigma)$ \Comment{Select attributes based on probability $\sigma$}
-    \State $\mathbf{X}_\Phi \leftarrow$ subset of $\mathbf{X}$ using $\Phi$
-    \State Model $\mathcal{M} \leftarrow \text{Define Impurity}(\dim(\mathbf{X}_\Phi), 1, \theta)$
-    \State Guide $\mathcal{G} \leftarrow \text{Define AutoDiagonalNormal}(\mathcal{M})$
-    \State Define optimization parameters $\alpha \leftarrow \{\text{"lr"}: \lambda\}$
-    \State Initialise optimizer $\text{Optimizer} \leftarrow \text{Adam}(\alpha)$
-    \State Early stopping mechanism $\text{EarlyStopping} \leftarrow \text{Initialize}(\beta)$
-
-    \State Initialise loss list $L$
-    \State Calculate number of batches $B \leftarrow \lceil \frac{|\mathcal{D}|}{\beta} \rceil$
-
-    \For{$e \in \{1, \dots, E\}$}
-        \State Initialise training loss $\mathcal{L}_{\text{train}} \leftarrow 0$
-        \For{$b \in \{1, \dots, B\}$}
-            \State $(\mathbf{X}_b, \mathbf{Y}_b) \leftarrow \text{Get Batch}(\mathbf{X}_\Phi, \mathbf{Y}, b, \beta)$
-            \State Update training loss $\mathcal{L}_{\text{train}} \leftarrow \mathcal{L}_{\text{train}} + \text{SVI Step}(\mathcal{M}, \mathcal{G}, \mathbf{X}_b, \mathbf{Y}_b)$
+    \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
+        \State Initialize $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialize variational parameters}
+        \State Initialize optimizer: $\alpha \gets \lambda$
+        \For{$e \in \{1, \dots, E\}$} \Comment{Iterate over epochs}
+            \For{each mini-batch $\mathcal{B} \subseteq \mathcal{D}$ of size $\beta$}
+                \State Sample $\mathbf{w} \sim \mathcal{N}(\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w)$, $b \sim \mathcal{N}(\mu_b, \sigma_b^2)$
+                \State Compute impurity $\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)$ as:
+                \[
+                \Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b) = \sum_{\text{right}} \rho(\mathbf{w}^\top \mathbf{x} + b) \cdot \text{Var}_\rho(\mathbf{Y}) + \sum_{\text{left}} (1 - \rho(\mathbf{w}^\top \mathbf{x} + b)) \cdot \text{Var}_\lambda(\mathbf{Y})
+                \]
+                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$
+                \State Condition on the observed impurity using:
+                \[
+                \log p(\Omega_\text{obs}|\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b))
+                \]
+                \State Compute ELBO $\mathcal{L}(\mathcal{B}; \Theta)$ as per~\eqref{eq:elbo}
+                \State Compute gradient $\nabla_{\Theta} \text{ELBO}$
+                \State Update $\Theta \gets \Theta + \alpha \nabla_{\Theta} \text{ELBO}$ \Comment{Gradient descent update}
             \EndFor
-        \State Append $\mathcal{L}_{\text{train}}$ to $L$
-        \If{$\text{EarlyStopping}(\mathcal{L}_{\text{train}})$}
+            \If{early stopping criterion satisfied}
                 \State \textbf{break}
@@
         \EndFor
-    \State \Return $\mathcal{M}$, $\mathcal{G}$
+        \State \textbf{return} $\Theta$
     \EndProcedure
-
-\end{algorithmic}
-\end{algorithm}
-
-\begin{algorithm}[h!]
-\caption{Impurity}
-\begin{algorithmic}[1]
-
-\Procedure{Impurity}{$\xi$, $\zeta$, $\theta$}
-    \State $\mathbf{w}, b \sim \mathcal{N}(0, I)$ \Comment{Weight and bias priors}
-    \State $\rho(\mathbf{x}) \gets \sigma(\mathbf{w}^\top \mathbf{x} + b)$ \Comment{Right selection probability}
-    \State $\lambda(\mathbf{x}) \gets 1 - \rho(\mathbf{x})$ \Comment{Left selection probability}
-    \State Compute $\text{Var}_\rho(\mathbf{Y})$ and $\text{Var}_\lambda(\mathbf{Y})$ \Comment{Weighted variances}
-    \State $\Omega \gets \lambda \cdot \text{Var}_\lambda(\mathbf{Y}) + \rho \cdot \text{Var}_\rho(\mathbf{Y}) + \|\mathbf{w}\|_{0.5}$
-%    \State Observe $\Omega$ with likelihood $\mathcal{N}(0, 1)$
-    \State \textbf{return} $\Omega$
-\EndProcedure
-
 \end{algorithmic}
@@
 
-The procedure for making a prediction for a new instance using the \gls{vspyct} model is given in Algorithm~\ref{alg:make_pred}.
-It involves traversing the tree from the root to a leaf, at each internal node using the variational posterior to sample parameters and compute the probability of branching right or left.
-The prediction of the reached leaf's prototype is returned as the output.
-Operationally, when a new data point is to be classified or a prediction is to be made, it traverses the tree from the root to a leaf.
-At each node, the oblique split is determined by evaluating the linear combination of the input features, with parameters sampled from the learned Gaussian distributions.
-This probabilistic approach to decision-making within the tree allows for a more nuanced interpretation of the data, accounting for uncertainty in the model parameters.
+The procedure for making a prediction for a new instance using the \gls{vspyct} model is detailed in Algorithm~\ref{alg:make_pred}.
+The process involves traversing the tree from the root node to a leaf node, with decisions at each internal node being made probabilistically based on the parameters sampled from the learned variational posterior distributions.
+The final prediction is determined by the prototype value stored in the reached leaf node.
 
-\begin{algorithm}
-\caption{\gls{vspyct} Prediction Process}
+Mathematically, at each internal node, the oblique split is determined by evaluating the linear combination of input features, with the parameters \(\Theta = (\mathbf{w}, b)\) drawn from their respective variational posterior distributions.
+The decision to branch left or right is based on the probability computed using the sigmoid function applied to this linear combination.
+This method accounts for the uncertainty in the model parameters, providing a probabilistic framework for decision-making.
+
+The prediction \(\hat{y}\) returned by the model is the prototype of the reached leaf node, where the prototype is defined as the mean of the target values associated with the training instances that were routed to that leaf during training.
+
+\begin{equation}
+\hat{y} = \frac{1}{|\mathbf{Y}|} \sum_{i=1}^{|\mathbf{Y}|} y_i
+\end{equation}
+where \( y_i \) represents the target value of the \(i\)-th instance in \(\mathbf{Y}\).
+This value $\hat{y}$ serves as the prediction for any new instance that reaches the particular leaf node.
+
+\begin{algorithm}[h!]
+\caption{Prediction Process in \gls{vspyct} using Monte Carlo Sampling}
 \label{alg:make_pred}
@@
 
-\State \textbf{Input:} Feature vector $\mathbf{x}$, decision tree $T$ rooted at node $\Omega$
+\State \textbf{Input:} Feature vector $\mathbf{x}$, decision tree $T$ rooted at node $\Omega$, number of samples $M$
 \State \textbf{Output:} Prediction $\hat{y}$
 
-\Function{Predict}{$\Omega$, $\mathbf{x}$}
-    \If{$\Omega$ is a leaf}
-        \State \Return $\Omega.prototype$
+\Function{MC\_Predict}{$\Omega$, $\mathbf{x}$, $M$}
+    \State Initialize $\hat{y}_{\text{sum}} = 0$
+    \For{$m = 1$ to $M$}
+        \State $\hat{y}_{\text{sum}} \gets \hat{y}_{\text{sum}} + \text{Predict}(\Omega, \mathbf{x}, m)$
+    \EndFor
+    \State \Return $\hat{y} = \frac{\hat{y}_{\text{sum}}}{M}$
+\EndFunction
+
+\Function{Predict}{$\Omega$, $\mathbf{x}$, $m$}
+    \If{$\Omega$ is a leaf node}
+        \State \Return $\hat{y}_\Omega = \frac{1}{|\mathcal{D}_\Omega|} \sum_{i \in \mathcal{D}_\Omega} y_i$
     \Else
-        \State $\Theta \gets$ \Call{SampleParameters}{$\Omega.split\_model$, $\Omega.guide$}
-        \State Define sigmoid function $\rho(\Theta, \mathbf{x}) = \frac{1}{1 + \exp(-\Theta^\top \mathbf{x})}$
-        \If{$\rho(\Theta, \mathbf{x}) \leq 0.5$}
-            \State \Return \Call{Predict}{$\Omega.left$, $\mathbf{x}$}
+        \State Sample $\Theta^{(m)} = (\mathbf{w}^{(m)}, b^{(m)})$ from $q(\mathbf{w}, b|\mathcal{D})$
+        \State Compute $\rho(\Theta^{(m)}, \mathbf{x}) = \sigma(\mathbf{w}^{(m)\top} \mathbf{x} + b^{(m)})$
+        \If{$\rho(\Theta^{(m)}, \mathbf{x}) \leq 0.5$}
+            \State \Return \Call{Predict}{$\Omega_\text{left}$, $\mathbf{x}$, $m$}
         \Else
-            \State \Return \Call{Predict}{$\Omega.right$, $\mathbf{x}$}
+            \State \Return \Call{Predict}{$\Omega_\text{right}$, $\mathbf{x}$, $m$}
         \EndIf
@@
 
-\Function{SampleParameters}{model, guide}
-    \State Let $\Theta = (\mathbf{w}, b)$ where:
-    \State $\mathbf{w}$ are weights sampled from variational posterior $q_{\omega}(\mathbf{w}|\mathcal{D})$
-    \State $b$ is bias sampled from variational posterior $q_{\omega}(b|\mathcal{D})$
-    \State \Return $\Theta$
-\EndFunction
-
-\State Compute prediction $\hat{y}$ by invoking \Call{Predict}{$\Omega$, $\mathbf{x}$}
+\State \(\hat{y} \gets \) \Call{MC\_Predict}{$\Omega$, $\mathbf{x}$, $M$}
 \State \textbf{Return} $\hat{y}$
@@
 
-In this section, we analyse the time complexity of learning a split in \gls{vspyct} and compare it with \gls{spyct}.
+We analyse the time complexity of learning a split in \gls{vspyct} and compare it with \gls{spyct}.
 Our analysis focuses on the computational expense associated with determining a split, as this represents the principal distinction between the two approaches.
 In the predictive clustering framework, the set of clustering attributes $K$, which contributes to the variance calculation, encompasses both the features and the target variables.
-As a result, $K$ equals $N + T$, with $N$ representing the feature count and $T$ denoting the target count.
-The time complexity for learning in \glspl{spyct} can be expressed as $\mathcal{O}(NI_o(D+K))$, wherein $I_o$ stands for the iteration count needed for optimisation.
-Notably, the computational demand for the \glspl{spyct}' gradient-based version increases linearly with the attribute count.
+As a result, $K$ equals $D + T$, with $D$ representing the feature count and $T$ denoting the target count.
+The time complexity for learning in \glspl{spyct} can be expressed as $\mathcal{O}(NI_o(D+K))$, wherein $N$ is the number of training instances, and $I_o$ is the iteration count needed for optimisation~\citep{Stepi_nik_2021}.
+Notably, the computational demand for the \glspl{spyct}'s gradient-based version increases linearly with the attribute count.
 
-The introduction of variational Bayes method in the architecture of \gls{spyct} to create \gls{vspyct} adds a new layer of computational complexity.
-Specifically, the variational approach requires iterating over the dataset multiple times to optimise the ELBO for each split.
-This iterative process involves sampling from the posterior distributions of split parameters, evaluating the ELBO, and updating the parameters accordingly.
-Consequently, the time complexity for learning a split in \gls{vspyct} can be described as $\mathcal{O}(MI_oNI_{vb})$, where $M$ is the number of Monte Carlo samples used to approximate the posterior distributions, $I_{vb}$ represents the number of iterations required for variational inference to converge, and the other variables retain their meanings from the \gls{spyct} context.
+The introduction of variational Bayes method in the architecture of the model adds a new layer of computational complexity.
+Specifically, the complexity of learning a split in \gls{vspyct} is influenced by the following factors: the need to sample parameter sets from the approximative posterior distributions, the iterative optimisation required to refine these estimates towards minimising the \gls{elbo}, and the dimensionality imposed by the number of features.
+The combined computational complexity for learning a split in \gls{vspyct} can thus be expressed as $\mathcal{O}(MNI_{vb}(D+K))$, where:
+\begin{itemize}
+\item $M$ represents the number of Monte Carlo samples needed to approximate the posterior distributions adequately.
+\item $N$ is the number of data points evaluated during each iteration.
+\item $D$ indicates the number of features, each contributing to the parameter space that must be sampled and optimised.
+\item $K$ indicates the number of clustering attributes, used to calculate the variance in each of the node.
+\item $I_{vb}$ indicates the number of iterations required for the variational inference process to achieve convergence.
+\end{itemize}
+This complexity reflects the multiplicative impact of sampling, feature dimensionality, and iterative optimisation, highlighting how the variational approach scales with the size and feature richness of the dataset.
 
-Comparing \glspl{vspyct} to \glspl{spyct}, the introduction of variational Bayes method increases the computational load due to the necessity of Monte Carlo sampling and iterative ELBO optimisation.
-However, this additional complexity is counterbalanced by the enhanced predictive performance, model interpretability, and uncertainty quantification offered by \glspl{vspyct}.
-%While \glspl{spyct} are computationally efficient due to their linear scaling with the number of attributes and utilisation of gradient-based optimisation, \glspl{vspyct} leverage the power of variational inference to provide a deeper understanding of the data and model behaviour, even if at the cost of increased time complexity.
+Comparing \glspl{vspyct} to \glspl{spyct}, the VB method increases the computational load due to the necessity of Monte Carlo sampling and iterative \gls{elbo} optimisation.
+However, this additional complexity is counterbalanced by the enhanced model interpretability, and uncertainty quantification offered by \glspl{vspyct}.
 
-The choice between using \glspl{vspyct} over \glspl{spyct} hinges on the specific requirements of the application scenario.
-For tasks where model interpretability and uncertainty quantification are paramount, and computational resources are sufficient, \glspl{vspyct} offer a compelling advantage.
-Conversely, for applications where computational efficiency is critical, and the interpretability requirements are less stringent, \glspl{spyct} may still be favourable.
+\subsection{Feature Importance}
 
-\subsection{Feature importance}
+The procedure for computing feature importance scores in \gls{vspyct} is inspired by the methodology introduced in the original \gls{spyct} model~\citep{Stepi_nik_2021}, while extending it to incorporate the model’s probabilistic nature.
+Let \(\mathbb{E}[\mathbf{w}_s]\) and \(\text{Var}(\mathbf{w}_s)\) denote the element-wise mean and variance of the weights at node \(s\). We define:
+\[
+\mathbf{u}_s = \frac{\mathbb{E}[\mathbf{w}_s]}{\text{Var}(\mathbf{w}_s) + \epsilon}
+\]
+where \(\epsilon\) is a small constant added for numerical stability.
 
-Feature importance is derived from the aggregated impact of features across all splits, quantified by examining the variational posterior distributions of the weights.
-The importance of each feature is proportional to its contribution to reducing impurity across splits.
+The feature importance vector \(\text{Imp}(T)\) is then computed as:
+\begin{equation}\label{eq:fimp}
+\text{Imp}(T) = \sum_{s \in T} \left( \frac{s_n}{N} \right) \cdot \frac{|\mathbf{u}_s|}{\|\mathbf{u}_s\|_2}
+\end{equation}
+Here, \(\frac{s_n}{N}\) weights the contribution of node \(s\) by the fraction of training samples passing through it, and \(\|\mathbf{u}_s\|_2\) is the L2 norm of \(\mathbf{u}_s\).
+This formulation integrates both the magnitude and the uncertainty of the weight parameters, potentially yielding a more robust measure of feature importance under a probabilistic model.
+The vector $\mathbf{w}_s$ specifies the weights that define the oblique split at node $s$.
+The expected value of the weights, $\mathbb{E}[\mathbf{w}_s]$, is computed through Monte Carlo simulation, where weights are sampled $M$ times from their approximation of the posterior distribution, and the mean of these samples is taken as the representative value for each weight.
+This integrates the stochastic nature of the model parameters into the calculation of feature importance.
+The repeated sampling and averaging process ensures that the estimated importance reflects both the central tendency and the variability of the weight parameters, crucial for interpreting the model's decision-making process under uncertainty.
 
-Given a \gls{vspyct} model, the feature importances are derived from the variational parameters defining the split decisions across the tree.
-The essence of the computation lies in evaluating the expected influence of each feature, taking into account the probabilistic nature of the model parameters.
-This is encapsulated by the following formula:
+\paragraph{Toy Example}
+Consider a simple dataset where two features, \textit{age} and \textit{salary}, predict the likelihood of purchasing a car.
+In a \gls{vspyct} model, the influence of each feature at a node depends not only on the magnitude of its mean weight but also on how this mean compares to the feature’s weight variance.
 
-\begin{equation}
-\text{Imp}(T) = \sum_{s \in T} \left( \frac{s_n}{N} \right) \left( \frac{\mathbb{E}[\mathbf{w}_s]}{|\mathbb{E}[\mathbf{w}_s]|_1} \right)
-\end{equation}
+Suppose that at the root node, the mean weights associated with \textit{age} and \textit{salary} are \( \mathbb{E}[w_{\text{age}}] = 0.2 \) and \( \mathbb{E}[w_{\text{salary}}] = 0.8 \), and their respective variances are \( \text{Var}(w_{\text{age}}) \) and \( \text{Var}(w_{\text{salary}}) \).
+The importance at this node will now reflect the ratio \( \frac{\mathbb{E}[w]}{\text{Var}(w)} \), such that a feature with a relatively high mean weight but also a high variance will not necessarily dominate one with a slightly lower mean but substantially more stable (lower variance) weights.
+Thus, if \textit{salary} maintains a higher \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio than \textit{age}, it indicates that \textit{salary} is not only influential but also more reliably so at this particular node.
 
-where $T$ represents the variational oblique predictive clustering tree, $s$ iterates over the set of split nodes within $T$, $s_n$ is the number of training examples that passed through node $s$, $N$ denotes the total number of training examples used in the tree, $\mathbf{w}_s$ symbolizes the weight vector defining the oblique split at node $s$, treated as a random variable with a Gaussian distribution, $\mathbb{E}[\mathbf{w}_s]$ signifies the expected value (mean) of the weight vector $\mathbf{w}_s$, calculated from its posterior distribution as inferred by the variational Bayesian method.
-The normalization factor $|\mathbb{E}[\mathbf{w}_s]|_1$ ensures the comparability of feature importances by scaling them relative to the aggregate magnitude of expected weights.
-This formulation accounts for the stochastic nature of each parameter, reflecting the central tendency of the distributions defining the split hyperplanes.
+However, the overall feature importance is derived by aggregating these normalized ratios across all splits within the tree.
+Even if \textit{salary} appears strongly influential at the root node, the final importance calculation considers both the magnitude and the stability of the weights across all nodes.
+A feature consistently demonstrating a favorable \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio throughout multiple splits will be deemed more important globally.
+In this way, the final importance assessment accounts for both the strength and certainty of each feature’s contribution to the model’s decision-making process.
 
-For practical computation, the expected weights $\mathbb{E}[\mathbf{w}_s]$ are obtained by sampling from the posterior distributions inferred during the learning process.
-Each weight vector $\mathbf{w}_s$ is drawn such that $\mathbf{w}_s \sim q(\mathbf{w}_s|\mathcal{D})$, where $q(\mathbf{w}_s|\mathcal{D})$ represents the variational approximation to the posterior distribution of weights given the data $\mathcal{D}$.
-The importance of each feature is then a function of these expected weights, modulated by the frequency of training instances influencing each split, thereby emphasising the splits that impact a larger portion of the data.
-This mathematical framework for computing feature importances in \gls{vspyct} not only highlights the contribution of individual features to the model's decision-making process but also integrates the inherent uncertainty of these contributions, offering a richer, more nuanced understanding of feature relevance within a probabilistic modelling context.
+\section{Experimental setting}
 
-\section{Data and evaluation metric}
+The experimental evaluation was conducted to compare the \gls{vspyct} model with several benchmarks, including the \gls{spyct} model, \glspl{opct}, and ensemble of \glspl{pct}.
+Specifically, we considered both a single \gls{spyct} and an ensemble of \glspl{spyct}, as well as a single \gls{opct} and an ensemble of \glspl{pct}.
+We chose these methods as primary comparisons because they have been extensively evaluated in the literature, demonstrating strong performance across various \gls{sop} tasks~\citep{Andonovikj_2024}.
+
+The comparison encompassed a range of predictive modelling tasks, including single-target regression, multi-target regression, binary classification, and multi-class classification.
+By evaluating not only a single \gls{spyct} tree, but also an ensemble of \glspl{spyct}, a single \gls{opct} tree, and an ensemble of \glspl{pct}, we aimed to assess how \gls{vspyct}'s predictive performance compares against a spectrum of models, from single decision trees to more complex ensembles and option-based trees.
+This comprehensive approach allowed us to rigorously evaluate \gls{vspyct}'s effectiveness across different modelling scenarios and relative to both well-established baseline models and state-of-the-art tree-based methods.
 
@@
 
-\subsection{R Survival datasets}
-
-Our evaluation encompasses four benchmark survival analysis datasets, each serving as a standard reference within the domain and accessible via the survival R package~\citep{survival-package}.
-
-\begin{enumerate}
-\item \textit{Retinopathy Dataset}: Investigates the efficacy of laser coagulation in delaying vision loss due to diabetic retinopathy among 197 patients, with survival times adjusted for a minimum event horizon of 6.5 months.
-\item \textit{Lung Dataset}: Analyses survival data for patients with advanced lung cancer, focusing on their lifespan post-diagnosis and daily activity performance levels.
-\item \textit{Primary Biliary Cirrhosis (PBC) Dataset}: Examines the progression of primary biliary cholangitis in 424 patients, with data drawn from a clinical trial conducted between 1974 and 1984, exploring disease progression towards cirrhosis.
-\item \textit{Heart Dataset}: Studies the survival of patients on the waiting list for heart transplants at Stanford, evaluating the length of time patients remain on the waiting list.
-\end{enumerate}
+The experiments were conducted on a diverse set of datasets, encompassing different types of predictive modelling tasks, including single-target regression, multi-target regression, and classification (binary and multi-class).
+The general characteristics of the datasets are summarised in Table~\ref{tab:datasets}.
+The table includes the number of examples \(N\), the number of features \(D\), the number of targets \(T\), and the number of classes \(C\), with \(C\) being applicable only to classification tasks.
 
@@
 \centering
-\caption{Main properties of the datasets. Censoring shows the percentage of censored observations.}
-\adjustbox{max width=\textwidth}{
-\begin{tabular}{llll}
+\caption{Summary of the datasets used in the experiments.}
+\label{tab:datasets}
+\begin{tabular}{@{}lcccccc@{}}
 \toprule
-Dataset& \# of examples& \# of features& Censoring\\
-\midrule
-Mayo Clinic Primary Biliary Cholangitis&312&17&60\%\\
-NCCTG Lung Cancer&228&8&\\
-Stanford Heart Transplant&172&4&\\
-Retinopathy&394&9&\\
-\bottomrule
+\textbf{Dataset} & \textbf{Type of Task}          & \textbf{N} & \textbf{D} & \textbf{T} & \textbf{C} \\ \midrule
+rf1~\citep{mulan}             & Multi-target regression        & 9125         & 64         & 8         &        \\
+rf2~\citep{mulan}             & Multi-target regression        & 9125         & 576         & 8         &        \\
+atp1d~\citep{mulan}           & Multi-target regression        & 337         & 411         & 6         &        \\
+atp7d~\citep{mulan}           & Multi-target regression        & 296         & 411         & 6         &        \\
+scm1d~\citep{mulan}           & Multi-target regression        & 9803         & 280         & 16         &        \\
+house\_8L~\citep{openml}       & Single-target regression       &  22784        & 8         & 1          &        \\
+puma8NH~\citep{openml}         & Single-target regression       & 8192         & 8         & 1          &        \\
+diabetes~\citep{openml}        & Binary classification          & 768         & 8         & 1          & 2          \\
+banknote~\citep{openml} & Binary classification          & 1372         & 4         & 1          & 2          \\
+gas-drift~\citep{openml}  & Multi-class classification     & 13910         & 128         & 1          & 6         \\
+balance~\citep{openml}         & Multi-class classification     & 625         & 4         & 1          & 3         \\ \bottomrule
 \end{tabular}
-}
-\label{table:dataset}
-\end{table}%
+\end{table}
 
-\subsection{Unemployment dataset}
-The dataset we consider consists of $74,086$ instances and contains anonymised personal and professional characteristics of jobseekers that have entered the Slovenian \gls{pes}.
-The dataset is complex because its attributes come in different forms (categorical, numerical, date and time), and most have to undergo transformation.
-An overview of the dataset and its features is given in \tablename~\ref{table:dataset}.
+For multi-target regression, the target variable is a vector, while for single-target regression, it is a scalar.
+In binary classification, there are two possible classes, and for multi-class classification, the number of classes is dataset-dependent.
 
-\begin{table}[h!]
-\centering
-\caption{Features of the dataset and their cardinalities \citep{Andonovikj_2024}}
-\begin{tabular}{ll}
-\toprule
-Continuous features& Discrete features\\
-\midrule
-Day of PES entry&Education direction (30)\\
-Months of work experience&Occupation (ISCO) (123)\\
-Age&Reason for PES entry (5)\\
-Month of PES entry&Reason for termination (43)\\
-&Social benefits (2)\\
-&Gender (2)\\
-&eApplication (2)\\
-&Unemployment benefits (2)\\
-\bottomrule
-\end{tabular}
-\label{table:dataset}
-\end{table}%
+\subsection{Evaluation}
 
-The general structure of the jobseekers' attributes is described by dividing the attributes into several prominent groups: socioeconomic variables (gender, age), information on job readiness (education, health limitations), and opportunities (regional labour market development), and all available labour market history information, such as prior work experience.
-The target variable is in numeric form.
-It is a counter of days till a jobseeker gets employed or gets out of the study and leaves no further information about the jobseeker's employment status.
+The performance of the models was evaluated using appropriate metrics for each predictive modelling task.
+For single-target regression, the Mean Absolute Error (\textit{MAE}) was employed as the evaluation metric:
+\[
+\textit{MAE} = \frac{1}{N} \sum_{i=1}^{N} \left| y_i - \hat{y}_i \right|
+\]
+where \( y_i \) denotes the true value, \(\hat{y}_i\) is the predicted value, and \(N\) is the total number of instances. A lower \textit{MAE} indicates better predictive accuracy.
 
-\figurename~\ref{fig:grid_prob} shows the distributions of some features in the dataset.
-The sample of jobseekers is balanced in terms of gender.
-Most jobseekers enter the \gls{pes} without any working experience. 
-As expected, the age distribution is between 18 and 64 since this is the legal age interval of the labour-active population.
-It should be noted that the age distribution is bi-modal.
-The first peak is around 23-26 years since this is the expected graduation age for a university degree.
-The second peak is just before 60 years of age, i.e. jobseekers who became unemployed just before retirement due to various reasons.
-The final histogram shows that the dataset contains 30\% of right censored data.
+For multi-target regression, a normalized Mean Absolute Error (\textit{NMAE}) was used to ensure comparability across targets with different value ranges.
+Letting \(y_{ij}\) and \(\hat{y}_{ij}\) be the true and predicted values for the \(j\)-th target of the \(i\)-th instance, and \(T\) be the total number of target variables, the \(\textit{NMAE}\) is defined as:
+\[
+\textit{NMAE} = \frac{1}{T} \sum_{j=1}^{T} \left( \frac{\frac{1}{N} \sum_{i=1}^{N} | y_{ij} - \hat{y}_{ij} |}{\max_i(y_{ij}) - \min_i(y_{ij})} \right)
+\]
+If \(\max_i(y_{ij}) - \min_i(y_{ij}) = 0\) for a given target \(j\), this denominator is replaced by 1, effectively leaving that target unscaled.
+
+A lower \(\textit{NMAE}\) indicates better predictive performance, as it accounts for the range of each target variable to produce a comparable error measure across multiple targets.
+
+For binary and multi-class classification tasks, the \textit{F1} score was used to measure model performance.
+In the case of multi-class classification, the \textit{F1} score was macro-averaged across all classes.
+A higher \textit{F1} score indicates better classification performance, as it reflects a balance between precision and recall.
+
+All evaluations were conducted using 5-fold cross-validation, to ensure a comprehensive assessment of the model's performance across different subsets of the data, reducing the potential for bias and overfitting.
+
+\section{Results}
+
+Figure~\ref{fig:avg_rank} shows the average rankings of the evaluated methods—\gls{vspyct}, \gls{opct}, single \gls{spyct}, ensemble of \glspl{spyct}, and ensemble of \glspl{pct}—across four task types: binary classification (BC), multi-class classification (MCC), single-target regression (STR), and multi-target regression (MTR).
 
@@
     \centering
-    \includegraphics[width=0.95\textwidth]{feature_distributions.pdf}
-    \caption{Distributions of some of the dataset features \citep{Andonovikj_2024}}
-    \label{fig:grid_prob}
+    \subfloat[BC]{%
+        \includegraphics{critical_distance_bc.pdf}
+        \label{fig:subfig1}
+    }
+    \subfloat[MCC]{%
+        \includegraphics{critical_distance_mcc.pdf}
+        \label{fig:subfig2}
+    } \\
+    \subfloat[STR]{%
+        \includegraphics{critical_distance_str.pdf}
+        \label{fig:subfig3}
+    }
+    \subfloat[MTR]{%
+        \includegraphics{critical_distance_mtr.pdf}
+        \label{fig:subfig4}
+    }
+    \caption{Average rank of predictive performance. Lower ranks indicate better performance.}
+    \label{fig:avg_rank}
 \end{figure}
 
-The categorical features are diverse with respect to their cardinality.
-The categories of the features with high cardinality ($>1000$ different class values per feature) are transformed using domain knowledge and the number of class values is significantly reduced.
-Because the \emph{ISCO} standard and \emph{Education direction} provide a hierarchical classification of occupations and education directions, we took the higher-level classifications and thus significantly reduced the number of categories.
-The remaining categorical features were encoded using one-hot encoding.
-The temporal (cyclical) features, such as ``Day of PES entry" and ``Month of PES entry" were transformed by calculating the feature's $\sin$ and $\cos$ components.
-In this way, the artificially significant difference between, for example, the last day of the previous month and the first day of the current month is removed, and the value of the feature can be viewed as a point $\left(\sin(x), \cos(x)\right)$ of a circle.
+In BC tasks, \gls{vspyct} ranks first, outperforming ensembles of \glspl{spyct} and \glspl{pct}, which tie for second.
+This suggests that \gls{vspyct}’s probabilistic modeling effectively handles simple binary boundaries, surpassing both ensemble variance reduction and structural options (\gls{opct}).
+In MCC tasks, \gls{vspyct} and the ensemble of \glspl{spyct} lead, followed by the ensemble of \glspl{pct}, single \gls{spyct}, and \gls{opct}, indicating that \gls{vspyct}’s uncertainty-aware splits can handle more complex class structures as well as ensembles can.
 
-\subsection{Evaluation metric}
+For STR tasks, both ensemble-based approaches—those combining multiple \glspl{spyct} and those combining multiple \glspl{pct}—achieve the highest ranks, followed by \gls{vspyct}.
+It can be seen that ensembles appear particularly effective for precise single-target continuous predictions.
+However, in MTR scenarios, \gls{vspyct} surpasses the ensemble of \glspl{pct}, indicating that its probabilistic representation of oblique splits can better capture complex inter-target relationships.
+By contrast, \gls{opct} and single \gls{spyct} lag behind, indicating that their individual enhancements—option nodes or a single-tree oblique structure—do not offer sufficient complexity or flexibility to match the performance gains achieved by either ensembles or the probabilistic modelling of \gls{vspyct}.
 
-The efficacy of survival models is predominantly assessed through the Concordance Index~\citep{brentnall2018use} and the Brier score~\citep{gerds2006consistent}.
-While the Concordance Index offers valuable insights, its inability to account for the error magnitude in predicted probabilities limits its utility.
-The Brier score addresses this gap by factoring in the error magnitude, making it a more comprehensive metric for our experimental evaluation.
-The Brier score assesses the precision of a forecasted survival function at a specific time $t$, quantifying the mean squared differences between observed survival statuses and forecasted survival probabilities.
-It ranges from $0$ to $1$, where a score of $0$ indicates perfect accuracy.
-This metric not only assesses the precision of time-to-event forecasts, akin to a squared error loss, but also necessitates adjustments for right-censored samples in the dataset.
-This adjustment employs the inverse probability of censoring weights method to weight the squared differences.
+\subsection{Discussion}
 
-The conditional survival function estimator for censoring times, $\hat{G}(t)=P(C>t)$, utilises the Kaplan-Meier method, with $C$ denoting the censoring time.
-The Brier score is thus calculated as~\citep{trickbrier}:
+The results underscore the interplay between model complexity, uncertainty quantification, and variance reduction.
+Ensemble methods consistently perform well in scenarios demanding high precision (e.g., STR) or broad generalisation over multiple outputs (MTR).
+\Gls{vspyct}, capitalises on its Bayesian inference framework to excel in tasks where modeling uncertainty and complex decision boundaries are crucial (e.g., BC and MCC), and remains competitive even in more challenging regression contexts.
 
-\begin{equation}
-BS(t) = \frac{1}{M}\sum\limits_{i=1}^N \left(\frac{\hat{S}(t_i,\mathbf{X}i)\cdot1{T_i\geq{t},\delta_i=1}}{\hat{G}({T_i}^-)}+\frac{(1-\hat{S}(t_i,\mathbf{X}i))\cdot1{T_i>t, \delta_i=0}}{\hat{G}(T_i)}\right)
-\label{eq:bs}
-\end{equation}
+These findings underscore that \gls{vspyct} is not a universal replacement for ensembles, but rather a complementary alternative whose benefits depend on the task at hand.
+When dealing with classification problems—binary or multi-class—where interpretability and uncertainty quantification are critical, \gls{vspyct} offers a potent blend of predictive performance, clarity, and inherent uncertainty modeling.
+Its capability to handle intricate class boundaries and its flexibility in representing parameter uncertainties make it a valuable method for domains demanding transparency and robustness.
 
-Here, $M$ signifies the count of test instances, $\mathbf{X}_i$ the feature vector for each instance, $\hat{S}(t, \mathbf{X}_i)$ the predicted survival probability at time $t_i$, $\delta_i$ a boolean indicator, and $T_i$ and $T_i^-$ the survival times for $\delta_i$ being 0 and 1, respectively.
-The Brier score is computed at various timestamps to evaluate how the method's performance varies with predictions extended into the future.
+In continuous-output scenarios, ensembles still exhibit an edge, suggesting that if raw predictive precision is paramount and interpretability or uncertainty modeling are secondary considerations, ensemble models may remain the preferred choice.
+Nevertheless, \gls{vspyct} provides a single-tree model that narrows the performance gap considerably and is particularly appealing when balancing predictive accuracy with the need for more direct insight into the model’s decision process.
 
-The performance was estimated by using $k$-fold cross-validation with the parameter $k$ set to 5~\citep{elstatl}.
-The prediction model yields an estimated survival curve for each individual in the study.
-
-\section{Results}
-
-The resulting BS curves are shown in Figure~\ref{fig:brier_scores}.
-It should be noted that lower BS indicates a more accurate forecast.
-The analysis revealed distinct performance patterns between the two models across different datasets.
-Specifically, in the PBC and lung cancer datasets, \gls{vspyct} outperformed \gls{spyct}, evidenced by consistently lower Brier scores throughout the observation period.
-This superior performance of \gls{vspyct} could be attributed to its robust handling of parameter uncertainty through the VB method, which likely improved its predictive accuracy in these contexts.
-The \gls{vspyct} model demonstrated a particularly notable advantage in maintaining lower peak Brier scores, suggesting enhanced reliability in predicting survival outcomes for patients with PBC and lung cancer.
-
-Conversely, in the heart transplant and retinopathy datasets, \gls{spyct} showed better performance compared to \gls{vspyct}, as indicated by lower Brier scores over time.
-This suggests that the simpler, more direct approach of \gls{spyct} may be more effective in certain contexts, possibly due to less complex data structures or the nature of the underlying survival processes in these specific conditions.
-The lower Brier scores achieved by \gls{spyct} in these cases highlight its efficiency in accurately predicting the survival outcomes of heart transplant patients and those undergoing treatment for diabetic retinopathy.
-
-% average rank diagram - zemi go kodot od Matej Petkovic
+\subsection{Feature importance scores}
 
@@
     \centering
-\subfloat[Lung]{%
-\includegraphics{bs_r_cancer.pdf}
-\label{fig:bs_cancer}}
-\quad % Adds some space between the figures in the same row
-\subfloat[Retinopathy]{%
-\includegraphics{bs_r_retinopathy.pdf}
-\label{fig:bs_retinopathy}}\\ % Ends the first row and starts a new one
-\subfloat[Heart]{%
-\includegraphics{bs_r_heart_truncated.pdf}
-\label{fig:bs_heart}}
-\quad % Adds some space between the figures in the same row
-\subfloat[PBC]{%
-\includegraphics{bs_r_pbc.pdf}
-\label{fig:bs_pbc}}
+    \includegraphics{../figures/feat_importances.pdf}
+    \caption{Feature importance scores obtained on a single dataset from per category. The results indicate that the model does not depend on irrelevant features.}
+    \label{fig:feature_importances}
+\end{figure}
 
-\caption{Performance evaluation of \gls{vspyct} and \gls{spyct} models measured through the Brier score on the R survival datasets}
-\label{fig:brier_scores}
-\end{figure}
+As illustrated in Figure~\ref{fig:feature_importances}, the proposed \gls{vspyct} model yields meaningful feature importance scores.
+To rigorously evaluate this property, we introduced synthetic noise features into the datasets before model training.
+Let the original dataset contain \(d\) features.
+We then augmented it by adding \(d\) additional features, each generated by random noise.
+Thus, the total number of features in the augmented dataset is \(2d\).
+
+The resulting feature importance scores reveal that the original (real) features exhibit a broad range of importance values.
+This heterogeneity is expected, as real-world datasets often comprise both highly informative variables and those with negligible predictive power.
+In contrast, the artificially introduced noise features consistently received low importance scores.
+Formally, if we let \(\mathcal{F} = \{f_1, f_2, \ldots, f_d\}\) denote the set of original features, and \(\mathcal{F}' = \{f_{d+1}, \ldots, f_{2d}\}\) the set of noise features, the estimated importances \(I(f_j)\) for \(f_j \in \mathcal{F}'\) are significantly lower than those for at least a subset of \(\mathcal{F}\).
+This indicates that the \gls{vspyct}-derived importance measures are not only able to identify relevant features but also robust to the inclusion of spurious, non-informative features.
 
@@
 
-%slika od drvoto so oboeni edges
+The dataset used for the interpretability analysis consists of $74,086$ anonymised instances of jobseekers from the Slovenian Public Employment Service (PES).
+We will refer to it as the Unemployment dataset.
+It includes a variety of personal and professional characteristics, such as age, gender, education, and work experience.
+The target variable is the time until the jobseeker either becomes employed or exits the study, which is measured in days.
+The dataset is challenging due to the different forms of its attributes (categorical, numerical, temporal) and the presence of censored data, where some jobseekers’ outcomes are not fully observed.
 
-Figure~\ref{fig:inter_tree} presents the visual representation of the \gls{vspyct} applied to the Unemployment dataset.
-Based on that, one can distill a concise interpretation enriched with practical insights.
+The goal is to predict the time-to-event (employment) for jobseekers, with some records being right-censored—meaning the exact time of the event is not observed.
+Each data instance is described by features \(\mathbf{X}_i\), observed time \(t_i\), and a censoring indicator \(\delta_i\), where \(\delta_i = 1\) indicates the event was observed, and \(\delta_i = 0\) indicates a right-censored record.
+This setup, with its inherent missing data due to censoring, makes it suitable for semi-supervised learning approaches.
+Detailed information on the dataset and the semi-supervised multi-target regression framework can be found in~\cite{Andonovikj_2024}.
+
+Figure~\ref{fig:pred_ci} presents the survival curve predicted by the \gls{vspyct} model for a specific jobseeker in the Unemployment dataset.
+The curve shows the estimated probability of remaining unemployed over time, with the blue line representing the mean prediction.
+The shaded area around the curve illustrates the uncertainty in the prediction, captured as a confidence interval between the 10th and 90th percentiles.
+This visualisation effectively communicates the model's predictions along with its confidence, providing valuable insights into the jobseeker's likelihood of finding employment over time.
+The dashed vertical line marks the actual observed time of employment, allowing for an immediate comparison between the predicted and actual outcomes.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1.0\textwidth]{pred_with_ci.pdf}
+    \caption{Survival curve predicted by the \gls{vspyct} model on the Unemployment dataset.
+The blue line represents the estimated mean survival time, while the shaded area indicates the confidence interval (10th to 90th percentile) around the prediction.
+The vertical dashed line marks the actual time of the observed event.}
+    \label{fig:pred_ci}
+\end{figure}
+
+Figure~\ref{fig:feature_importance} highlights the feature importance scores obtained from the \gls{vspyct} model, focusing on the most influential features that contribute to predicting the unemployment duration.
+The horizontal bars show the relative importance of each feature, with \textit{Months of work experience} emerging as the most critical factor, followed by \textit{Age} and \textit{Entry month}.
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics[width=1.0\textwidth]{feature_importance.pdf}
+    \caption{Feature importance scores extracted from the \gls{vspyct} model applied to the Unemployment dataset.
+The horizontal bars represent the relative importance of each feature, with the most influential feature being "Months of work experience."}
+    \label{fig:feature_importance}
+\end{figure}
 
@@
 \end{figure}
+
+Figure~\ref{fig:inter_tree} presents the visual representation of the \gls{vspyct} applied to the Unemployment dataset.
+Based on that, one can distill a concise interpretation enriched with practical insights.
 
@@
 
-We introduced the \gls{vspyct} model, an approach that strategically combines the benefits of oblique splits, variational Bayes, and Bayesian inference within a single, interpretable framework.
+We introduced the \gls{vspyct} model, an approach that combines the benefits of oblique splits, variational Bayes, and Bayesian inference within a single, interpretable predictive framework.
 The \gls{vspyct} model addresses the trade-off between the performance of ensemble models and the interpretability of single decision trees, and the absence of inherent uncertainty quantification in predictions.
-Our results demonstrate that the \gls{vspyct} model is competitive to, and on some data sets exceeds the performance of ensemble of \glspl{spyct}, without sacrificing the model's interpretability.
+Our results demonstrate that the \gls{vspyct} model is competitive to, and on some datasets exceeds the performance of ensemble of \glspl{spyct}, without sacrificing the model's interpretability.
 By employing variational Bayes method, we introduce a mechanism for uncertainty quantification directly into the decision-making process, enhancing the reliability and applicability of the model across various domains where decision certainty is paramount.
@@
 The model opens new possibilities for research in semi-supervised learning and Bayesian inference, promising to bridge the gap between theoretical advancements and practical applications.
-Future work will focus on further refining the model's performance, and extending its applications in diverse \gls{sop} settings.
+Future work will focus on further refining the model's performance, and extending its applications in diverse predictive tasks.
 
+\begin{comment}
 \section*{Acknowledgement}
 The authors acknowledge the research core funding No.\ P2-0001,  P2-0103 and P2-0383 financially supported by the Slovenian Research Agency.
-The authors also acknowledge the funding received from the European Union's Horizon 2020 research and innovation programme project HECAT under grant agreement No.\ 870702 .
+The authors also acknowledge the funding received from the European Union's Horizon 2020 research and innovation programme project HECAT under grant agreement No.\ 870702.
+\end{comment}
 
@@
 
+\begin{comment}
+\section*{Appendix}
+
+The evaluation scores on the regression datasets are presented in Table~\ref{tab:regression_results}.
+
+\begin{table}[h!]
+\centering
+\caption{\textit{MAE} Scores for Regression Datasets}
+\label{tab:regression_results}
+\begin{tabular}{@{}lccc@{}}
+\toprule
+Dataset & SPYCT-single tree & SPYCT-ensemble & VSPYCT \\ \midrule
+rf1             & 29.38                            & \textbf{29.34}                        & 29.37                \\
+rf2             & 29.37                            & 29.49                        & \textbf{29.36}                \\
+atp1d           & 95.17                            & \textbf{70.93}                        &  77.24               \\
+atp7d           & 115.54                            & \textbf{73.08}                        & 91.28                \\
+scm1d           & 205.99                            & 206.01                        & \textbf{205.72}                \\
+house\_8L       & 25378.02                            & \textbf{19817.56}                        & 22992.24                \\
+puma8NH         & 2.80                            & \textbf{2.62}                        & 2.74                \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[h!]
+\centering
+\caption{F1 Scores for Classification Datasets}
+\label{tab:classification_results}
+\begin{tabular}{@{}lccc@{}}
+\toprule
+Dataset & SPYCT-single tree & SPYCT-ensemble & VSPYCT \\ \midrule
+diabetes              & 0.53                            & \textbf{0.60}                        & 0.59                \\
+banknote       & 0.97                            & \textbf{0.99}                        & 0.98                \\
+gas-drift        & 0.98                            & 0.99                        & \textbf{0.99}                \\
+balance               & 0.60                            & 0.66                        & \textbf{0.73}                \\ \bottomrule
+\end{tabular}
+\end{table}
+
+The performance of the models on the classification datasets is shown in Table~\ref{tab:classification_results}.
+
+\end{comment}
+
 \end{document}

### 2024-12-13 reports/presentation/pres.tex
new version

\documentclass{beamer}

\usepackage[backend=biber,style=ieee]{biblatex} % Load biblatex package
\addbibresource{../paper/bibliography.bib}
%\usepackage{natbib}

% Set the theme to a simple, clean style
\usetheme{default}
\usecolortheme{default}
\setbeamertemplate{itemize item}{\normalsize\textbullet} % Change bullets to dots
\setbeamertemplate{footline}[frame number]

\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{pythontex}
\usepackage{float}
\usepackage{booktabs}

\title{Variational Oblique Predictive Clustering Trees}
\author{Viktor Andonovikj}
\institute{Jožef Stefan Institute \& Jožef Stefan International Postgraduate School}
\date{\today}

\begin{document}

\setbeamertemplate{section in toc}[ball unnumbered]

% Title slide
\begin{frame}[plain]
  \titlepage
\end{frame}

% Outline slide
\begin{frame}[plain]
  \frametitle{Outline}

  {\small % Reduce font size
    \setlength{\baselineskip}{10pt} % Adjust line spacing
    \tableofcontents
  }

\end{frame}

\addtocounter{framenumber}{-2}

\section{Key Idea}
\begin{frame}
  \frametitle{Key Idea}
  \begin{itemize}
    \item We focus on \textbf{decision tree} type of models - popular due to their interpretability and simplicity.
    \item We aim to combine the \textbf{predictive power} of \textbf{ensemble} methods with the \textbf{interpretability} of a \textbf{single} decision tree.
    \item \textbf{Variational SPYCT} incorporates Bayesian inference to enhance both predictive performance and decision-making transparency.
  \end{itemize}
\end{frame}

\section{Introduction to Structured Output Prediction}
\begin{frame}
  \frametitle{Introduction to Structured Output Prediction (SOP)}
  \begin{itemize}
    \item \textbf{Structured Output Prediction (SOP)} involves predicting multiple interdependent outputs.
    \item SOP tasks include simultaneous prediction of:
      \begin{itemize}
        \item Multiple continuous values.
        \item Multiple discrete values.
        \item Hierarchically organized discrete values.
      \end{itemize}
    \item \textbf{Real-world applications}:
      \begin{itemize}
        \item Drug discovery: predicting multiple biological properties of molecules.
        \item Recommender systems: predicting user preferences across multiple items.
        \item Genomics: predicting gene functions based on interdependent traits.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Predictive Clustering Framework}
  \begin{itemize}
    \item \textbf{Predictive Clustering}: Combines clustering and prediction by treating prediction as a hierarchical clustering task where similar instances are grouped and predictions are made for each cluster.
    \item \textbf{Characteristics}:
      \begin{itemize}
        \item Supports both supervised and semi-supervised learning.
        \item State-of-the-art performance via ensemble learning.
        \item Offers interpretable models through feature importance analysis.
        \item Provides a unified framework for predictive modeling across multiple tasks.
      \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Predictive Clustering Trees (PCT) vs Oblique Predictive Clustering Trees (SPYCT)}

  \begin{figure}
    \centering
    \includegraphics[width=0.75\textwidth]{images/pct_spyct.jpg}
    {\tiny \caption{A) Learned split - SPYCT in blue, PCT in red B) PCT C) SPYCT~\cite{Stepi_nik_2021}}}
  \end{figure}

  \vspace{0.05cm} % Reduces space between the image and text

  \begin{itemize}
    \item \textbf{PCT}: Uses axis-aligned splits (based on single features), fully interpretable models.
    \item \textbf{SPYCT}: Uses oblique splits (linear combinations of features), more flexibility for high-dimensional and sparse data, suited for more complex tasks with intricate decision boundaries.
  \end{itemize}

\end{frame}

\begin{frame}
\frametitle{Introduction to Variational SPYCT}
  \begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{../figures/main_flowchart.pdf}
    {\tiny \caption{Ensemble of SPYCTs vs Variational SPYCT}}
  \end{figure}
\end{frame}

\section{Variational SPYCT}
\begin{frame}
  \frametitle{Introduction to Variational SPYCT}
  \begin{itemize}
    \item \textbf{Motivation}: Variational SPYCT (VSPYCT) integrates variational Bayes for improved decision-making within a single model, eliminating the need for ensembles.
    \item \textbf{Uncertainty quantification}: Embeds Bayesian inference directly into the decision tree structure, providing insight into decision processes and confidence levels.
    \item \textbf{Novelty}: VSPYCT introduces probabilistic treatment of oblique splits, offering a paradigm shift toward interpretable, and reliable machine learning models.
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Optimization through Variational Bayes}

  \small % Reduce text size to fit within margins

  \begin{itemize}
    \item \textbf{Bayes' Theorem}: 
      \[
      p(\theta|x) = \frac{p(x|\theta) p(\theta)}{p(x)}
      \]
    \item The primary computational challenge lies in calculating the posterior $p(\theta|x)$, where the denominator $p(x)$ requires:
      \[
      p(x) = \int_{\theta} p(x|\theta) p(\theta) \,d\theta
      \]
    \item \textbf{Variational Bayes (VB)} approximates the posterior with $q_{\omega^*}(\theta)$, minimizing the KL divergence:
      \[
      \omega^* = \arg\min_{\omega \in \Omega} KL(q_\omega(\theta) \parallel p(\theta|x))
      \]
    \item The Evidence Lower Bound (ELBO) is maximized to improve the approximation:
      \[
      KL(q_\omega(\theta) \parallel p(\theta|x)) = - \mathbb{E}_q[\log p(x,\theta) - \log q_\omega(\theta)] + \log p(x)
      \]
    \item VB's computational efficiency surpasses methods like MCMC, though it introduces bias based on the choice of the variational family $\mathcal{Q}$.
  \end{itemize}

\end{frame}

\begin{frame}
\frametitle{Optimization through Variational Bayes}
  \begin{figure}
    \centering
    \includegraphics[width=1.0\textwidth]{images/vb_optim.jpg}
    {\tiny \caption{Optimisation process of finding the closest variational distribution $q_\omega(\theta)$ over the set of latent variables $\omega$.~\cite{}}}
    \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Methodology Overview}
  \begin{itemize}
    \item VSPYCT follows SPYCT's tree-based architecture, but replaces fixed parameters with random variables.
    \item \textbf{Key difference}: VSPYCT uses variational Bayes (VB) for probabilistic split optimization, improving handling of noisy data and uncertainty.
    \item Split parameters (weights $\mathbf{w}$ and bias $b$) are modeled as random variables, allowing uncertainty estimation.
  \end{itemize}
  \[
  f(\mathbf{x}) = \sigma \left(\mathbf{w}^\top \mathbf{x} + b \right)
  \]
\end{frame}

\begin{frame}
  \frametitle{Learning Splits through Variational Bayes}
  \begin{itemize}
    \item Weights $\mathbf{w}$ and bias $b$ have Gaussian priors:
    \[
    \mathbf{w} \sim \mathcal{N}(\mathbf{0}, \mathbf{I}), \quad b \sim \mathcal{N}(0, 1)
    \]
    \item Variational Bayes approximates the posterior distribution with:
    \[
    q(\mathbf{w}, b|\mathcal{D}) = \mathcal{N}(\mathbf{w}|\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w) \mathcal{N}(b|\mu_b, \sigma_b^2)
    \]
    \item ELBO maximization: 
    \[
    \mathcal{L}(\boldsymbol{\mu}, \boldsymbol{\Sigma}) = \mathbb{E}_{q(\mathbf{w}, b|\mathcal{D})}[\log p(\mathcal{D}|\mathbf{w}, b)] - KL(q(\mathbf{w}, b|\mathcal{D}) \parallel p(\mathbf{w}, b))
    \]
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Algorithm for Learning a Split}

  \centering
  \begin{algorithm}[H]
  \caption{Variational Learning of Split Parameters}
  \label{alg:learn_split_vb}
  \begin{algorithmic}[1]
      \State \textbf{Input:} $\mathcal{D} = \{\mathbf{X}, \mathbf{Y}\}$, $\theta = \{\mathbf{w}_0, b_0\}$, $E$ (epochs), $\lambda$ (learning rate), $\beta$ (batch size), $\sigma$ (selection probability)
      \State \textbf{Output:} $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ (variational parameters)

      \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
          \State Initialize $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$
          \For{$e \in \{1, \dots, E\}$}
              \For{each mini-batch $\mathcal{B} \subseteq \mathcal{D}$ of size $\beta$}
                  \State Sample $\mathbf{w} \sim \mathcal{N}(\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w)$, $b \sim \mathcal{N}(\mu_b, \sigma_b^2)$
                  \State Compute impurity $\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)$
                  \State Compute ELBO $\mathcal{L}(\mathcal{B}; \Theta)$
                  \State Update $\Theta$
              \EndFor
          \EndFor
          \State \textbf{return} $\Theta$
      \EndProcedure
  \end{algorithmic}
  \end{algorithm}
\end{frame}

\begin{frame}
  \frametitle{Deriving the Impurity Function}

  \begin{itemize}
    \item For a given split node \(i\), we define the fuzzy membership:
    \[
    FM_i = \sigma(\mathbf{x}^\top \mathbf{w_i} + b_i), \quad 1 - FM_i = \text{right group membership}
    \]

    \item The split fitness function is the following:
    \[
    f(\mathbf{w_i}, b_i) = Z \cdot imp(FM) + (L + U - Z) \cdot imp(1 - FM)
    \]
    \item Impurity is given by the variances of the target \(Y\) and features \(X\):
    \[
    imp(FM) = \sum_{k=1}^{N} \sigma_{\bar{X}_k}^2 + \sum_{k=1}^{T} \sigma_{\bar{Y}_k}^2
    \]
  \end{itemize}

  \begin{itemize}
    \item In VSPYCT, we minimize this impurity by observing a target impurity of $impurity/2$ at each step using Variational Bayes.
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Making a Prediction}
  \begin{itemize}
    \item Prediction involves traversing the tree, making probabilistic decisions at each node.
    \item The split is determined by evaluating the function:
 \[
  f(\mathbf{x}) = \sigma \left(\mathbf{w}^\top \mathbf{x} + b \right)
  \]
    \item The final prediction $\hat{y}$ is the prototype value at the reached leaf:
    \[
    \hat{y} = \frac{1}{|\mathbf{Y}|} \sum_{i=1}^{|\mathbf{Y}|} y_i
    \]
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Prediction Process in VSPYCT}

  \footnotesize
  \setlength{\intextsep}{1pt}
  \setlength{\textfloatsep}{1pt}

  \begin{algorithm}[H]
    \caption{Prediction using Monte Carlo Sampling}
    \label{alg:make_pred}
    \begin{algorithmic}[1]
      \State \textbf{Input:} Feature vector $\mathbf{x}$, tree $T$, samples $M$
      \State \textbf{Output:} Prediction $\hat{y}$

      \State Initialize $\hat{y}_{\text{sum}} = 0$
      \For{$m = 1$ to $M$}
        \State Traverse $T$ from root to leaf with sampled $\mathbf{w}^{(m)}, b^{(m)}$
        \State $\hat{y}_{\text{sum}} \gets \hat{y}_{\text{sum}} + $ prediction at leaf
      \EndFor
      \State \Return $\hat{y} = \frac{\hat{y}_{\text{sum}}}{M}$
    \end{algorithmic}
  \end{algorithm}

\end{frame}

\begin{frame}
  \frametitle{Feature Importance}
  \begin{itemize}
    \item Feature importance in VSPYCT: Calculated as the influence of features across all splits.
    \item The importance of a feature is determined by:
    \[
    \text{Imp}(T) = \sum_{s \in T} \left( \frac{s_n}{N} \right) \left( \frac{\mathbb{E}[\mathbf{w}_s]}{|\mathbb{E}[\mathbf{w}_s]|_1} \right)
    \]
    \item Weights $\mathbf{w}$ are sampled from the posterior distribution, and importance is aggregated over all splits.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Time Complexity Analysis}
  \begin{itemize}
    \item Time complexity of learning a split in VSPYCT:
    \[
    \mathcal{O}(MNI_{vb}(D+K))
    \]
    \item $M$: Number of Monte Carlo samples, $N$: Data points, $D$: Features, $K$: Clustering attributes, $I_{VB}$: Number of optimization iterations.
  \end{itemize}
\end{frame}

\section{Experimental Setting}
\begin{frame}
  \frametitle{Experimental Setting}
  \begin{itemize}
    \item \textbf{Comparison}: VSPYCT vs SPYCT (single tree and ensemble).
    \item \textbf{Tasks}: Single-target and multi-target regression, binary and multi-class classification.
    \item \textbf{Goal}: Evaluate predictive performance of VSPYCT across different modeling scenarios.
  \end{itemize}
\end{frame}

\section{Data}
\begin{frame}
  \frametitle{Data}
  \begin{itemize}
    \item Datasets cover regression and classification tasks.
    \item Number of examples (N), features (D), targets (T), and classes (C).
  \end{itemize}

  \begin{table}[h!]
    \centering
    \caption{Summary of the datasets used in the experiments.}
    \label{tab:datasets}
    \scriptsize % Reduce font size to fit the table
    \begin{tabular}{lcccccc}
    \toprule
    \textbf{Dataset} & \textbf{Type of Task} & \textbf{N} & \textbf{D} & \textbf{T} & \textbf{C} \\
    \midrule
    rf1~\cite{mulan}              & Multi-target regression  & 9125   & 64   & 8  & -- \\
    rf2~\cite{mulan}              & Multi-target regression  & 9125   & 576  & 8  & -- \\
    atp1d~\cite{mulan}            & Multi-target regression  & 337    & 411  & 6  & -- \\
    atp7d~\cite{mulan}            & Multi-target regression  & 296    & 411  & 6  & -- \\
    scm1d~\cite{mulan}            & Multi-target regression  & 9803   & 280  & 16 & -- \\
    house\_8L~\cite{openml}        & Single-target regression & 22784  & 8    & 1  & -- \\
    puma8NH~\cite{openml}          & Single-target regression & 8192   & 8    & 1  & -- \\
    diabetes~\cite{openml}         & Binary classification    & 768    & 8    & 1  & 2 \\
    banknote~\cite{openml}         & Binary classification    & 1372   & 4    & 1  & 2 \\
    gas-drift~\cite{openml}        & Multi-class classification & 13910 & 128  & 1  & 6 \\
    balance~\cite{openml}          & Multi-class classification & 625   & 4    & 1  & 3 \\
    \bottomrule
    \end{tabular}
  \end{table}

\end{frame}

\begin{frame}
  \frametitle{Evaluation Metrics}
  \begin{itemize}
    \item \textbf{Regression}: Mean Absolute Error (MAE).
    \[
    \textit{MAE} = \frac{1}{N} \sum_{i=1}^{N} \left| y_i - \hat{y}_i \right|
    \]
    \item \textbf{Classification}: F1 score (macro-averaged for multi-class tasks).
    \item \textbf{Cross-validation}: 5-fold cross-validation used to reduce bias.
  \end{itemize}
\end{frame}

\section{Results}
\begin{frame}
  \frametitle{Regression Results}

  \begin{table}[h!]
    \centering
    \caption{\textit{MAE} Scores for Regression Datasets}
    \label{tab:regression_results}
    \scriptsize
    \begin{tabular}{lccc}
    \toprule
    \textbf{Dataset} & \textbf{SPYCT-single tree} & \textbf{SPYCT-ensemble} & \textbf{VSPYCT} \\
    \midrule
    rf1            & 29.38         & \textbf{29.34}  & 29.37 \\
    rf2             & 29.37         & 29.49           & \textbf{29.36} \\
    atp1d           & 95.17         & \textbf{70.93}  & 77.24 \\
    atp7d           & 115.54        & \textbf{73.08}  & 91.28 \\
    scm1d           & 205.99        & 206.01          & \textbf{205.72} \\
    house\_8L       & 25378.02      & \textbf{19817.56} & 22992.24 \\
    puma8NH         & 2.80          & \textbf{2.62}   & 2.74 \\
    \bottomrule
    \end{tabular}
  \end{table}

  \vspace{0.2cm}
  \begin{itemize}
    \item VSPYCT outperforms or closely matches the SPYCT ensemble on several datasets (rf2, scm1d).
    \item Ensemble methods generally achieve lower MAE, but VSPYCT provides competitive results.
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Classification Results}

  \begin{table}[h!]
    \centering
    \caption{F1 Scores for Classification Datasets}
    \label{tab:classification_results}
    \scriptsize
    \begin{tabular}{lccc}
    \toprule
    \textbf{Dataset} & \textbf{SPYCT-single tree} & \textbf{SPYCT-ensemble} & \textbf{VSPYCT} \\
    \midrule
    diabetes              & 0.53                            & \textbf{0.60}            & 0.59 \\
    banknote              & 0.97                            & \textbf{0.99}            & 0.98 \\
    gas-drift             & 0.98                            & 0.99            & \textbf{0.99} \\
    balance               & 0.60                            & 0.66                     & \textbf{0.73} \\
    \bottomrule
    \end{tabular}
  \end{table}

  \vspace{0.2cm}
  \begin{itemize}
    \item VSPYCT achieves competitive performance across datasets, often matching or surpassing ensembles.
    \item In the balance dataset, VSPYCT achieves the best F1 score.
  \end{itemize}

\end{frame}

\section{Interpretability}
\begin{frame}
  \frametitle{Interpretability - An example with real dataset}

  \begin{itemize}
    \item \textbf{Dataset:} Unemployment dataset from the Slovenian Public Employment Service (PES), consisting of 74,086 anonymized instances.
    \item \textbf{Features:} Age, gender, education, work experience, and other personal/professional characteristics.
    \item \textbf{Target:} Time until the jobseeker becomes employed or exits the study (measured in days). Some records are right-censored (event not observed).
    \item \textbf{Goal:} Predict the time-to-event (employment) with censored data, using a semi-supervised learning approach for handling the inherent missing information.
    \item \textbf{Challenges:} Diverse attribute types (categorical, numerical, temporal) and the presence of censored data.
  \end{itemize}

\end{frame}

\begin{frame}
  \frametitle{Structure of the learned VSPYCT model}
  \begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../figures/hecat_tree_old_temp.pdf}
    \caption{VSPYCT tree structure. Colors represent different feature groups (e.g., education, profession).}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Predicted Survival Curve}
  \begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../figures//pred_with_ci.pdf}
    \caption{Predicted survival curve for a jobseeker. Blue: mean prediction; shaded area: confidence interval; vertical line: true event time.}
  \end{figure}
  \begin{itemize}
    \item The confidence interval captures the uncertainty behind the predictions.
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Feature Importance}
  \begin{figure}
    \centering
    \includegraphics[width=0.8\textwidth]{../figures/feature_importance.pdf}
    \caption{Feature importance from the VSPYCT model. "Months of work experience" is the most important factor in the model's decisions.}
  \end{figure}
\end{frame}

\section{Conclusion}
\begin{frame}
  \frametitle{Conclusion}

  \begin{itemize}
    \item \textbf{VSPYCT Model:} Combines oblique splits, variational Bayes, and Bayesian inference in an interpretable predictive framework.
    \item \textbf{Key Characteristic:} Balances the performance of ensemble models with the interpretability of single decision trees, while introducing uncertainty quantification.
    \item \textbf{Results:} Competitive performance, sometimes surpassing ensembles of SPYCT.
    \item \textbf{Uncertainty Quantification:} Enhances decision-making reliability, making the model applicable to domains where certainty in predictions is critical.
    \item \textbf{Future Work:} Further performance improvements and extending applications to diverse predictive tasks.
  \end{itemize}

\end{frame}

\addtocounter{framenumber}{-2}

% Bibliography slide
\begin{frame}[allowframebreaks, plain] % Allows the bibliography to continue on multiple slides if necessary
  \frametitle{References}
  \printbibliography
\end{frame}

% Thank You Slide
\begin{frame}[plain]
  \frametitle{Thank You}
  \centering
  Questions?
\end{frame}

\end{document}

### 2024-12-13 src/data/dataset.py

@@ -256,7 +256,7 @@ class ReducedDataset:
 
     def __init__(self, path, filename, random_state=11):
         self.dataset = pd.read_pickle(os.path.join(path, filename))
-        self.dataset = self.dataset.sample(frac=0.05)
+        self.dataset = self.dataset.sample(frac=0.005)
         self.random_state = random_state
         self.skp = pd.read_csv(os.path.join(DIR_PATH, '../../data/raw/dimSKP08.csv'))[['IDpoklicaSKP','SFpoklicaSKP']]
         self.skp.SFpoklicaSKP = self.skp.SFpoklicaSKP.astype(int).astype(str).str.zfill(4)

### 2024-12-13 src/models/model.py

@@ -4,19 +4,36 @@ from src.models.split import learn_split, learn_split_vb
 import numpy as np
 import pyro
 from pyro.infer import Predictive
+from sklearn.preprocessing import MinMaxScaler
 
-def nanvar(y, dim=0):
-    mean = torch.nanmean(y, dim)
-    squared_diff = (y - mean.unsqueeze(dim)) ** 2
-    valid_count = torch.sum(~torch.isnan(y), dim)
-    variance = torch.nansum(squared_diff, dim) / (valid_count - 1) 
-    return variance
+def nanvar(tensor, dim=None, keepdim=False):
+    tensor_mean = tensor.nanmean(dim=dim, keepdim=True)
+    output = (tensor - tensor_mean).square().nanmean(dim=dim, keepdim=keepdim)
+    return output
 
-def impurity(values): return torch.sum(nanvar(values, dim=0))
+# Adjusted function to handle multiple target variables
+def multi_target_impurity(tensor, target_dim=1):
+    """
+    Compute impurity for multiple target variables.
+    
+    tensor: Input tensor where each row represents a sample and each column represents a different target variable.
+    target_dim: The dimension along which the targets are stored (e.g., 1 if columns are different targets).
+    """
+    # Remove NaN values and check for empty tensor
+    if tensor.isnan().all():
+        return float('inf')
+    
+    # Compute variances for each target variable (along target_dim)
+    variances = nanvar(tensor, dim=0 if target_dim == 1 else 1)
+    
+    # Return the sum of variances across all target variables
+    total_variance = torch.sum(variances)
+
+    return total_variance
 
 class Spyct:
     def __init__(self, max_depth=np.inf, subspace_size=1, minimum_examples_to_split=2,
-                 device='cpu', epochs=200, bs=None, lr=0.01):
+                 device='cpu', epochs=100, bs=None, lr=0.1):
         self.minimum_examples_to_split = minimum_examples_to_split
         self.root_node = None
         self.num_nodes = 0
@@ -32,7 +49,7 @@ class Spyct:
         if rows is None: rows = torch.arange(descriptive_data.shape[0])
 
         self.num_training_instances = descriptive_data.shape[0]
-        total_variance = impurity(clustering_data)
+        total_variance = multi_target_impurity(clustering_data)
         self.root_node = Node(depth=0, enable_mc_dropout=enable_mc_dropout)
         splitting_queue = [(self.root_node, rows, total_variance)]
[…]

### 2024-12-13 src/models/node.py

@@ -41,7 +41,7 @@ class Node:
 
 
 class VNode:
-    def __init__(self, depth=0, num_instances=0, root=False):
+    def __init__(self, descriptive_values, depth=0, num_instances=0, root=False):
         self.left = None
         self.right = None
         self.prototype = None   
@@ -51,7 +51,9 @@ class VNode:
         self.depth = depth
         self.is_root = root
         self.param_store = None
+        self.losses = None
         self.num_instances = num_instances
+        self.descriptive_values = descriptive_values
 
     def is_leaf(self): return self.left is None and self.right is None
 
@@ -68,7 +70,7 @@ class VNode:
                 if splits <= 0.5: return self.left.predict(x)
                 else: return self.right.predict(x)
 
-    def mc_predict(self, x, num_samples=10):
+    def mc_predict(self, x, num_samples=30):
         pred_lst = []
         for i in range(num_samples):
             pred_lst.append(self.predict(x))

### 2024-12-13 src/models/split.py

@@ -34,14 +34,21 @@ DIR_PATH = os.path.abspath(os.path.dirname(__file__))
 
 # def weighted_variance(values, weights, weight_sum):
 #     values = torch.nan_to_num(values, nan=0.0)
-#     weighted_squares = torch.matmul(weights, values) / weight_sum
-#     return -torch.sum(weighted_squares*weighted_squares)
+#     mean = torch.matmul(weights, values) / weight_sum
+#     return -torch.sum(mean*mean)
+
+# def weighted_variance(values, weights, weight_sum):
+#     values = torch.nan_to_num(values, nan=0.0)
+#     weights = weights.view(weights.shape[0], 1)
+#     weighted_squares = weights*values
+#     return torch.sum((weighted_squares-weighted_squares.mean(axis=0))**2)/ weight_sum
 
 def weighted_variance(values, weights, weight_sum):
-    values = torch.nan_to_num(values, nan=0.0)
-    weights = weights.view(weights.shape[0], 1)
-    weighted_squares = weights*values
-    return torch.sum((weighted_squares-weighted_squares.mean(axis=0))**2)/ weight_sum
+    mean = torch.matmul(weights, values) / weight_sum
+    squared_diff = (values - mean) ** 2
+    weighted_var = torch.sum(torch.matmul(weights, squared_diff)) / weight_sum
+    return weighted_var
+
 
 class MC_Dropout_Linear(nn.Module):
     def __init__(self, input_dim, output_dim, dropout_prob=0.2):
@@ -73,10 +80,9 @@ class Impurity(PyroModule):
         self.linear.weight = PyroSample(dist.Normal(0, 1).expand([self.output_dim, self.input_dim]).to_event(2))
         self.linear.bias = PyroSample(dist.Normal(0, 1).expand([self.output_dim]).to_event(1))
 
-    #fix this with the guide
     def forward(self, descriptive_data, clustering_data):
-        # sigma = pyro.sample("sigma", dist.Normal(1., 10.))
-        right_selection = self.linear(descriptive_data).reshape(-1).sigmoid()
+        # sigma = pyro.sample("sigma", dist.Normal(0., 1.))
+        right_selection = self.linear(descriptive_data).reshape(-1).sigmoid()   
         left_selection = torch.tensor(1., device=self.device) - right_selection
 
         right_weight_sum = torch.sum(right_selection)
@@ -85,19 +91,17 @@ class Impurity(PyroModule):
         var_left = weighted_variance(clustering_data, left_selection, left_weight_sum)
[…]

### 2024-12-13 reports/paper/vspyct.tex
grammar fixes

@@
 The VB approach offers an approximation to this issue by closely mimicking the true posterior $p(\theta|x)$ with an approximate distribution $q_{\omega^*}(\theta)$.
-This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterized by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
+This variational distribution is part of a family of distributions $q_\omega(\theta)\in \mathcal{Q}$, parameterised by $\omega \in \Omega$, where $\Omega$ represents the potential latent parameter values.
 Typically, the mean-field variational family is utilised, assuming independence among the parameters $\omega$~\citep{Blei_2017}. 
@@
     \Procedure{LearnSplit}{$\mathcal{D}, \theta, E, \lambda, \beta, \sigma$}
-        \State Initialize $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialize variational parameters}
-        \State Initialize optimizer: $\alpha \gets \lambda$
+        \State Initialise $\Theta = \{\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w, \mu_b, \sigma_b^2\}$ \Comment{Initialise variational parameters}
+        \State Initialise optimiser: $\alpha \gets \lambda$
         \For{$e \in \{1, \dots, E\}$} \Comment{Iterate over epochs}
@@
 \Function{MC\_Predict}{$\Omega$, $\mathbf{x}$, $M$}
-    \State Initialize $\hat{y}_{\text{sum}} = 0$
+    \State Initialise $\hat{y}_{\text{sum}} = 0$
     \For{$m = 1$ to $M$}
@@
 
-However, the overall feature importance is derived by aggregating these normalized ratios across all splits within the tree.
+However, the overall feature importance is derived by aggregating these normalised ratios across all splits within the tree.
 Even if \textit{salary} appears strongly influential at the root node, the final importance calculation considers both the magnitude and the stability of the weights across all nodes.
-A feature consistently demonstrating a favorable \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio throughout multiple splits will be deemed more important globally.
+A feature consistently demonstrating a favourable \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio throughout multiple splits will be deemed more important globally.
 In this way, the final importance assessment accounts for both the strength and certainty of each feature’s contribution to the model’s decision-making process.
@@
 
-For multi-target regression, a normalized Mean Absolute Error (\textit{NMAE}) was used to ensure comparability across targets with different value ranges.
+For multi-target regression, a normalised Mean Absolute Error (\textit{NMAE}) was used to ensure comparability across targets with different value ranges.
 Letting \(y_{ij}\) and \(\hat{y}_{ij}\) be the true and predicted values for the \(j\)-th target of the \(i\)-th instance, and \(T\) be the total number of target variables, the \(\textit{NMAE}\) is defined as:

### 2024-12-16 reports/paper/vspyct.tex
added separate tables with datasets

@@
 
-The experimental evaluation was conducted to compare the \gls{vspyct} model with several benchmarks, including the \gls{spyct} model, \glspl{opct}, and ensemble of \glspl{pct}.
+The experimental evaluation was conducted to compare the \gls{vspyct} model with several benchmarks, including the \gls{spyct} model, \glspl{opct}~\citep{Stepisnik_2020}, and ensemble of \glspl{pct}.
 Specifically, we considered both a single \gls{spyct} and an ensemble of \glspl{spyct}, as well as a single \gls{opct} and an ensemble of \glspl{pct}.
-We chose these methods as primary comparisons because they have been extensively evaluated in the literature, demonstrating strong performance across various \gls{sop} tasks~\citep{Andonovikj_2024}.
+We chose these methods as primary comparisons because they have been extensively evaluated in the literature, demonstrating strong performance across various \gls{sop} tasks~\citep{Andonovikj_2024, Stepi_nik_2021}.
 
@@
 
-The experiments were conducted on a diverse set of datasets, encompassing different types of predictive modelling tasks, including single-target regression, multi-target regression, and classification (binary and multi-class).
-The general characteristics of the datasets are summarised in Table~\ref{tab:datasets}.
-The table includes the number of examples \(N\), the number of features \(D\), the number of targets \(T\), and the number of classes \(C\), with \(C\) being applicable only to classification tasks.
+The experimental evaluation was performed on a diverse collection of datasets, each reflecting a distinct predictive modelling task: single-target regression (STR), multi-target regression (MTR), binary classification (BC), and multi-class classification (MCC).
+The properties of these benchmark datasets are detailed in Tables~\ref{tab:str} through \ref{tab:mcc}, respectively.
+Each table lists the number of examples (\(N\)) and the number of features (\(D\)) for all datasets, along with the number of targets (\(T\)) for MTR datasets and the number of classes (\(C\)) for MCC datasets.
 
@@
 \centering
-\caption{Summary of the datasets used in the experiments.}
-\label{tab:datasets}
-\begin{tabular}{@{}lcccccc@{}}
+\caption{Properties of the benchmark STR datasets}
+\label{tab:str}
+\begin{tabular}{@{}lcccc@{}}
 \toprule
-\textbf{Dataset} & \textbf{Type of Task}          & \textbf{N} & \textbf{D} & \textbf{T} & \textbf{C} \\ \midrule
-rf1~\citep{mulan}             & Multi-target regression        & 9125         & 64         & 8         &        \\
-rf2~\citep{mulan}             & Multi-target regression        & 9125         & 576         & 8         &        \\
-atp1d~\citep{mulan}           & Multi-target regression        & 337         & 411         & 6         &        \\
-atp7d~\citep{mulan}           & Multi-target regression        & 296         & 411         & 6         &        \\
-scm1d~\citep{mulan}           & Multi-target regression        & 9803         & 280         & 16         &        \\
-house\_8L~\citep{openml}       & Single-target regression       &  22784        & 8         & 1          &        \\
-puma8NH~\citep{openml}         & Single-target regression       & 8192         & 8         & 1          &        \\
-diabetes~\citep{openml}        & Binary classification          & 768         & 8         & 1          & 2          \\
-banknote~\citep{openml} & Binary classification          & 1372         & 4         & 1          & 2          \\
-gas-drift~\citep{openml}  & Multi-class classification     & 13910         & 128         & 1          & 6         \\
-balance~\citep{openml}         & Multi-class classification     & 625         & 4         & 1          & 3         \\ \bottomrule
+\textbf{Dataset}   & \textbf{N} & \textbf{D} \\ \midrule
+era~\citep{openml}                  & 1000         & 5          \\
+cpmp~\citep{openml}                     & 2108         & 27          \\
+qsar234~\citep{openml}                   & 2145         & 1026             \\
+concrete\_compressive\_strength~\citep{openml}                   & 1030         & 9          \\
+yprop~\citep{openml}                   & 8885         & 252       \\
+puma8NH~\citep{openml}                & 8192         & 8        \\
+spacega~\citep{openml}                  & 3107         & 6             \\
+bike\_sharing~\citep{openml}           & 17379         & 6            \\
+quake~\citep{openml}       & 2178         & 3         \\
+ailerons~\citep{openml}         & 13750         & 40           \\ \bottomrule
 \end{tabular}
@@
 
-For multi-target regression, the target variable is a vector, while for single-target regression, it is a scalar.
-In binary classification, there are two possible classes, and for multi-class classification, the number of classes is dataset-dependent.
+\begin{table}[h!]
+\centering
+\caption{Properties of the benchmark MTR datasets}
+\label{tab:mtr}
+\begin{tabular}{@{}lccccc@{}}
+\toprule
+\textbf{Dataset}   & \textbf{N} & \textbf{D} & \textbf{T}     \\ \midrule
+wq~\citep{openml}                  & 1060         & 16         & 14 \\
+atp1d~\citep{openml}               &       337         & 411   &  6     \\
+atp7d~\citep{openml}                   & 296         & 411       &  6    \\
+scpf~\citep{openml}                   & 1137         & 23       &  3  \\
+osales~\citep{openml}                   & 639         & 411   &  12  \\
+sf1~\citep{openml}             &  323        & 10       &  3   \\
+sf2~\citep{openml}                & 1066         & 10     & 3  \\
+rf1~\citep{openml}                  & 9125         & 64       &  8    \\
+rf2~\citep{openml}           &  9125        & 576       &  8   \\
+slump~\citep{openml}       & 103         & 7   &  3    \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[h!]
+\centering
+\caption{Properties of the benchmark BC datasets}
+\label{tab:str}
+\begin{tabular}{@{}lcccc@{}}
+\toprule
+\textbf{Dataset}   & \textbf{N} & \textbf{D} \\ \midrule
+scene~\citep{mulan}                  & 2407         & 299          \\
+ova\_breast~\citep{openml}                     & 1545         & 10935          \\
+ova\_lung~\citep{openml}                   & 1545         & 10935             \\
+bio\_response~\citep{openml}                   & 3751         & 1777          \\
+chronic\_kidney\_disease~\citep{openml}                   & 400         & 25       \\
+compas\_two\_years~\citep{openml}             &  5278        & 13          \\
+spambase~\citep{openml}                & 4601         & 57        \\
+gina\_agnostic~\citep{openml}                  & 3468         & 970             \\
+credit-g~\citep{openml}           & 1000         & 20            \\
+diabetes\citep{openml}       & 768         & 8         \\ \bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}[h!]
+\centering
+\caption{Properties of the benchmark MCC datasets}
+\label{tab:mcc}
+\begin{tabular}{@{}lccccc@{}}
+\toprule
+\textbf{Dataset}   & \textbf{N} & \textbf{D} & \textbf{C}     \\ \midrule
+amazon\_reviews~\citep{openml}                  & 1500         & 10000   &  50     \\
+dermatology~\citep{openml}                     & 366         & 34     &  6   \\
+micro\_mass~\citep{openml}                   & 360         & 1300      &   10    \\
+balance~\citep{openml}                   & 625         & 4       & 3  \\
+yeast~\citep{mulan}                   & 1484         & 8    & 10  \\
+wine~\citep{openml}             &  178        & 13     &  3   \\
+mfeat\_zernike~\citep{openml}                & 2000         & 47   &  10   \\
+har~\citep{openml}                  & 10299         & 561        &  6   \\
+gas\_drift~\citep{openml}           & 13910         & 128      &   6   \\
+semeion~\citep{openml}       & 1593         & 257    &   10  \\ \bottomrule
+\end{tabular}
+\end{table}
+
+In STR tasks, the output is a single continuous variable, while in MTR tasks, the output comprises multiple continuous target variables.
+BC tasks involve two possible classes, and MCC tasks extend to multiple classes, with the exact number of classes depending on the specific dataset.
+This comprehensive range of datasets ensures a robust evaluation of the models across varying complexity and output structures.
 
@@
 Figure~\ref{fig:avg_rank} shows the average rankings of the evaluated methods—\gls{vspyct}, \gls{opct}, single \gls{spyct}, ensemble of \glspl{spyct}, and ensemble of \glspl{pct}—across four task types: binary classification (BC), multi-class classification (MCC), single-target regression (STR), and multi-target regression (MTR).
+For improved readability, the actual performance of the models on each dataset is given in the Appendix.
 

### 2024-12-17 reports/paper/vspyct.tex
appendix

@@
 
-\begin{comment}
 \section*{Appendix}
-
-The evaluation scores on the regression datasets are presented in Table~\ref{tab:regression_results}.
 
@@
 \centering
-\caption{\textit{MAE} Scores for Regression Datasets}
-\label{tab:regression_results}
-\begin{tabular}{@{}lccc@{}}
+\caption{\textit{MAE} scores for STR (lowest is best)}
+\label{tab:str_results}
+\begin{tabular}{@{}lccccc@{}}
 \toprule
-Dataset & SPYCT-single tree & SPYCT-ensemble & VSPYCT \\ \midrule
-rf1             & 29.38                            & \textbf{29.34}                        & 29.37                \\
-rf2             & 29.37                            & 29.49                        & \textbf{29.36}                \\
-atp1d           & 95.17                            & \textbf{70.93}                        &  77.24               \\
-atp7d           & 115.54                            & \textbf{73.08}                        & 91.28                \\
-scm1d           & 205.99                            & 206.01                        & \textbf{205.72}                \\
-house\_8L       & 25378.02                            & \textbf{19817.56}                        & 22992.24                \\
-puma8NH         & 2.80                            & \textbf{2.62}                        & 2.74                \\ \bottomrule
+Dataset & SPYCT-single & SPYCT-ensemble & VSPYCT & PCT-Ensemble & OPCT \\ \midrule
+era                             & 1.296625 & \textbf{1.249352} & 1.287840 & 1.277622 & 1.277424 \\
+cpmp2015                        & 384.377425 & 326.497515 & \textbf{285.105799} & 1211.124409 & 848.721320 \\
+qsar234                         & 0.572577 & \textbf{0.415656} & 0.428724 & 0.428721 & 0.441020 \\
+puma8NH                         & 2.795463 & \textbf{2.632242} & 2.711678 & - & - \\
+yprop                           & 0.020126 & 0.019693 & 0.019969 & \textbf{0.019682} & 0.023687 \\
+space\_ga                       & 0.105716 & 0.090193 & 0.104501 & \textbf{0.084793} & 0.099387 \\
+bike\_sharing                   & 7.866288 & 3.101784 & 11.288880 & 10.610927 & \textbf{2.509829} \\
+quake                           & 0.144549 & 0.143972 & \textbf{0.143805} & 0.147743 & 0.172350 \\
+concrete\_compressive\_strength & 4.500918 & \textbf{4.121670} & 7.740371 & 4.164038 & 4.216046 \\
+ailerons                        & 0.000135 & \textbf{0.000124} & 0.000127 & 0.000133 & 0.000161 \\
+\bottomrule
 \end{tabular}
@@
 \centering
-\caption{F1 Scores for Classification Datasets}
-\label{tab:classification_results}
-\begin{tabular}{@{}lccc@{}}
+\caption{\textit{NMAE} scores for MTR (lowest is best)}
+\label{tab:regression_results}
+\begin{tabular}{@{}lccccc@{}}
 \toprule
-Dataset & SPYCT-single tree & SPYCT-ensemble & VSPYCT \\ \midrule
-diabetes              & 0.53                            & \textbf{0.60}                        & 0.59                \\
-banknote       & 0.97                            & \textbf{0.99}                        & 0.98                \\
-gas-drift        & 0.98                            & 0.99                        & \textbf{0.99}                \\
-balance               & 0.60                            & 0.66                        & \textbf{0.73}                \\ \bottomrule
+Dataset & SPYCT-single & SPYCT-ensemble & VSPYCT & PCT-Ensemble & OPCT \\ \midrule
+wq       & 0.184654 & \textbf{0.167491} & 0.183236 & 0.169793 & 0.175325 \\
+atp7d    & 0.321386 & 0.157284 & 0.167780 & \textbf{0.147780} & 0.176141 \\
+atp1d    & 0.229711 & \textbf{0.104547} & 0.125300 & 0.106302 & 0.115277 \\
+scpf     & 0.031613 & 0.027816 & \textbf{0.027291} & 0.047811 & 0.052046 \\
+osales   & 0.068986 & 0.064457 & \textbf{0.057865} & 0.060054 & 0.058278 \\
+sf2      & 0.042276 & 0.038311 & \textbf{0.037199} & 0.039234 & 0.041858 \\
+sf1      & 0.097729 & 0.095937 & \textbf{0.092266} & 0.108493 & 0.109708 \\
+slump    & 0.182132 & \textbf{0.165759} & 0.239199 & 0.175965 & 0.243881 \\
+rf1      & 0.197493 & 0.183436 & \textbf{0.166669} & 0.167392 & 0.168493 \\
+rf2      & 0.212579 & \textbf{0.176247} & 0.193914 & 0.177062 & 0.194180 \\
+\bottomrule
 \end{tabular}
@@
 
-The performance of the models on the classification datasets is shown in Table~\ref{tab:classification_results}.
+\begin{table}[h!]
+\centering
+\caption{\textit{F1} scores for BC (highest is best)}
+\label{tab:classification_results}
+\begin{tabular}{@{}lccccc@{}}
+\toprule
+Dataset & SPYCT-single & SPYCT-ensemble & VSPYCT & PCT-Ensemble & OPCT \\ \midrule
+scene                  & 0.944059 & \textbf{0.986098} & 0.972439 & 0.882295 & 0.935982 \\
+ova\_breast            & 0.928256 & 0.946192 & 0.959834 & \textbf{0.966468} & 0.956409 \\
+bioresponse            & 0.729417 & 0.777438 & 0.768986 & \textbf{0.777509} & 0.753144 \\
+ova\_lung              & 0.802281 & 0.905439 & \textbf{0.920463} & 0.886492 & 0.904865 \\
+chronic\_kidney\_disease & 0.961777 & 0.991777 & \textbf{1.000000} & 0.997178 & 0.997178 \\
+compas\_two\_years     & 0.645271 & 0.641390 & 0.652540 & \textbf{0.666824} & 0.615251 \\
+spambase               & 0.903289 & \textbf{0.934069} & 0.926491 & 0.924987 & 0.912740 \\
+gina\_agnostic         & 0.856799 & 0.885768 & 0.858579 & 0.909199 & \textbf{0.922184} \\
+credit-g               & 0.632353 & 0.657097 & \textbf{0.681566} & 0.615338 & 0.547619 \\
+diabetes               & 0.698525 & 0.739710 & 0.715029 & \textbf{0.778026} & 0.694843 \\
+\bottomrule
+\end{tabular}
+\end{table}
 
-\end{comment}
+\begin{table}[h!]
+\centering
+\caption{\textit{F1} scores for MCC (highest is best)}
+\label{tab:mcc_results}
+\begin{tabular}{@{}lccccc@{}}
+\toprule
+Dataset & SPYCT-single & SPYCT-ensemble & VSPYCT & PCT-Ensemble & OPCT \\ \midrule
+amazon\_reviews & 0.143184 & 0.331889 & 0.172791 & \textbf{0.333139} & 0.293184 \\
+dermatology      & 0.852633 & 0.978408 & \textbf{0.980117} & 0.893882 & 0.922443 \\
+micro\_mass      & 0.845445 & 0.940100 & \textbf{0.954908} & 0.940397 & 0.905830 \\
+balance          & 0.645499 & 0.618803 & \textbf{0.825152} & 0.595340 & 0.585561 \\
+yeast            & 0.443345 & 0.535824 & 0.471005 & \textbf{0.575664} & 0.443048 \\
+wine             & 0.975580 & 0.965350 & 0.966350 & \textbf{1.000000} & 0.945825 \\
+mfeat\_zernike   & 0.786454 & \textbf{0.804786} & 0.792283 & 0.759459 & 0.693469 \\
+har              & 0.969661 & 0.982193 & \textbf{0.984686} & 0.942095 & 0.942095 \\
+gas\_drift        & 0.974804 & \textbf{0.993256} & 0.988948 & 0.988421 & 0.981730 \\
+semeion          & 0.846045 & \textbf{0.923838} & 0.868518 & 0.897004 & 0.811956 \\
+\bottomrule
+\end{tabular}
+\end{table}
 

### 2025-01-06 reports/paper/vspyct.tex
Comments

@@
 	\ead{biljana.mileva@ijs.si}
-	\author[1]{Pavle Boškoski}
+	\author[1,3]{Pavle Boškoski}
 	\ead{pavle.boskoski@ijs.si}
@@
 In the evolving field of machine learning, the fusion of semi-supervised learning and decision tree models has led to the development of models like \glspl{pct}~\citep{Kocev_2013} and their advanced variant, \glspl{spyct}~\citep{Stepi_nik_2021}.
-These models leverage both labeled and unlabelled data to enhance prediction accuracy and efficiency.
+These models leverage labeled and unlabelled data to enhance prediction accuracy and efficiency.
 \Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
 Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal performance.
-While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, crucial for informed decision-making across various applications.
+While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, which is crucial for informed decision-making across various applications.
 
@@
 %This approach not only elevates the explainability of semi-supervised decision trees by consolidating predictive capabilities into a single model but also marks a significant advancement in machine learning, particularly within semi-supervised learning domains.
-By combining the robustness of \gls{spyct} ensembles with the clarity and interpretability of a single tree model, alongside introducing uncertainty quantification, our research provides a comprehensive tool aimed to overcome current limitations of predictive clustering methodologies.
+Combining the robustness of \gls{spyct} ensembles with the clarity and interpretability of a single tree model, alongside introducing uncertainty quantification, our research provides a comprehensive tool to overcome the current limitations of predictive clustering methodologies.
 
@@
 This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.
-The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, modelling these parameters as random variables with probability distributions.
+The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, and modelling these parameters as random variables with probability distributions.
 This represents a paradigm shift toward a probabilistic understanding of decision-making processes within tree-based models, significantly enhancing interpretability, especially in uncertainty estimation domains.
 It sets a new benchmark for interpretable, efficient, and reliable machine learning models.
+\pbnote{Check Bayesian Additive Regression Trees (BART). Maybe add a comment/reference what is different}
 
@@
 
-\Glspl{pct} have been an essential development in extending decision tree capabilities to a variety of predictive modelling tasks, including structured output prediction.
+\Glspl{pct} have been an essential development in extending decision tree capabilities to various predictive modelling tasks, including structured output prediction.
 \Glspl{pct} are versatile and can be combined into ensembles to achieve state-of-the-art performance~\citep{Kocev_2013}.
@@
 
-\Glspl{pct} are celebrated for their interpretability and efficiency, yet they often fall short in predictive performance due to their myopic, greedy nature of selecting splits.
-The introduction of option predictive clustering trees (OPCTs)~\citep{Stepisnik_2020} aimed to address this limitation by incorporating multiple alternative splits within option nodes.
-This approach mitigates the inherent myopia and achieves competitive performance with ensemble methods like bagging and random forests, while still maintaining a degree of interpretability.
-OPCTs, despite mitigating myopia and enhancing performance, still face challenges in maintaining interpretability when extended to complex tasks.
+\Glspl{pct} are celebrated for their interpretability and efficiency, yet they often fall short in predictive performance due to their myopic, greedy selection of splits.
+The introduction of option predictive clustering trees \pbnote*{use gls}{(OPCTs)}~\citep{Stepisnik_2020} aimed to address this limitation by incorporating multiple alternative splits within option nodes.
+This approach mitigates the inherent myopia and achieves competitive performance with ensemble methods like bagging and random forests while still maintaining a degree of interpretability.
+\pbnote*{use gls}{OPCTs}, despite mitigating myopia and enhancing performance, still face challenges in maintaining interpretability when extended to complex tasks.
 
@@
 
-Despite these advancements, ensemble models such as random forests and gradient boosting machines remain popular due to their superior predictive accuracy.
+Despite these advancements, ensemble models such as random forests and gradient-boosting machines remain popular due to their superior predictive accuracy.
 However, their complexity often obscures interpretability, making it challenging for users to understand the model’s decision-making process.
 Efforts to bridge this gap include the iForest visual analytics system~\citep{Zhao_2019}, which summarises decision paths and tree structures within random forests, thereby making the ensemble's decisions more transparent.
-Additionally, the Tree Space Prototypes approach~\citep{Tan_2020} uses representative points (prototypes) for each class, offering a more intuitive understanding of ensemble classifiers compared to traditional feature-based explanations.
+Additionally, the Tree Space Prototypes approach~\citep{Tan_2020} uses representative points (prototypes) for each class, offering a more intuitive understanding of ensemble classifiers than traditional feature-based explanations.
 Visual analytics systems like iForest and prototype-based approaches offer some clarity but do not fully resolve the complexity issue.
@@
 \end{equation}
-
-Here, the model-prescribed likelihood $p(x|\theta)$ and the selected prior $p(\theta)$ are generally well-defined.
+The model-prescribed likelihood $p(x|\theta)$ and the selected prior $p(\theta)$ are generally well-defined.
 The primary computational challenge arises in determining the posterior distribution $p(\theta|x)$ due to the denominator $p(x)$, which is derived as:
@@
 \end{equation}
-
 This integral rarely yields a closed-form solution.
@@
 
-\begin{align*}
-KL(q_\omega(\theta) \parallel p(\theta|x)) = - & \mathbb{E}_q\left[\log p(x,\theta) - \log q_\omega(\theta)\right] + \log p(x)\\
-& \underbrace{\phantom{\mathbb{E}_q\left[\log p(x,\theta) - \log q_\omega(\theta)\right]}}_{\text{ELBO}}
+\begin{equation}
+KL(q_\omega(\theta) \parallel p(\theta|x)) = -\underbrace{\mathbb{E}_q\left[\log p(x,\theta) - \log q_\omega(\theta)\right]}_{\text{ELBO}}+ \log p(x)\\
 \label{eq:kl_divergence}
-\end{align*}
+\end{equation}
 
@@
 The adoption of an approximate variational distribution introduces bias contingent on the chosen variational family $\mathcal{Q}$, necessitating a decision grounded in empirical evidence or expert knowledge.
-This bias arises because the family \(\mathcal{Q}\) may not be capable of capturing the true complexity of the posterior distribution \(p_\omega(\theta)\).
+This bias arises because the family \(\mathcal{Q}\) may not be capable of capturing the true complexity of the posterior distribution~\(p_\omega(\theta)\).
 As a result, the accuracy of the approximation depends significantly on how well \(\mathcal{Q}\) aligns with the true underlying distribution.
@@
 While \gls{spyct} utilises a deterministic approach to optimise split parameters, \gls{vspyct} introduces a probabilistic framework through variational Bayes (VB) optimisation.
+\pbnote*{Maybe remove this whole part and merge it with the paragraph below since it repeats below~\eqref{eq:linear_model}.}{
 In \gls{vspyct}, the parameters of the linear combination at each split—specifically, the weight vector \(\mathbf{w}\) and the bias term \(b\)—are modelled as random variables with Gaussian distributions.
 This probabilistic representation of split parameters not only enhances the model's ability to handle uncertainty and noise in the data but also provides a measure of predictive uncertainty.
+}
 
@@
 \begin{equation}
-q(\mathbf{w}, b|\mathcal{D}) = \mathcal{N}(\mathbf{w}|\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w) \mathcal{N}(b|\mu_b, \sigma_b^2)
+q(\mathbf{w}, b|\mathcal{D}) = \mathcal{N}(\mathbf{w}|\boldsymbol{\mu}_w, \boldsymbol{\Sigma}_w) \mathcal{N}(b|\mu_b, \sigma_b^2),
 \end{equation}
-
 where \(\boldsymbol{\mu}_w\), \(\boldsymbol{\Sigma}_w\), \(\mu_b\), and \(\sigma_b^2\) are the parameters of the variational distribution, learned through the optimisation process.
@@
 \begin{equation}
-\text{Impurity} = \sum_{i=1}^{N} \left(\rho_i \cdot \text{Var}_{\rho}(\mathbf{Y}) + \lambda_i \cdot \text{Var}_{\lambda}(\mathbf{Y})\right)
+\text{Impurity} = \sum_{i=1}^{N} \left(\rho_i \text{Var}_{\rho}(\mathbf{Y}) + \lambda_i  \text{Var}_{\lambda}(\mathbf{Y})\right),
 \end{equation}
-
 where \(\rho_i = \sigma(\mathbf{w}^\top \mathbf{x}_i + b)\) is the probability of an instance being routed to the right child node, and \(\lambda_i = 1 - \rho_i\) is the probability of routing to the left child node.
 The variances \(\text{Var}_{\rho}(\mathbf{Y})\) and \(\text{Var}_{\lambda}(\mathbf{Y})\) represent the variance of the target variables in the right and left child nodes, respectively, weighted by their selection probabilities.
-
 The objective is to minimise the impurity by optimising the parameters \(\mathbf{w}\) and \(b\) using the VB method, thereby ensuring that the split at each node effectively separates the data into homogeneous subsets.
-
 The overall procedure for learning a split in \gls{vspyct} is detailed in Algorithm~\ref{alg:learn_split_vb}.
@@
 The procedure for making a prediction for a new instance using the \gls{vspyct} model is detailed in Algorithm~\ref{alg:make_pred}.
-The process involves traversing the tree from the root node to a leaf node, with decisions at each internal node being made probabilistically based on the parameters sampled from the learned variational posterior distributions.
+The process involves traversing the tree from the root node to a leaf node, with decisions at each internal node being made \pbnote*{what do you mean by this?
+Are you referring to line 16 in algorithm 2?}{probabilistically} based on the parameters sampled from the learned variational posterior distributions.
 The final prediction is determined by the prototype value stored in the reached leaf node.
@@
 where \(\epsilon\) is a small constant added for numerical stability.
+\pbnote{$\text{Var}(\mathbf{w}_s) > 0$, why do you need $\epsilon$?
+I know that we were talking about this, but there is a concept called coefficient of variation and it is defined as $\sigma/\mu$, std.
+(not variance) vs mean.
+We have to discuss this.}
 
@@
 
+\pbnote{For the toy example below, I would suggest to change the name from toy example to something else. Or completely remove the title.
+Instead of using \emph{age} and \emph{salary} you can use $x_1$ and $x_2$.}
 \paragraph{Toy Example}
@@
 Suppose that at the root node, the mean weights associated with \textit{age} and \textit{salary} are \( \mathbb{E}[w_{\text{age}}] = 0.2 \) and \( \mathbb{E}[w_{\text{salary}}] = 0.8 \), and their respective variances are \( \text{Var}(w_{\text{age}}) \) and \( \text{Var}(w_{\text{salary}}) \).
-The importance at this node will now reflect the ratio \( \frac{\mathbb{E}[w]}{\text{Var}(w)} \), such that a feature with a relatively high mean weight but also a high variance will not necessarily dominate one with a slightly lower mean but substantially more stable (lower variance) weights.
+The importance at this node will now reflect the ratio \( \frac{\mathbb{E}[w]}{\text{Var}(w)} \), such that a feature with a relatively high mean weight but also a high variance will not necessarily dominate one with a slightly lower mean but substantially more \pbnote*{I don't know wheter the word stable is a correct one.
+May be precise because precission is the inverse of the variance.}{stable} (lower variance) weights.
 Thus, if \textit{salary} maintains a higher \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio than \textit{age}, it indicates that \textit{salary} is not only influential but also more reliably so at this particular node.
@@
 \section{Experimental setting}
-
+\pbnote{The paragraph below is a little messy. I assume the idea is to say that we compare VBSPYCT on all benchmarks where OPCT and ensemble PCT were evaluated? If so write it like that.}
 The experimental evaluation was conducted to compare the \gls{vspyct} model with several benchmarks, including the \gls{spyct} model, \glspl{opct}~\citep{Stepisnik_2020}, and ensemble of \glspl{pct}.
@@
 
-In STR tasks, the output is a single continuous variable, while in MTR tasks, the output comprises multiple continuous target variables.
+\mpop{In STR tasks, the output is a single continuous variable, while in MTR tasks, the output comprises multiple continuous target variables.
 BC tasks involve two possible classes, and MCC tasks extend to multiple classes, with the exact number of classes depending on the specific dataset.
-This comprehensive range of datasets ensures a robust evaluation of the models across varying complexity and output structures.
+This comprehensive range of datasets ensures a robust evaluation of the models across varying complexity and output structures.}{}
 
@@
 
-Figure~\ref{fig:avg_rank} shows the average rankings of the evaluated methods—\gls{vspyct}, \gls{opct}, single \gls{spyct}, ensemble of \glspl{spyct}, and ensemble of \glspl{pct}—across four task types: binary classification (BC), multi-class classification (MCC), single-target regression (STR), and multi-target regression (MTR).
-For improved readability, the actual performance of the models on each dataset is given in the Appendix.
+Figure~\ref{fig:avg_rank} shows the average rankings of the evaluated methods \gls{vspyct}, \gls{opct}, single \gls{spyct}, ensemble of \glspl{spyct}, and ensemble of \glspl{pct} across four task types.
+For improved readability, the actual performance of the models on each dataset is given in~\ref{app:res}.
 
@@
 \subsection{Feature importance scores}
+\pbnote{Here you are missing some introduction why do you need this subsection.
+Furthermore, when reading the complete subsection you have to clarify how did you add noisy features. What is the distribution of these features with respect to the original features?
+Does this means that we have fearures that are completely useless in the original data, that even noise that does not influence the output is more important (since we have blue dots below the orange ones)?
+One would expect only cases like in columns 2 and 3. 
+}
+
+\pbnote{
+Just to be clear, originally you have this true model:
+$$y=model(\mathcal{F}).$$
+Based on this you fit one tree.
+The you make
+$$y_1=y+0\cdot\mathcal{F}'.$$
+Now when I wrote it, it seems rather stupid notation.
+Anyway, my question is how did you introduce this features.
+Did you just add columns in the dataset with noise which in essence makes my model with $0\cdot\mathcal{F}'$ the correct explenation how you added the additional features?
+}
+
+\pbnote{
+Is there a citation that can be used that discribes this way of thinking, how to evaluate feature importance metrics?
+Or this is completely your idea?
+}
 
@@
 
-\section*{Appendix}
+\appendix
+\section{Detailed results}
+\label{app:res}
 

### 2025-01-23 reports/paper/vspyct.tex
paper changes and find_threshold script

@@
 In the evolving field of machine learning, the fusion of semi-supervised learning and decision tree models has led to the development of models like \glspl{pct}~\citep{Kocev_2013} and their advanced variant, \glspl{spyct}~\citep{Stepi_nik_2021}.
-These models leverage labeled and unlabelled data to enhance prediction accuracy and efficiency.
+These models leverage labeled and unlabelled data to enhance predictive accuracy and efficiency.
 \Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
-Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal performance.
+Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
+\vanote{Ensemble methods, such as random forests and gradient-boosted trees, achieve high predictive performance by reducing variance and/or bias through the aggregation of diverse models.
+These techniques ensure robustness by capturing slightly different patterns or residuals from the data.}
 While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, which is crucial for informed decision-making across various applications.
@@
 \gls{vspyct} integrates the variational Bayes method~\citep{Blei_2017} to facilitate complex decision-making within a singular model framework, thus maintaining the interpretability inherent to decision trees without necessitating ensembles.
+\vanote{Unlike ensembles, which explicitly combine multiple \gls{spyct} trees, a single \gls{vspyct} attempts to incorporate uncertainty by sampling from the posterior distribution of tree parameters.}
 Furthermore, by embedding model-specific uncertainty quantification directly into the decision tree structure through Bayesian inference, \gls{vspyct} enhances the depth of insight into the model's decision process and confidence levels.
 %This approach not only elevates the explainability of semi-supervised decision trees by consolidating predictive capabilities into a single model but also marks a significant advancement in machine learning, particularly within semi-supervised learning domains.
-Combining the robustness of \gls{spyct} ensembles with the clarity and interpretability of a single tree model, alongside introducing uncertainty quantification, our research provides a comprehensive tool to overcome the current limitations of predictive clustering methodologies.
+Combining the robustness of \gls{spyct} ensembles with the clarity and interpretability of a single tree model, alongside introducing uncertainty quantification, our research provides a comprehensive tool as an alternative with a goal to overcome the current limitations of predictive clustering methodologies.
 
@@
 \Glspl{pct} are celebrated for their interpretability and efficiency, yet they often fall short in predictive performance due to their myopic, greedy selection of splits.
-The introduction of option predictive clustering trees \pbnote*{use gls}{(OPCTs)}~\citep{Stepisnik_2020} aimed to address this limitation by incorporating multiple alternative splits within option nodes.
+The introduction of \glspl{opct}~\citep{Stepisnik_2020} aimed to address this limitation by incorporating multiple alternative splits within option nodes.
 This approach mitigates the inherent myopia and achieves competitive performance with ensemble methods like bagging and random forests while still maintaining a degree of interpretability.
-\pbnote*{use gls}{OPCTs}, despite mitigating myopia and enhancing performance, still face challenges in maintaining interpretability when extended to complex tasks.
+\Glspl{opct}, despite mitigating myopia and enhancing performance, still face challenges in maintaining interpretability when extended to complex tasks.
 
@@
 Similar to \gls{spyct}, \gls{vspyct} constructs a decision tree with an oblique structure, where each split is defined by a linear combination of input features, rather than relying on single-feature thresholds as in axis-aligned trees.
-
 The key distinction between \gls{vspyct} and \gls{spyct} lies in the optimisation step used to determine the parameters of these oblique splits.
 While \gls{spyct} utilises a deterministic approach to optimise split parameters, \gls{vspyct} introduces a probabilistic framework through variational Bayes (VB) optimisation.
-\pbnote*{Maybe remove this whole part and merge it with the paragraph below since it repeats below~\eqref{eq:linear_model}.}{
-In \gls{vspyct}, the parameters of the linear combination at each split—specifically, the weight vector \(\mathbf{w}\) and the bias term \(b\)—are modelled as random variables with Gaussian distributions.
-This probabilistic representation of split parameters not only enhances the model's ability to handle uncertainty and noise in the data but also provides a measure of predictive uncertainty.
-}
 
@@
 where \(\mathbf{x} \in \mathbb{R}^N\) is the feature vector, \(\mathbf{w} \in \mathbb{R}^N\) represents the weight vector, \(b \in \mathbb{R}\) is the bias term, and \(\sigma(\cdot)\) denotes the sigmoid function, which ensures the output is a probability, determining whether an instance is sent to the right or left child node.
-
 While the tree structure and the decision-making process follow the same principles as in \gls{spyct}, the optimisation of \(\mathbf{w}\) and \(b\) in \gls{vspyct} differs fundamentally.
@@
 
+\vanote{While parameter uncertainty is effectively addressed, structural uncertainty is inherently more challenging to capture in a single tree.
+The variational Bayes approach approximates the posterior over parameters, but the inclusion of priors on tree structure (e.g., split selection) is an area for further exploration.}
+
 \begin{algorithm}[h!]
@@
 The procedure for making a prediction for a new instance using the \gls{vspyct} model is detailed in Algorithm~\ref{alg:make_pred}.
-The process involves traversing the tree from the root node to a leaf node, with decisions at each internal node being made \pbnote*{what do you mean by this?
-Are you referring to line 16 in algorithm 2?}{probabilistically} based on the parameters sampled from the learned variational posterior distributions.
+The process involves traversing the tree from the root node to a leaf node, with decisions at each internal node being made based on the parameters sampled from the learned variational posterior distributions.
 The final prediction is determined by the prototype value stored in the reached leaf node.
@@
 This method accounts for the uncertainty in the model parameters, providing a probabilistic framework for decision-making.
+The prediction \(\hat{y}\) returned by the model is the prototype of the reached leaf node, where the prototype is defined as the mean of the target values associated with the training instances that were routed to that leaf during training.
 
-The prediction \(\hat{y}\) returned by the model is the prototype of the reached leaf node, where the prototype is defined as the mean of the target values associated with the training instances that were routed to that leaf during training.
+\vanote{The posterior sampling allows the model to integrate over the uncertainty of the parameters at each node, providing smoothed and robust decision boundaries.
+The model captures variability in the decision process, reducing the risk of overfitting to a single parameter set.
+While this approach effectively accounts for parameter uncertainty, the reliance on a single tree structure limits the model’s ability to replicate the diversity inherent in ensemble methods.
+Ensembles, achieve variance reduction by combining predictions from multiple distinct structures, a capability that a single \gls{vspyct} cannot fully emulate.}
 
@@
 
-\pbnote{For the toy example below, I would suggest to change the name from toy example to something else. Or completely remove the title.
-Instead of using \emph{age} and \emph{salary} you can use $x_1$ and $x_2$.}
-\paragraph{Toy Example}
-Consider a simple dataset where two features, \textit{age} and \textit{salary}, predict the likelihood of purchasing a car.
+Consider a simple dataset where two features, $x_1$ and $x_2$, predict the likelihood of purchasing a car.
 In a \gls{vspyct} model, the influence of each feature at a node depends not only on the magnitude of its mean weight but also on how this mean compares to the feature’s weight variance.
 
-Suppose that at the root node, the mean weights associated with \textit{age} and \textit{salary} are \( \mathbb{E}[w_{\text{age}}] = 0.2 \) and \( \mathbb{E}[w_{\text{salary}}] = 0.8 \), and their respective variances are \( \text{Var}(w_{\text{age}}) \) and \( \text{Var}(w_{\text{salary}}) \).
-The importance at this node will now reflect the ratio \( \frac{\mathbb{E}[w]}{\text{Var}(w)} \), such that a feature with a relatively high mean weight but also a high variance will not necessarily dominate one with a slightly lower mean but substantially more \pbnote*{I don't know wheter the word stable is a correct one.
-May be precise because precission is the inverse of the variance.}{stable} (lower variance) weights.
-Thus, if \textit{salary} maintains a higher \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio than \textit{age}, it indicates that \textit{salary} is not only influential but also more reliably so at this particular node.
+Suppose that at the root node, the mean weights associated with $x_1$ and $x_2$ are \( \mathbb{E}[w_{x_1}] = 0.2 \) and \( \mathbb{E}[w_{x_2}] = 0.8 \), and their respective variances are \( \text{Var}(w_{x_1}) \) and \( \text{Var}(w_{x_2}) \).
+The importance at this node will now reflect the ratio \( \frac{\mathbb{E}[w]}{\text{Var}(w)} \), such that a feature with a relatively high mean weight but also a high variance will not necessarily dominate one with a slightly lower mean but substantially lower variance.
+Thus, if $x_2$ maintains a higher \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio than $x_1$, it indicates that $x_2$ is not only influential but also more reliable at this particular node.
 
 However, the overall feature importance is derived by aggregating these normalised ratios across all splits within the tree.
-Even if \textit{salary} appears strongly influential at the root node, the final importance calculation considers both the magnitude and the stability of the weights across all nodes.
+Even if $x_2$ appears strongly influential at the root node, the final importance calculation considers both the magnitude and the stability of the weights across all nodes.
 A feature consistently demonstrating a favourable \(\frac{\mathbb{E}[w]}{\text{Var}(w)}\) ratio throughout multiple splits will be deemed more important globally.
@@
 
-\mpop{In STR tasks, the output is a single continuous variable, while in MTR tasks, the output comprises multiple continuous target variables.
-BC tasks involve two possible classes, and MCC tasks extend to multiple classes, with the exact number of classes depending on the specific dataset.
-This comprehensive range of datasets ensures a robust evaluation of the models across varying complexity and output structures.}{}
-
 \subsection{Evaluation}
@@
 
-\subsection{Feature importance scores}
-\pbnote{Here you are missing some introduction why do you need this subsection.
-Furthermore, when reading the complete subsection you have to clarify how did you add noisy features. What is the distribution of these features with respect to the original features?
-Does this means that we have fearures that are completely useless in the original data, that even noise that does not influence the output is more important (since we have blue dots below the orange ones)?
-One would expect only cases like in columns 2 and 3. 
-}
+\subsection{Robustness to Spurious Features}%
+\label{sec:noise_features}
 
+\begin{comment}
 \pbnote{
@@
 }
+\end{comment}
+
+A critical step in validating the proposed \gls{vspyct} model is to test its robustness against spurious features. 
+In many real-world scenarios, the data may contain variables that do not contribute to the predictive task at hand -- hence, it is paramount to ensure that the model assigns low importance to such non-informative features.
+Accordingly, this subsection investigates whether \gls{vspyct} can distinguish relevant features from artificially generated noise.
 
@@
     \includegraphics{../figures/feat_importances.pdf}
-    \caption{Feature importance scores obtained on a single dataset from per category. The results indicate that the model does not depend on irrelevant features.}
+    \caption{Feature importance scores for both the original and noise-augmented datasets.
+    Orange markers represent the importance of original features, whereas blue markers indicate the importance of noise features.}
     \label{fig:feature_importances}
@@
 
-As illustrated in Figure~\ref{fig:feature_importances}, the proposed \gls{vspyct} model yields meaningful feature importance scores.
-To rigorously evaluate this property, we introduced synthetic noise features into the datasets before model training.
-Let the original dataset contain \(d\) features.
-We then augmented it by adding \(d\) additional features, each generated by random noise.
-Thus, the total number of features in the augmented dataset is \(2d\).
+As illustrated in Figure~\ref{fig:feature_importances}, the \gls{vspyct}-derived importance measures generally assign higher scores to features that provide genuine predictive information.
+To rigorously evaluate the extent to which the model remains robust to irrelevant variables, let the original dataset contain \(d\) features, denoted by the set
+\(\mathcal{F} = \{f_1, f_2, \ldots, f_d\}\).
+We augment this dataset by adding \(d\) synthetic noise features
+\(\mathcal{F}' = \{f_{d+1}, \ldots, f_{2d}\}\),
+each of which is drawn independently from a Gaussian distribution with mean \(0\) and variance \(1\):
+\[
+f_{d + j} \sim \mathcal{N}(0, 1), 
+\quad \text{for } j = 1, \ldots, d.
+\]
+Hence, the total number of features in the augmented dataset is \(2d\).
 
-The resulting feature importance scores reveal that the original (real) features exhibit a broad range of importance values.
-This heterogeneity is expected, as real-world datasets often comprise both highly informative variables and those with negligible predictive power.
-In contrast, the artificially introduced noise features consistently received low importance scores.
-Formally, if we let \(\mathcal{F} = \{f_1, f_2, \ldots, f_d\}\) denote the set of original features, and \(\mathcal{F}' = \{f_{d+1}, \ldots, f_{2d}\}\) the set of noise features, the estimated importances \(I(f_j)\) for \(f_j \in \mathcal{F}'\) are significantly lower than those for at least a subset of \(\mathcal{F}\).
-This indicates that the \gls{vspyct}-derived importance measures are not only able to identify relevant features but also robust to the inclusion of spurious, non-informative features.
+Figure~\ref{fig:feature_importances} reveals that most of the original (real) features exhibit a broad range of importance values.
+This heterogeneity is expected because many real-world datasets naturally include some highly informative predictors, as well as others that provide minimal predictive power.
+Importantly, the noise features typically receive lower importance values, confirming that \gls{vspyct} can detect their lack of relevance.
+Formally, if \(I(f)\) denotes the estimated importance of a feature \(f\), then
+\[
+I(f_j) \ll I(f_k)
+\quad \text{for at least a subset of } f_j \in \mathcal{F}' \text{ and } f_k \in \mathcal{F}.
+\]
+In other words, noise features \(\{f_{d+1}, \ldots, f_{2d}\}\) are consistently outperformed by at least some subset of the original features.
+
+Noteworthy, a few original features may show importance values that are comparable to or even lower than those of some noise features. 
+This outcome does not necessarily imply that the noise features carry genuine signal; rather, it highlights that certain original features are also uninformative for the target variable.
+Overall, these results confirm that the \gls{vspyct}-derived feature importance measures reliably rank truly predictive variables above most spurious ones, underscoring the model's robustness to irrelevant inputs.
 
@@
 Figure~\ref{fig:feature_importance} highlights the feature importance scores obtained from the \gls{vspyct} model, focusing on the most influential features that contribute to predicting the unemployment duration.
-The horizontal bars show the relative importance of each feature, with \textit{Months of work experience} emerging as the most critical factor, followed by \textit{Age} and \textit{Entry month}.
+The horizontal bars show the relative importance of each feature, with \textit{Entry month} emerging as the most critical factor, followed by \textit{Months of work experience} and \textit{ISCO\_Health associate professionals}.
 
@@
     \centering
-    \includegraphics[width=1.0\textwidth]{feature_importance.pdf}
+    \includegraphics[width=1.0\textwidth]{feature_importance_colored.pdf}
     \caption{Feature importance scores extracted from the \gls{vspyct} model applied to the Unemployment dataset.
-The horizontal bars represent the relative importance of each feature, with the most influential feature being "Months of work experience."}
+The horizontal bars represent the relative importance of each feature, with the most influential feature being "Entry month"}
     \label{fig:feature_importance}
@@
     \centering
-    \includegraphics{hecat_tree_old_temp.pdf}
+    \includegraphics{hecat_tree_new_temp.pdf}
     \caption{The \gls{vspyct} model applied to the Unemployment dataset. The colouring distinguishes the different feature groups of the data set}
@@
 
-Figure~\ref{fig:inter_tree} presents the visual representation of the \gls{vspyct} applied to the Unemployment dataset.
-Based on that, one can distill a concise interpretation enriched with practical insights.
+\vanote{Figure~\ref{fig:inter_tree} presents the visual representation of the \gls{vspyct} applied to the Unemployment dataset.
+It provides clear insights into how different feature groups influence predictions of unemployment duration.
+Each node represents a decision point, with the edges coloured based on the dominant feature group—\textit{ISCO} (occupational classifications), \textit{P16} (education-related features), or demographic features—used in the split.
+This colour-coded representation highlights the relative importance of feature groups across different levels of the tree.}
 
-At the heart of this \gls{vspyct} model lies the ability to visually demarcate the influence of various feature categories on unemployment duration predictions.
-The tree's root node is predominantly influenced by features categorised under ``other" (represented in green), encompassing age, months of work experience, and various demographic characteristics.
-This indicates that such features play a pivotal role in initial split decisions, underscoring the broad impact of demographic and experiential factors on unemployment trends.
+\vanote{Demographic features dominate higher levels of the tree, indicating their broad and primary influence on unemployment duration.
+Features such as \textit{Months of work experience} and \textit{Age} frequently appear at the root or upper nodes, suggesting that prior professional experience and age are strong, general predictors of re-employment likelihood.
+For instance, individuals with less work experience are often routed towards branches with longer unemployment times, reflecting the challenges faced by those lacking experience in securing jobs.}
 
-As we navigate through the tree, a dichotomy emerges based on the expected survival times—essentially, the projected duration an individual is likely to remain unemployed.
-On branches leading to higher expected survival times, features categorised under ``p16 level of education" (coloured in blue) become increasingly dominant.
-This suggests that higher levels of education are associated with longer periods of unemployment, potentially reflecting challenges in finding employment that matches an individual's qualifications or aspirations.
+\vanote{In contrast, ISCO features become more prominent in lower-level splits, capturing occupation-specific influences on unemployment.
+Branches dominated by ISCO-related splits suggest that certain occupational groups have distinct re-employment trends.
+For example, occupations such as \textit{Health associate professionals} or \textit{Labourers in mining, construction, manufacturing, and transport} are likely associated with specific labor market demands that either shorten or extend unemployment durations.}
 
-Conversely, on branches associated with lower expected survival times, the dominance shifts towards features represented by the ``ISCO classification of professional occupation" (coloured in red).
-This dominance implies that certain professional occupations, as defined by the ISCO classification, may expedite reemployment, possibly due to higher demand or greater flexibility in job roles within these sectors.
+\vanote{P16 features emerge in intermediate levels of the tree, reflecting the nuanced role of education in employment outcomes.
+For instance, splits involving education levels such as \textit{Business, administration, and law} or \textit{Services} suggest that individuals’ educational backgrounds influence their unemployment duration in ways that interact with occupational or demographic factors.}
 
-This visual breakdown not only highlights the model's interpretability but also offers tangible insights into how different factors—ranging from personal demographics and work experience to education and professional classifications—impact unemployment durations.
-Such insights are invaluable for policymakers and stakeholders aiming to devise targeted interventions that address specific barriers to employment faced by diverse demographic groups.
+\vanote{The leaf nodes, representing the predicted expected survival times (i.e., unemployment duration in days), provide a clear endpoint for each pathway.
+Branches leading to longer survival times often involve demographic splits, highlighting structural barriers such as insufficient work experience or higher age, while branches associated with shorter durations often incorporate ISCO and P16 splits, reflecting the effect of occupation- and education-specific factors.}
 
@@
 We introduced the \gls{vspyct} model, an approach that combines the benefits of oblique splits, variational Bayes, and Bayesian inference within a single, interpretable predictive framework.
-The \gls{vspyct} model addresses the trade-off between the performance of ensemble models and the interpretability of single decision trees, and the absence of inherent uncertainty quantification in predictions.
-Our results demonstrate that the \gls{vspyct} model is competitive to, and on some datasets exceeds the performance of ensemble of \glspl{spyct}, without sacrificing the model's interpretability.
-By employing variational Bayes method, we introduce a mechanism for uncertainty quantification directly into the decision-making process, enhancing the reliability and applicability of the model across various domains where decision certainty is paramount.
-%The \gls{vspyct} model marks a significant step forward in making complex decision trees more accessible, interpretable, and effective.
+The \gls{vspyct} model addresses the trade-off between the performance of ensemble models and the interpretability of single decision trees, while also tackling the absence of inherent uncertainty quantification in predictions.
+
+Our results demonstrate that the \gls{vspyct} model is competitive with, and on some datasets exceeds, the performance of ensembles of \glspl{spyct}, without sacrificing the model's interpretability.
+By employing the variational Bayes method, we introduce a mechanism for uncertainty quantification directly into the decision-making process, enhancing the reliability and applicability of the model across various domains where decision certainty is paramount.
+
 The model opens new possibilities for research in semi-supervised learning and Bayesian inference, promising to bridge the gap between theoretical advancements and practical applications.
-Future work will focus on further refining the model's performance, and extending its applications in diverse predictive tasks.
+\vanote{Future work will focus on further refining the model's performance, exploring the inclusion of soft splits to enable smoother decision boundaries and probabilistic routing, and extending its applications in diverse predictive tasks.}
 

### 2025-01-23 find_threshold.py

@@ -1,0 +1,138 @@
+import pickle
+from sklearn.metrics import mean_absolute_error
+
+
+import os
+import sys
+import numpy as np
+import pandas as pd
+
+# sys.path.append('..')
+from src.models.model import VSpyct
+
+import torch
+import pyro
+import random
+
+torch.manual_seed(0)
+torch.cuda.manual_seed(0)  # Set seed for CUDA
+random.seed(0)
+np.random.seed(0)
+
+
+# --- Hyperparams / Setup ---
+max_depth = 5
+minimum_examples_to_split = 2
+epochs = 500
+bs = 128
+lr = 0.01
+subspace_size = 1
+device = 'cpu'
+
+# --- Load data ---
+with open('cpmp2015_str_files.pcl', 'rb') as f:
+    X_train, X_test, y_train, y_test = pickle.load(f)
+
+with open('cpmp2015_str_feat_imp.pcl', 'rb') as f:
+    feature_importance_scores = pickle.load(f)
+
+feature_names = X_train.columns.tolist()
+features_dict = dict(zip(feature_names, feature_importance_scores))
+
+# Sort features by importance ascending so that
+# the "least important" feature is first.
+sorted_features_least_first = sorted(features_dict, key=features_dict.get)
+
+# --- Define a helper function to train & evaluate ---
+def train_and_evaluate_model(X_train, y_train, X_test, y_test, features):
+    """
+    Subset the training/test data to the given 'features',
+    train the VSpyct model, and return the mean_absolute_error on test.
+    """
+    # Subset columns
+    X_train_sub = X_train[features]
+    X_test_sub = X_test[features]
+    
+    # Initialize model
+    vspyct = VSpyct(
+        bs=bs,
+        max_depth=max_depth,
+        epochs=epochs,
+        lr=lr,
+        minimum_examples_to_split=minimum_examples_to_split
+    )
+    
+    # Fit
+    vspyct.fit(torch.Tensor(X_train_sub.values), torch.Tensor(y_train.values))
+    
+    # Predict
+    vspyct_preds = vspyct.predict(torch.Tensor(X_test_sub.values))
+    vspyct_preds_mean = vspyct_preds.mean(axis=1).numpy()
+    
+    # Score
+    score = mean_absolute_error(y_test.values, vspyct_preds_mean)
+    return score
+
+# --- 1) Baseline score with ALL features ---
+current_feature_set = list(feature_names)  # start with all features
[…]

### 2025-01-23 src/data/dataset.py

@@ -256,7 +256,7 @@ class ReducedDataset:
 
     def __init__(self, path, filename, random_state=11):
         self.dataset = pd.read_pickle(os.path.join(path, filename))
-        self.dataset = self.dataset.sample(frac=0.005)
+        self.dataset = self.dataset.sample(frac=0.1)
         self.random_state = random_state
         self.skp = pd.read_csv(os.path.join(DIR_PATH, '../../data/raw/dimSKP08.csv'))[['IDpoklicaSKP','SFpoklicaSKP']]
         self.skp.SFpoklicaSKP = self.skp.SFpoklicaSKP.astype(int).astype(str).str.zfill(4)
@@ -392,6 +392,7 @@ class ReducedDataset:
     def rsf_split(self, test = 0.2, to_pcl = False):
         X = self.rsf_dataset().drop(columns=['duration', 'truncated'])
         y = self.get_survival_vector()
+        # y = torch.tensor(self.rsf_dataset().duration.fillna(self.rsf_dataset().duration.mean()).values)
         from sklearn.model_selection import train_test_split
         X_train, X_test, y_train, y_test = train_test_split(X, y, test_size = test, random_state = self.random_state)
         print(X_train.shape, y_train.shape)
@@ -425,7 +426,7 @@ class ReducedDataset:
         time_tensor = torch.empty((survival_time.shape[0], time_upper_limit))
         for i,el in enumerate(survival_time):
             for j in range(time_tensor.shape[1]):
-                if (survival_status[i] == 1) and (j>=el): time_tensor[i][j] = np.nan
+                if (survival_status[i] == 1) and (j>=el): time_tensor[i][j] = 0 #np.nan
                 elif j<el: time_tensor[i][j] = 1
                 else: time_tensor[i][j] = 0
         return time_tensor

### 2025-01-23 find_threshold.py
find_threshold change

@@ -43,6 +43,7 @@ features_dict = dict(zip(feature_names, feature_importance_scores))
 # the "least important" feature is first.
 sorted_features_least_first = sorted(features_dict, key=features_dict.get)
 
+
 # --- Define a helper function to train & evaluate ---
 def train_and_evaluate_model(X_train, y_train, X_test, y_test, features):
     """
@@ -52,7 +53,7 @@ def train_and_evaluate_model(X_train, y_train, X_test, y_test, features):
     # Subset columns
     X_train_sub = X_train[features]
     X_test_sub = X_test[features]
-    
+
     # Initialize model
     vspyct = VSpyct(
         bs=bs,
@@ -61,78 +62,37 @@ def train_and_evaluate_model(X_train, y_train, X_test, y_test, features):
         lr=lr,
         minimum_examples_to_split=minimum_examples_to_split
     )
-    
+
     # Fit
     vspyct.fit(torch.Tensor(X_train_sub.values), torch.Tensor(y_train.values))
-    
+
     # Predict
     vspyct_preds = vspyct.predict(torch.Tensor(X_test_sub.values))
     vspyct_preds_mean = vspyct_preds.mean(axis=1).numpy()
-    
+
     # Score
     score = mean_absolute_error(y_test.values, vspyct_preds_mean)
     return score
 
-# --- 1) Baseline score with ALL features ---
-current_feature_set = list(feature_names)  # start with all features
-baseline_score = train_and_evaluate_model(X_train, y_train, X_test, y_test, current_feature_set)
-print(f"Baseline MAE: {baseline_score:.4f}")
 
-removed_features = []
-scores = [baseline_score]  # keep track of scores at each step
+def main():
+    # Get number of features to remove from command line argument
+    print(len(sys.argv))
+    if len(sys.argv) != 2:
+        print("Usage: python script.py <num_features_to_remove>")
+        sys.exit(1)
 
-# --- 2) Iteratively remove the least important feature ---
-for f in sorted_features_least_first:
-    # Create a candidate set that removes 'f'
-    candidate_feature_set = [feat for feat in current_feature_set if feat != f]
-    
-    # Train & evaluate
-    candidate_score = train_and_evaluate_model(
-        X_train, y_train, 
-        X_test, y_test, 
-        candidate_feature_set
-    )
-    
-    # Check if removing this feature degrades performance by >10% 
-    # i.e. if MAE is 10% higher than the initial baseline
-    if candidate_score > baseline_score * 1.10:
[…]

### 2025-01-23 src/models/model.py
new feat imp cpmp

@@ -262,6 +262,7 @@ class VSpyct:
         
         _traverse(self.root_node)
         weights = []
+        vars = []
         for node in non_leaves:
             predictive = Predictive(model=node.split_model.to(self.device),
                                     guide=node.guide,
@@ -269,12 +270,13 @@ class VSpyct:
                                     return_sites=(["linear.weight"]))
             data = predictive(node.descriptive_values)['linear.weight']
             weights.append(data.mean(axis=0).numpy())
+            vars.append(data.var(axis=0).numpy())
         weights = np.abs(np.array(weights))
         weights = weights.reshape(weights.shape[0], -1)
         weights_normalized = weights # (weights - weights.min()) / (weights.max() - weights.min() + 0.0001)
         importances = torch.zeros((weights_normalized.shape[1]))
         for i, node in enumerate(non_leaves):
-            importances += (node.num_instances/self.num_training_instances)*(weights_normalized[i, :]/(np.linalg.norm(weights_normalized[i, :])+ 0.0001))
+            importances += (node.num_instances/self.num_training_instances)*(weights_normalized[i, :]/(vars[i]+ 0.0001))
         if k is not None: return dict(zip(torch.topk(importances, k=k).indices.tolist(), torch.topk(importances, k=k).values.tolist()))
         else: return importances
         

### 2025-01-24 find_threshold.py
updated script and new dataset pickles

@@ -1,11 +1,11 @@
 import pickle
 from sklearn.metrics import mean_absolute_error
 
-
 import os
 import sys
 import numpy as np
 import pandas as pd
+import argparse
 
 # sys.path.append('..')
 from src.models.model import VSpyct
@@ -19,7 +19,6 @@ torch.cuda.manual_seed(0)  # Set seed for CUDA
 random.seed(0)
 np.random.seed(0)
 
-
 # --- Hyperparams / Setup ---
 max_depth = 5
 minimum_examples_to_split = 2
@@ -28,20 +27,6 @@ bs = 128
 lr = 0.01
 subspace_size = 1
 device = 'cpu'
-
-# --- Load data ---
-with open('cpmp2015_str_files.pcl', 'rb') as f:
-    X_train, X_test, y_train, y_test = pickle.load(f)
-
-with open('cpmp2015_str_feat_imp.pcl', 'rb') as f:
-    feature_importance_scores = pickle.load(f)
-
-feature_names = X_train.columns.tolist()
-features_dict = dict(zip(feature_names, feature_importance_scores))
-
-# Sort features by importance ascending so that
-# the "least important" feature is first.
-sorted_features_least_first = sorted(features_dict, key=features_dict.get)
 
 
 # --- Define a helper function to train & evaluate ---
@@ -76,23 +61,38 @@ def train_and_evaluate_model(X_train, y_train, X_test, y_test, features):
 
 
 def main():
-    # Get number of features to remove from command line argument
-    print(len(sys.argv))
-    if len(sys.argv) != 2:
-        print("Usage: python script.py <num_features_to_remove>")
-        sys.exit(1)
+    # Parse command-line arguments
+    parser = argparse.ArgumentParser(description="Train and evaluate VSpyct model with feature removal.")
+    parser.add_argument("num_features_to_remove", type=int, help="Number of least important features to remove.")
+    parser.add_argument("data_pickle", type=str, help="Path to the data pickle file.")
+    parser.add_argument("feature_importance_pickle", type=str, help="Path to the feature importance pickle file.")
+    args = parser.parse_args()
 
-    num_features_to_remove = int(sys.argv[1])
+    # Load data
+    with open(args.data_pickle, 'rb') as f:
+        X_train, X_test, y_train, y_test = pickle.load(f)
 
-    # Ensure num_features_to_remove is within valid range
-    if num_features_to_remove < 0 or num_features_to_remove >= len(sorted_features_least_first):
[…]

### 2025-01-24 src/models/model.py

@@ -273,6 +273,8 @@ class VSpyct:
             vars.append(data.var(axis=0).numpy())
         weights = np.abs(np.array(weights))
         weights = weights.reshape(weights.shape[0], -1)
+        vars = np.abs(np.array(vars))
+        vars = vars.reshape(vars.shape[0], -1)
         weights_normalized = weights # (weights - weights.min()) / (weights.max() - weights.min() + 0.0001)
         importances = torch.zeros((weights_normalized.shape[1]))
         for i, node in enumerate(non_leaves):

### 2025-01-24 find_threshold.py
fix numpy in find_threshold.py

@@ -49,7 +49,7 @@ def train_and_evaluate_model(X_train, y_train, X_test, y_test, features):
     )
 
     # Fit
-    vspyct.fit(torch.Tensor(X_train_sub.values), torch.Tensor(y_train.values))
+    vspyct.fit(torch.Tensor(X_train_sub.values), torch.Tensor(y_train))
 
     # Predict
     vspyct_preds = vspyct.predict(torch.Tensor(X_test_sub.values))

### 2025-02-06 reports/paper/vspyct.tex
feature importance changes in paper

@@
 \section{Experimental setting}
-\pbnote{The paragraph below is a little messy. I assume the idea is to say that we compare VBSPYCT on all benchmarks where OPCT and ensemble PCT were evaluated? If so write it like that.}
+
 The experimental evaluation was conducted to compare the \gls{vspyct} model with several benchmarks, including the \gls{spyct} model, \glspl{opct}~\citep{Stepisnik_2020}, and ensemble of \glspl{pct}.
-Specifically, we considered both a single \gls{spyct} and an ensemble of \glspl{spyct}, as well as a single \gls{opct} and an ensemble of \glspl{pct}.
+Specifically, we considered both a single \gls{spyct} and an ensemble of \glspl{spyct}, an ensemble of \glspl{pct}, and the \gls{opct} model.
 We chose these methods as primary comparisons because they have been extensively evaluated in the literature, demonstrating strong performance across various \gls{sop} tasks~\citep{Andonovikj_2024, Stepi_nik_2021}.
@@
 The comparison encompassed a range of predictive modelling tasks, including single-target regression, multi-target regression, binary classification, and multi-class classification.
-By evaluating not only a single \gls{spyct} tree, but also an ensemble of \glspl{spyct}, a single \gls{opct} tree, and an ensemble of \glspl{pct}, we aimed to assess how \gls{vspyct}'s predictive performance compares against a spectrum of models, from single decision trees to more complex ensembles and option-based trees.
-This comprehensive approach allowed us to rigorously evaluate \gls{vspyct}'s effectiveness across different modelling scenarios and relative to both well-established baseline models and state-of-the-art tree-based methods.
+By evaluating against a single \gls{spyct} tree, an ensemble of \glspl{spyct}, \gls{opct} tree, and an ensemble of \glspl{pct}, we aimed to assess how \gls{vspyct}'s predictive performance compares against a spectrum of models, from single decision trees to more complex ensembles and option trees.
 
@@
     \subfloat[STR]{%
-        \includegraphics{critical_distance_str.pdf}
+        \includegraphics{critical_distance_str_2.pdf}
         \label{fig:subfig3}
@@
 In many real-world scenarios, the data may contain variables that do not contribute to the predictive task at hand -- hence, it is paramount to ensure that the model assigns low importance to such non-informative features.
-Accordingly, this subsection investigates whether \gls{vspyct} can distinguish relevant features from artificially generated noise.
+Accordingly, this subsection delves into whether \gls{vspyct} can distinguish relevant features from artificially generated noise.
 
@@
     \includegraphics{../figures/feat_importances.pdf}
-    \caption{Feature importance scores for both the original and noise-augmented datasets.
-    Orange markers represent the importance of original features, whereas blue markers indicate the importance of noise features.}
+    \caption{Feature importance scores for both the original and noise-augmented features.
+    Blue markers represent the importance of original features, while orange markers correspond to noise features.
+The horizontal threshold line marks the point where removing additional features starts degrading model performance.
+Features below this threshold are non-essential, as their removal improves the evaluation score—demonstrating that noise features (orange) contribute little to model performance.
+In contrast, features above the threshold are crucial, as removing them further degrades performance.}
     \label{fig:feature_importances}
@@
 era                             & 1.296625 & \textbf{1.249352} & 1.287840 & 1.277622 & 1.277424 \\
-cpmp2015                        & 384.377425 & 326.497515 & \textbf{285.105799} & 1211.124409 & 848.721320 \\
+cpmp2015                        & 1137.738259 & 997.261786 & \textbf{918.481147} & 1211.124409 & 848.721320 \\
 qsar234                         & 0.572577 & \textbf{0.415656} & 0.428724 & 0.428721 & 0.441020 \\

### 2025-02-06 reports/paper/vspyct.tex
fix in the STR table - OPCT best score on cpmp

@@
 era                             & 1.296625 & \textbf{1.249352} & 1.287840 & 1.277622 & 1.277424 \\
-cpmp2015                        & 1137.738259 & 997.261786 & \textbf{918.481147} & 1211.124409 & 848.721320 \\
+cpmp2015                        & 1137.738259 & 997.261786 & 918.481147 & 1211.124409 & \textbf{848.721320} \\
 qsar234                         & 0.572577 & \textbf{0.415656} & 0.428724 & 0.428721 & 0.441020 \\

### 2025-02-11 reports/paper/vspyct.tex
added loss threshold in appendix

@@
 Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
-\vanote{Ensemble methods, such as random forests and gradient-boosted trees, achieve high predictive performance by reducing variance and/or bias through the aggregation of diverse models.
-These techniques ensure robustness by capturing slightly different patterns or residuals from the data.}
+Ensemble methods, such as random forests and gradient-boosted trees, achieve high predictive performance by reducing variance and/or bias through the aggregation of diverse models~\cite{Breiman_2001, Kocev_2020, Friedman_2001}.
+These methods ensure robustness by capturing slightly different patterns or residuals from the data.
 While effective in improving accuracy, ensembles tend to obscure the model's interpretability—a fundamental attribute of decision trees—and do not inherently provide a means to quantify prediction uncertainty, which is crucial for informed decision-making across various applications.
@@
 \gls{vspyct} integrates the variational Bayes method~\citep{Blei_2017} to facilitate complex decision-making within a singular model framework, thus maintaining the interpretability inherent to decision trees without necessitating ensembles.
-\vanote{Unlike ensembles, which explicitly combine multiple \gls{spyct} trees, a single \gls{vspyct} attempts to incorporate uncertainty by sampling from the posterior distribution of tree parameters.}
+Unlike ensembles, which explicitly combine multiple \gls{spyct} trees, a single \gls{vspyct} attempts to incorporate uncertainty by sampling from the posterior distribution of tree parameters.
 Furthermore, by embedding model-specific uncertainty quantification directly into the decision tree structure through Bayesian inference, \gls{vspyct} enhances the depth of insight into the model's decision process and confidence levels.
@@
 Experimental evaluations on numerous benchmark datasets have shown that \glspl{spyct} achieve performance on par with state-of-the-art methods while being significantly faster than standard \glspl{pct}~\citep{Andonovikj_2024}.
-%Moreover, meaningful feature importance scores can be extracted from SPYCT models, enhancing their interpretability.
 
@@
 Our proposed \gls{vspyct} model addresses these critical challenges by integrating Bayesian principles directly into the predictive clustering tree framework.
-%Unlike traditional ensemble strategies, \glspl{vspyct} employs a probabilistic treatment of model parameters through the variational Bayes method, which enhances interpretability and provides inherent uncertainty quantification.
-This approach maintains the clarity and simplicity of a single tree model while achieving state-of-the-art performance, thus offering a significant improvement over existing methods.
+Such example where Bayesian principles, specifically the variational Bayes method, have been directly integrated in a predictive model can be found in~\cite{Bo_koski_2021, lopez2023informative}
+This approach maintains the clarity and simplicity of a single tree model while achieving on-par performance with the state-of-the-art methods.
 By introducing uncertainty quantification directly into the decision-making process, \gls{vspyct} enhances the reliability and applicability of the model across various domains where decision certainty is crucial.
-%This advancement bridges the gap between theoretical developments and practical applications, marking a significant step forward in making complex decision trees more accessible, interpretable, and effective.
 
 \section{The variational Bayes method}
-
-% optional slika - Optimisation process of finding the closest variational distribution qω(θ) over the set of latent variables ω. (da se generira vo python)
 
@@
 
-\vanote{While parameter uncertainty is effectively addressed, structural uncertainty is inherently more challenging to capture in a single tree.
-The variational Bayes approach approximates the posterior over parameters, but the inclusion of priors on tree structure (e.g., split selection) is an area for further exploration.}
+While parameter uncertainty is effectively addressed, structural uncertainty is inherently more challenging to capture in a single tree.
+The variational Bayes approach approximates the posterior over parameters, but the inclusion of priors on tree structure (e.g., split selection) is an area for further exploration.
 
 \begin{algorithm}[h!]
-\caption{Variational Learning of Split Parameters in \gls{vspyct} with Impurity Minimization}
+\caption{Variational Learning of Split Parameters in \gls{vspyct} with Impurity Minimisation}
 \label{alg:learn_split_vb}
@@
 
-\vanote{The posterior sampling allows the model to integrate over the uncertainty of the parameters at each node, providing smoothed and robust decision boundaries.
+The posterior sampling allows the model to integrate over the uncertainty of the parameters at each node, providing smoothed and robust decision boundaries.
 The model captures variability in the decision process, reducing the risk of overfitting to a single parameter set.
 While this approach effectively accounts for parameter uncertainty, the reliance on a single tree structure limits the model’s ability to replicate the diversity inherent in ensemble methods.
-Ensembles, achieve variance reduction by combining predictions from multiple distinct structures, a capability that a single \gls{vspyct} cannot fully emulate.}
+Ensembles, achieve variance reduction by combining predictions from multiple distinct structures, a capability that a single \gls{vspyct} cannot fully emulate.
 
@@
 where \(\epsilon\) is a small constant added for numerical stability.
-\pbnote{$\text{Var}(\mathbf{w}_s) > 0$, why do you need $\epsilon$?
-I know that we were talking about this, but there is a concept called coefficient of variation and it is defined as $\sigma/\mu$, std.
-(not variance) vs mean.
-We have to discuss this.}
 
@@
 In many real-world scenarios, the data may contain variables that do not contribute to the predictive task at hand -- hence, it is paramount to ensure that the model assigns low importance to such non-informative features.
-Accordingly, this subsection delves into whether \gls{vspyct} can distinguish relevant features from artificially generated noise.
+%Accordingly, this subsection delves into whether \gls{vspyct} can distinguish relevant features from artificially generated noise.
 
@@
 This heterogeneity is expected because many real-world datasets naturally include some highly informative predictors, as well as others that provide minimal predictive power.
-Importantly, the noise features typically receive lower importance values, confirming that \gls{vspyct} can detect their lack of relevance.
+
+To assess the significance of individual features in our model, we conducted an iterative feature removal process inspired by~\cite{guyon2003introduction}.
+We aim to identify the subset of features that meaningfully contribute to the model’s predictive performance while eliminating those that do not.
+Starting with the full set of features, we trained our model and recorded its evaluation score.
+Then, we systematically removed features in ascending order of their importance scores—starting with the least important feature and retraining the model after each removal.
+This iterative process continued until all features were removed.
+We set a threshold, which distinguishes important from non-important features, as the feature importance score at which the model performance begins to degrade.
+
+Importantly, the noise features in Figure~\ref{fig:feature_importances} typically receive lower importance values, confirming that \gls{vspyct} can detect their lack of relevance.
 Formally, if \(I(f)\) denotes the estimated importance of a feature \(f\), then
@@
 This outcome does not necessarily imply that the noise features carry genuine signal; rather, it highlights that certain original features are also uninformative for the target variable.
-Overall, these results confirm that the \gls{vspyct}-derived feature importance measures reliably rank truly predictive variables above most spurious ones, underscoring the model's robustness to irrelevant inputs.
+These results confirm that the \gls{vspyct}-derived feature importance scores reliably rank truly predictive variables above most spurious ones, underscoring the model's robustness to irrelevant inputs.
 
@@
 
-\vanote{Figure~\ref{fig:inter_tree} presents the visual representation of the \gls{vspyct} applied to the Unemployment dataset.
+Figure~\ref{fig:inter_tree} presents the visual representation of the \gls{vspyct} applied to the Unemployment dataset.
 It provides clear insights into how different feature groups influence predictions of unemployment duration.
 Each node represents a decision point, with the edges coloured based on the dominant feature group—\textit{ISCO} (occupational classifications), \textit{P16} (education-related features), or demographic features—used in the split.
-This colour-coded representation highlights the relative importance of feature groups across different levels of the tree.}
+This colour-coded representation highlights the relative importance of feature groups across different levels of the tree.
 
-\vanote{Demographic features dominate higher levels of the tree, indicating their broad and primary influence on unemployment duration.
+Demographic features dominate higher levels of the tree, indicating their broad and primary influence on unemployment duration.
 Features such as \textit{Months of work experience} and \textit{Age} frequently appear at the root or upper nodes, suggesting that prior professional experience and age are strong, general predictors of re-employment likelihood.
-For instance, individuals with less work experience are often routed towards branches with longer unemployment times, reflecting the challenges faced by those lacking experience in securing jobs.}
+For instance, individuals with less work experience are often routed towards branches with longer unemployment times, reflecting the challenges faced by those lacking experience in securing jobs.
 
-\vanote{In contrast, ISCO features become more prominent in lower-level splits, capturing occupation-specific influences on unemployment.
+In contrast, ISCO features become more prominent in lower-level splits, capturing occupation-specific influences on unemployment.
 Branches dominated by ISCO-related splits suggest that certain occupational groups have distinct re-employment trends.
-For example, occupations such as \textit{Health associate professionals} or \textit{Labourers in mining, construction, manufacturing, and transport} are likely associated with specific labor market demands that either shorten or extend unemployment durations.}
+For example, occupations such as \textit{Health associate professionals} or \textit{Labourers in mining, construction, manufacturing, and transport} are likely associated with specific labor market demands that either shorten or extend unemployment durations.
 
-\vanote{P16 features emerge in intermediate levels of the tree, reflecting the nuanced role of education in employment outcomes.
-For instance, splits involving education levels such as \textit{Business, administration, and law} or \textit{Services} suggest that individuals’ educational backgrounds influence their unemployment duration in ways that interact with occupational or demographic factors.}
+P16 features emerge in intermediate levels of the tree, reflecting the nuanced role of education in employment outcomes.
+For instance, splits involving education levels such as \textit{Business, administration, and law} or \textit{Services} suggest that individuals’ educational backgrounds influence their unemployment duration in ways that interact with occupational or demographic factors.
 
-\vanote{The leaf nodes, representing the predicted expected survival times (i.e., unemployment duration in days), provide a clear endpoint for each pathway.
-Branches leading to longer survival times often involve demographic splits, highlighting structural barriers such as insufficient work experience or higher age, while branches associated with shorter durations often incorporate ISCO and P16 splits, reflecting the effect of occupation- and education-specific factors.}
+The leaf nodes, representing the predicted expected survival times (i.e., unemployment duration in days), provide a clear endpoint for each pathway.
+Branches leading to longer survival times often involve demographic splits, highlighting structural barriers such as insufficient work experience or higher age, while branches associated with shorter durations often incorporate ISCO and P16 splits, reflecting the effect of occupation- and education-specific factors.
 
@@
 The model opens new possibilities for research in semi-supervised learning and Bayesian inference, promising to bridge the gap between theoretical advancements and practical applications.
-\vanote{Future work will focus on further refining the model's performance, exploring the inclusion of soft splits to enable smoother decision boundaries and probabilistic routing, and extending its applications in diverse predictive tasks.}
+Future work will focus on further refining the model's performance, exploring the inclusion of soft splits to enable smoother decision boundaries and probabilistic routing, and extending its applications in diverse predictive tasks.
 
@@
 
+\section{Evaluation score in relation to the number of removed features}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics{loss_rf2.pdf}
+    \caption{Dataset: rf2}
+    \label{fig:loss_rf2}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics{loss_cpm.pdf}
+    \caption{Dataset: cpmp2015}
+    \label{fig:loss_cpm}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics{loss_mfeat.pdf}
+    \caption{Dataset: mfeat\_zernike}
+    \label{fig:loss_mfeat}
+\end{figure}
+
+\begin{figure}[h!]
+    \centering
+    \includegraphics{loss_chronic.pdf}
+    \caption{Dataset: chronic\_kidney\_disease}
+    \label{fig:loss_chronic}
+\end{figure}
+
 \end{document}

### 2025-03-29 reports/paper/vspyct.tex
vspyct performance analysis through correlation

@@
 
+\subsection{Analysis of VSPYCT Performance under Varying Dataset Properties}
+
+In order to identify which types of datasets and predictive tasks \gls{vspyct} handles well or struggles with, we correlated each dataset’s final performance metric (MAE/NMAE or F1) against its size (\(N\)) and number of features (\(D\)), as summarised in Table~\ref{tab:vspyct_correlation}.
+
+\begin{table}[h!]
+\centering
+\caption{Spearman correlation between dataset size ($N$), number of features ($D$), and \gls{vspyct} performance (MAE/NMAE for regression and F1 for classification).}
+\label{tab:vspyct_correlation}
+\begin{tabular}{@{}lcccc@{}}
+\toprule
+\textbf{Task} & \multicolumn{2}{c}{\textbf{Corr.\ w.r.t. }$N$} & \multicolumn{2}{c}{\textbf{Corr.\ w.r.t. }$D$} \\
+\cmidrule(lr){2-3}\cmidrule(lr){4-5}
+ & $\rho$ & $p$ & $\rho$ & $p$ \\
+\midrule
+\textbf{STR} & $-0.345$ & $0.328$ & $-0.188$ & $0.602$\\
+\textbf{MTR} & $-0.231$ & $0.521$ & $0.068$ & $0.853$\\
+\textbf{BC}  & $0.286$  & $0.424$ & $-0.456$ & $0.185$\\
+\textbf{MCC} & $-0.152$ & $0.676$ & $-0.055$ & $0.881$\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+Using Spearman correlation to relate \gls{vspyct}’s scores to $N$ and $D$, we observed only weak or moderate (and statistically insignificant) relationships overall.
+For single-target regression, for instance, a small negative correlation with $N$ and $D$ indicates that increasing dataset size or dimensionality does not systematically improve or degrade \gls{vspyct}’s predictive error -- exemplified by its success on both large, high-dimensional tasks such as \emph{qsar234}, as well as on relatively smaller, simpler problems like \emph{quake}.
+In multi-target regression, similarly negligible correlations suggest that \gls{vspyct}’s performance does not critically depend on the sheer scale of the data; it produces low NMAE on cases such as \emph{scpf} (modest size and dimensionality) as well as \emph{rf1} and \emph{rf2} (substantial size and feature counts), revealing that its oblique splits can effectively model multiple continuous targets even when dimensionality is high.
+Turning to binary classification, where higher F1 scores are better, a slight positive correlation with $N$ implies that larger datasets might offer marginal gains, while a mild negative correlation with $D$ reflects that \gls{vspyct} can maintain robust performance despite high-dimensional problems (e.g., \emph{ova\_lung} with over $10\,000$ features).
+Finally, the multi-class classification results also show near-zero correlation with both $N$ and $D$, meaning \gls{vspyct} can handle diverse complexities without clear dependence on data size or feature count alone, as seen by strong F1 on both smaller sets (e.g., \emph{dermatology}) and larger ones (e.g., \emph{gas\_drift}).
+Across all four predictive tasks, the absence of any strong or statistically significant correlation with $N$ or $D$ underscores that \gls{vspyct}’s method of oblique, probabilistic splits adapts broadly: it is more successful in classification tasks, and less consistently dominant in single-target regression problems demanding extremely low errors (for example, on \emph{bike\_sharing} where ensembles tend to excel).
+Nonetheless, even in regression, \gls{vspyct} frequently remains competitive, especially on datasets with moderate variance, making it an appealing single-model alternative when interpretability and uncertainty quantification are valued.
+
 \subsection{Robustness to Spurious Features}%
@@
 
+\clearpage
 \section{Evaluation score in relation to the number of removed features}
+
+The following graphs show how the predictive performance of the VSPYCT model changes as features are iteratively removed, starting from the least important.
+
+Figure~\ref{fig:loss_rf2} illustrates the effect of iterative feature removal on the \textit{rf2} dataset.
 
@@
 
+Figure~\ref{fig:loss_cpm} shows the performance degradation on the \textit{cpmp2015} dataset.
+
 \begin{figure}[h!]
@@
 
+Figure~\ref{fig:loss_mfeat} depicts the model’s predictive performance as features are removed from the \textit{mfeat\_zernike} dataset.
+
 \begin{figure}[h!]
@@
 
+Figure~\ref{fig:loss_chronic} demonstrates the impact of feature removal on predictive accuracy for the \textit{chronic\_kidney\_disease} dataset.
+
 \begin{figure}[h!]

### 2025-03-29 correlation_analysis.py

@@ -1,0 +1,110 @@
+import numpy as np
+import pandas as pd
+from scipy.stats import spearmanr
+
+# 1. PREPARE DATA (N, D, VSPYCT performance) FOR EACH TASK
+
+# ---------- Single-Target Regression (STR) -------------
+# Datasets: (Name, N, D, VSPYCT_MAE)
+str_datasets = [
+    ("era", 1000, 5, 1.287840),
+    ("cpmp2015", 2108, 27, 918.481147),
+    ("qsar234", 2145, 1026, 0.428724),
+    ("concrete_compressive_strength", 1030, 9, 7.740371),
+    ("yprop", 8885, 252, 0.019969),
+    ("puma8NH", 8192, 8, 2.711678),
+    ("space_ga", 3107, 6, 0.104501),
+    ("bike_sharing", 17379, 6, 11.288880),
+    ("quake", 2178, 3, 0.143805),
+    ("ailerons", 13750, 40, 0.000127)
+]
+
+# ---------- Multi-Target Regression (MTR) -------------
+# Datasets: (Name, N, D, VSPYCT_NMAE)
+mtr_datasets = [
+    ("wq", 1060, 16, 0.183236),
+    ("atp1d", 337, 411, 0.125300),
+    ("atp7d", 296, 411, 0.167780),
+    ("scpf", 1137, 23, 0.027291),
+    ("osales", 639, 411, 0.057865),
+    ("sf1", 323, 10, 0.092266),
+    ("sf2", 1066, 10, 0.037199),
+    ("rf1", 9125, 64, 0.166669),
+    ("rf2", 9125, 576, 0.193914),
+    ("slump", 103, 7, 0.239199)
+]
+
+# ---------- Binary Classification (BC) -------------
+# Datasets: (Name, N, D, VSPYCT_F1)
+bc_datasets = [
+    ("scene", 2407, 299, 0.972439),
+    ("ova_breast", 1545, 10935, 0.959834),
+    ("bioresponse", 3751, 1777, 0.768986),
+    ("ova_lung", 1545, 10935, 0.920463),
+    ("chronic_kidney_disease", 400, 25, 1.000000),
+    ("compas_two_years", 5278, 13, 0.652540),
+    ("spambase", 4601, 57, 0.926491),
+    ("gina_agnostic", 3468, 970, 0.858579),
+    ("credit-g", 1000, 20, 0.681566),
+    ("diabetes", 768, 8, 0.715029)
+]
+
+# ---------- Multi-Class Classification (MCC) -------------
[…]

### 2025-06-19 reports/paper/response_to_reviewers.tex
First revision

\documentclass [a4paper,11pt]{article}
\usepackage{enumitem}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
%\usepackage{cite} 
%\usepackage[sort&compress,square,numbers]{natbib}
\usepackage[backend=biber,sorting=none]{biblatex}
\addbibresource{references.bib}
\usepackage{color}
\setlength{\itemsep}{0pt plus4pt minus2pt}
\renewcommand{\labelitemi}{\Alph{enumi}}
\setlength{\headheight}{0mm} \setlength{\topskip}{0mm}
\setlength{\textheight}{24cm} \setlength{\topmargin}{-15mm}
\setlength{\textwidth}{16cm}\setlength{\oddsidemargin}{0cm}\setlength{\evensidemargin}{0cm}
\setlength{\parindent}{0mm} \setlength{\parskip}{2mm}
\usepackage{url}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{eurosym}
\usepackage{enumitem}
\usepackage{graphicx}
\usepackage[colorinlistoftodos]{todonotes}
\usepackage{booktabs}
\usepackage[version=4]{mhchem}
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{comment}
\usepackage{bm}
\graphicspath{ {./}{./figs/}{./figs/laprint_figs/} }

\definecolor{SteelBlue}  {rgb}{0.275,0.51,0.706}

\usepackage{xr}
\externaldocument[m-]{main}

\newtheoremstyle{blue}% name
{10pt}%      Space above
{3pt}%      Space below
{\color{SteelBlue}}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{blue}
\newtheorem{question}{Comment}[section]

\definecolor{green-red}  {rgb}{0,0.,1}
\newtheoremstyle{blue}% name
{10pt}%      Space above
{3pt}%      Space below
{\color{green-red}}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{blue}
\newtheorem{general}{General Comment}[section]

\newtheoremstyle{note}% name
{3pt}%      Space above
{3pt}%      Space below
{}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{note}
\newtheorem{answer}{Response}[section]

\newtheoremstyle{note}% name
{3pt}%      Space above
{10pt}%      Space below
{}%         Body font
{}%         Indent amount (empty = no indent, \parindent = para indent)
{\bfseries}% Thm head font
{:}%        Punctuation after thm head
{.5em}%     Space after thm head: " " = normal interword space;
%       \newline = linebreak
{}%         Thm head spec (can be left empty, meaning `normal')

\theoremstyle{note}
\newtheorem{general_answer}{General Response}[section]

\newcommand{\todopb}[2][]
{\todo[backgroundcolor=blue!25, #1]{#2}}

\usepackage[normalem]{ulem}
\newcommand{\mkom}[1]{{\color{purple} #1}}
\newcommand{\mnew}[1]{{\color{Tan} #1}}
\newcommand{\mpop}[2]{{\color{orange}\sout{#1}\color{purple}{#2}}}
\newcommand{\mrev}[2]{{\color{red}\footnote{#1} #2 }}

\usepackage[nomargin,inline,marginclue,draft]{fixme}
 \fxsetup{theme=color, mode=multiuser} 
 \FXRegisterAuthor{pb}{apb}{\color{red}PB} 
 \FXRegisterAuthor{mp}{amp}{\color{red}mp}

\usepackage{xr}
\externaldocument[M-]{detcoupling}

\begin{document}

	\textbf{Detailed Response to Reviewers}\\[10pt]

%	{\large \textbf{Reply to the reviewers}}
%	This document contains the responses to the reviewers' questions.
%	In addition to the revised manuscript, we have added a marker revised manuscript that includes color-highlighted changes marked with a reference to the reviewers' questions.

%	The analysis of the mutual dependence of $(N,L)$ and the dependence on the parameters $k_1$ and $k_2$ are presented as Supplementary material.
%	We have also moved the original Appendix~A as a part of the Supplementary material.

	\section{Reviewer \#1:}

%{\color{blue}Section I)}

\begin{question}
Correct small typographical issues (e.g., replace "it's" with "its" in formal writing).
\end{question}

\begin{answer}

Thank you. This is now corrected.

\end{answer}

%{\color{blue}Section II)}

\begin{question}

Improve Figure 1's caption to clearly distinguish ensemble vs. variational trees.

\end{question}

\begin{answer}

The caption has been improved as per your suggestion.

\end{answer}

\begin{question}

Consider summarizing model characteristics in a comparative table in the Related Work section.

\end{question}

\begin{answer}

We have now included a comparative table in the Related Work section clearly summarising and comparing the key characteristics of the VSPYCT method and the other predictive clustering methods in the literature.

\end{answer}

\begin{question}

Clarify "structured output prediction" early for readers not familiar with the term.

\end{question}

\begin{answer}

We clarified the term Structured Output Prediction (SOP) explicitly in the Introduction section by providing a concise definition for readers who may not be familiar with this concept.

\end{answer}

\begin{question}

Include explanation of $\Omega_{obs} = \Omega (X,Y,w,b)/2$ mentioned in Algorithm1 and why this choice is made.

\end{question}

\begin{answer}

We have clarified this in the manuscript (inside Algorithm 1), explicitly stating that the factor of 1/2 in $\Omega_{obs} = \Omega (X,Y,w,b)/2$ is introduced to ensure that it is consistently lower than the current impurity $\Omega (X,Y,w,b)$. This guarantees that the optimisation is biased toward impurity reduction at each step, thereby encouraging effective split selection. The choice helps to drive the learning dynamics in the right direction by maintaining a useful margin between the observed and expected impurities.

\end{answer}

\begin{question}

Although the paper acknowledges the added complexity of variational inference compared to deterministic tree splits, it lacks empirical evidence regarding training time
or scalability.

\end{question}

\begin{answer}

Thank you for highlighting this important aspect. While we have not provided empirical training-time measurements due to their high dependence on specific implementations and hardware, we have explicitly derived and clearly presented the theoretical computational complexity of our proposed VSPYCT model compared to the deterministic SPYCT. The added complexity due to the variational inference framework is detailed explicitly in the manuscript. To clarify the limitations, we have now clearly stated in the manuscript that systematic empirical evaluations of training time and scalability were outside of the scope of this study, due to the above-mentioned reasons.

\end{answer}

\begin{question}

The model captures parameter uncertainty well, but structural uncertainty (e.g., tree topology) is briefly mentioned as future work.

\end{question}

\begin{answer}

Thank you for this important suggestion. The Discussion section now highlights structural uncertainty as a significant area for future research. Specifically, we mentioned potential strategies such as incorporating structural priors or employing ensemble-like approaches within the variational framework.

\end{answer}

\begin{question}

There is no mention of code availability, parameter tuning strategies, implementation frameworks or datasets . Consider including a statement on code or data
availability to enhance reproducibility.

\end{question}

\begin{answer}

We have now included a clear statement in the manuscript with a link to our code implementation.

\end{answer}

\begin{question}

While feature importance and visualisations are well described, a brief practical commentary on how domain experts might interact with or interpret the VSPYCT outputs
would be valuable.

\end{question}

\begin{answer}

We have explicitly added a brief practical commentary in the Interpretability subsection, clearly outlining how domain experts (e.g., clinicians, employment analysts, or material scientists) might effectively interact with and interpret VSPYCT outputs for informed decision-making in their respective fields.

\end{answer}

	\section{Reviewer \#2:}

\begin{question}

The novelty of the work should be clearly emphasized in the last paragraph of the introduction.

\end{question}

\begin{answer}

Thank you for this important remark. We have now explicitly emphasised the novelty of our approach in the final paragraph of the Introduction, clearly highlighting the integration of Bayesian inference with oblique splits, the direct quantification of uncertainty, and the unique capability to maintain interpretability while achieving competitive predictive performance.

\end{answer}

\begin{question}

Too many parameters have been examined in the study, but few results are presented in the Conclusion section. Please add a few points in the Conclusion section.

\end{question}

\begin{answer}

We have enhanced the Conclusion section with specific results summarizing the predictive performance of VSPYCT compared to benchmark methods. Additionally, we explicitly discussed key findings, limitations, and recommendations for future research clearly.

\end{answer}

\begin{question}

All parameters and symbols mentioned in this manuscript must be defined within the manuscript or in a separate list of nomenclature.

\end{question}

\begin{answer}

Thank you for this. All parameters and symbols used throughout the manuscript have now been explicitly defined at the point of introduction within the main text. We have ensured that no undefined notation remains, making the manuscript self-contained without the need for a separate nomenclature table.

\end{answer}

\begin{question}

The introduction is not well organized. The research gap and novelty of the research are missing. The author should clearly explain what has done previously and what is the research gap. The authors did not explain the cited articles properly.

\end{question}

\begin{answer}

Thank you for highlighting this issue. We reorganized the Introduction section by explicitly identifying the research gap—particularly emphasizing the lack of integrated uncertainty quantification and interpretability in previous studies. We have also clarified the context and relevance of the cited literature.

\end{answer}

\begin{question}

For the validity of the results, the study needs to be validated with previous studies graphically.

\end{question}

\begin{answer}

Thank you for highlighting this important point.
To clarify, our experimental design already includes comparisons with state-of-the-art models from previous literature, such as standard SPYCT, ensembles of SPYCTs, OPCT, and ensembles of PCTs.
These baseline models have been extensively evaluated and validated in previous published studies (Stepišnik and Kocev, 2021; Kocev et al., 2013; Stepisnik et al., 2020), and we explicitly included them in our empirical analyses. The ranking based on the predictive performance is given in Figure 2.

\end{answer}

\begin{question}

What is the novelty of this research?

\end{question}

\begin{answer}

The novelty of our research lies in the unique integration of variational Bayesian inference with oblique predictive clustering trees. This allows us to directly quantify uncertainty within a single interpretable tree model, achieving predictive performance comparable or superior to state-of-the-art ensemble methods, which has not been explored in previous studies.

\end{answer}

\begin{question}

The result and discussion section should be improved from the physical point of view.

\end{question}

\begin{answer}

We have introduced a new subsection "Practical Interpretation of Results" in the Results and Discussion section. This clearly addresses the physical and practical significance of our findings, explicitly linking the predictive outcomes to real-world contexts such as material properties, medical diagnoses, and employment data.

\end{answer}

\begin{question}

How were similarity parameters calculated?

\end{question}

\begin{answer}

We have explicitly clarified in the Methodology section how similarity parameters, specifically feature importance, are calculated. These parameters are computed based on the ratio of the expected values of the weights to their variances, which integrates both the magnitude and uncertainty of each feature's contribution to the predictions.

\end{answer}

\begin{question}

Include justifications for the choice of specific parameters and models.

\end{question}

\begin{answer}

We have clearly included justifications for our chosen parameters and models within the Methodology section. Specifically, we explained our choice of Gaussian priors for computational tractability, the selection of the ELBO optimisation approach, and parameters used in Monte Carlo sampling based on trade-offs between predictive accuracy and computational efficiency.

\end{answer}

\begin{question}

It is recommended to noticeably highlight the advantages, disadvantages, and limitations of the proposed approaches.

\end{question}

\begin{answer}

We have clearly added an explicit and structured discussion of the advantages, disadvantages, and limitations of the VSPYCT model within the Discussion subsection. Advantages include interpretability and uncertainty quantification, while disadvantages primarily involve computational complexity and potential scalability concerns.

\end{answer}

\begin{question}

It is necessary to incorporate further analysis and discussion comparing the current findings with previous publications. There must be an explanation for any differences between the present results and the data that has been published.

\end{question}

\begin{answer}

\end{answer}

\begin{question}

It is recommended to add a table representing all the investigated cases in the manuscript; please show the values/range of each tested parameter.

\end{question}

\begin{answer}

We have now added a comprehensive table under the "Experimental Setting" section, summarising all parameters investigated, their ranges, and their specific relevance. Furthermore, we explicitly stated the final selected parameter values used throughout the experiments clearly below the table.

\end{answer}

\begin{question}

The conclusion section should be better organized and it is recommended to include quantitative results. More specific aspects regarding the limitations, recommendations, and future works can be stated in the conclusion section. It is recommended to include qualitative and quantitative results.

\end{question}

\begin{answer}

We have reorganised and enhanced our Conclusion section to explicitly include both qualitative and quantitative summaries of our key findings.

\end{answer}

\begin{question}

Ensure that all sources and data are properly cited throughout the manuscript. For instance, the following papers could be helpful in improving the literature

\end{question}

\begin{answer}

Thank you for this suggestion. We have now cited the recommended references within our manuscript.

\end{answer}

\end{document}

### 2025-06-19 reports/paper/cover_letter.tex
First revision

\documentclass{article}

\usepackage{charter} % Use the Charter font for the document text
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{fancyhdr,graphicx}

\fancypagestyle{firstpage}{%
  \fancyhf{}% Clear header/footer
  \renewcommand{\headrulewidth}{0pt}%
  \renewcommand{\footrulewidth}{1pt}%
  % Add more detail here if needed
}
\fancypagestyle{otherpages}{%
  \fancyhf{}% Clear header/footer
  \renewcommand{\headrulewidth}{1pt}%
  \renewcommand{\footrulewidth}{1pt}%
  % Add more detail here if needed
}

\AtBeginDocument{\thispagestyle{firstpage}}
\pagestyle{otherpages}

\setlength{\parindent}{0pt}
\setlength{\parskip}{1ex}

\begin{document}

\vspace*{\dimexpr-\headsep-\headheight-1pt}

%\includegraphics[width=2.1in]{logo.png}

\rule{\linewidth}{1pt}

\bigskip\bigskip

\hfill
\begin{tabular}{ l @{} }
  \today \\[12pt] % Date
\end{tabular}

\bigskip

\begin{tabular}{ @{} l }
  Viktor Andonovikj \\ % Addressee of the letter above the to address
  Young researcher \\ % To address
  Jožef Stefan Institute \\
  Jamova cesta 39 \\
  1000 Ljubljana\\
  Slovenia
\end{tabular}

\bigskip

Dear Editor,

\bigskip

On behalf of my co-authors, I am submitting our paper titled ``Variational oblique predictive clustering trees''.

The paper presents a novel model that integrates Bayesian inference into oblique predictive clustering trees to enhance predictive performance, interpretability, and uncertainty quantification within a single-tree framework. Our results show that the proposed model achieves performance competitive with ensemble methods while retaining the transparency of interpretable models.

We believe that the paper provides a valuable contribution to the field, and we hope that it will be of interest to the readership of Expert Systems with Applications.

Thank you for considering our submission.

\bigskip

Sincerely yours,

\vspace{50pt}

Viktor Andonovikj

\end{document}

### 2025-06-19 reports/paper/credit_author_statement.tex
First revision

\documentclass{article}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage[T1]{fontenc}
\begin{document}
\begin{description}
\item[Viktor Andonovikj] Code writing, Manuscript writing, Data analysis, Modelling

\item[Sa\v{s}o D\v{z}eroski] Supervision, Concept validation, Reviewing and editing

\item[Biljana Mileva Boshkoska] Supervision, Conceptualisation, Data analysis, Reviewing and editing

\item[Pavle Bo\v{s}koski]  Supervision, Conceptualisation, Data provision, Data analysis, Reviewing and editing

\end{description}

\end{document}

### 2025-06-19 reports/paper/highlights_vspyct.tex
First revision

\documentclass{article}

\begin{document}
\begin{itemize}
\item Introduces VSPYCT, a variational Bayesian oblique decision tree for structured output prediction.
\item Captures parameter uncertainty through variational inference in each tree split.
\item Matches or outperforms SPYCT ensembles on classification and multi-target regression tasks.
\item Provides feature importance scores and visual interpretability within a single-tree framework.
\item Demonstrates robustness to spurious features and performs well across varied dataset properties.
\end{itemize}

\end{document}z

### 2025-06-19 reports/paper/markup_vspyct.tex
First revision

--- 2025-03-29 reports/paper/vspyct.tex
@@
 They are computationally efficient and when combined in ensembles they achieve state-of-the-art results.
-However, one major issue is that it's challenging to interpret an ensemble of \glspl{spyct} without the use of a model-agnostic method.
+However, one major issue is that it is challenging to interpret an ensemble of \glspl{spyct} without the use of a model-agnostic method.
 We propose variational oblique predictive clustering trees, which address this challenge.
@@
 \Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
+\added[id=r1, comment=C4]{In this work we use the term SOP to denote predictive tasks where each example’s label is itself a vector, set or hierarchy (e.g., multi-target regression, hierarchical multi-label, multi-output classification).}
 Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
@@
     \includegraphics[width=1.0\textwidth]{main_flowchart.pdf}
-    \caption{Ensemble of trees vs. Variational tree}
+    \caption{Visual contrast between a bagged ensemble of oblique predictive clustering trees (left) and the proposed single Variational SPYCT (right).
+The shading of hyper-planes illustrates parameter uncertainty that is captured by the variational posterior, rather than by averaging many independent trees.}
     \label{fig:main_flowchart}
@@
 This model avoids conventional ensemble strategies, which aggregate multiple trees' outputs, for a probabilistic treatment of model parameters, offering refined mechanisms for uncertainty quantification and predictive performance.
+
+\added[id=r2, comment=C4]{Despite the progress in predictive clustering methods, existing approaches either sacrifice interpretability by relying on ensembles or fail to directly incorporate uncertainty quantification.
+Recent works, such as \glspl{spyct}~\citep{Stepi_nik_2021} and \glspl{opct}~\citep{Stepisnik_2020}, have addressed some limitations but lack the integrated capability of uncertainty quantification, critical for real-world decision-making contexts.
+Hence, there is a clear research gap in developing methods that simultaneously ensure interpretability, accuracy, and robust uncertainty estimation.}
+
+\begin{comment}
 The core of \gls{vspyct}'s innovation lies in applying the variational Bayes method for optimising parameters defining the oblique splits, and modelling these parameters as random variables with probability distributions.
@@
 It sets a new benchmark for interpretable, efficient, and reliable machine learning models.
-\pbnote{Check Bayesian Additive Regression Trees (BART). Maybe add a comment/reference what is different}
+\end{comment}
+
+\added[id=r2, comment=C1C6]{In contrast to prior work, \glspl{vspyct} combine oblique splits from predictive clustering trees with a Bayesian inference framework using variational Bayes.
+Unlike traditional ensemble methods or single deterministic trees, \glspl{vspyct} explicitly model uncertainty in split parameters, providing inherent uncertainty quantification without sacrificing interpretability.
+This integration represents a significant novelty by directly embedding probabilistic modeling into oblique splits, ensuring a single interpretable model can achieve performance comparable to or surpassing state-of-the-art ensembles across various prediction tasks.}
+
+%\pbnote{Check Bayesian Additive Regression Trees (BART). Maybe add a comment/reference what is different}
 
@@
 
+\added[id=r1, comment=C8]{The complete implementation is available at \url{https://github.com/viksa98/variational_spyct}.}
+
 \section{Related work}
 
-%references for VB applications in ML and Survival analysis models. Cite pavles paper.
+\added[id=r2, comment=C14]{Advanced modeling methods, including fractional calculus and artificial neural networks, have been applied extensively across diverse domains for solving complex modeling and prediction problems~\cite{Abbas_2024,Abbas_2025,Ramzan_2025,Abbas_2025_2, Andonovikj_2022}.}
 
@@
 Methods like MAPLE and other interpretability frameworks strike a balance between accuracy and transparency, but they often fall short in handling high-dimensional, sparse data efficiently.
+
+\added[id=r1, comment=C3]{To provide a clearer overview of the related predictive clustering approaches discussed, we summarise and contrast their interpretability, capability to handle high-dimensional data, and uncertainty quantification properties in Table~\ref{tab:model_comparison}.}
+
+\begin{table}[htbp]
+\centering
+\caption{Comparative summary of predictive clustering methods}
+\label{tab:model_comparison}
+\resizebox{\textwidth}{!}{
+\begin{tabular}{lccc}
+\toprule
+\textbf{Model} & \textbf{Interpretability} & \textbf{Handles High-dimensional Data} & \textbf{Uncertainty Quantification}\\
+\midrule
+PCT~\citep{Kocev_2013} & High & Limited & No \\
+OPCT~\citep{Stepisnik_2020} & Moderate to High & Moderate & No \\
+SPYCT~\citep{Stepi_nik_2021} & Moderate & High & No \\
+Ensemble methods & Low & High & Partial (via variance) \\
+\textbf{VSPYCT (ours)} & \textbf{Moderate to High} & \textbf{High} & \textbf{Yes (inherent, Bayesian)} \\
+\bottomrule
+\end{tabular}
+}
+\end{table}
 
@@
 
+\added[id=r2, comment=C8C9]{We selected Gaussian priors for split parameters due to their computational tractability and effectiveness in Bayesian inference frameworks.
+The impurity function used for evaluating splits leverages the variance reduction weighted by selection probabilities, a common choice ensuring effective partitioning of data into homogeneous subsets.
+Feature importance is calculated using the normalised expected weight-to-variance ratio, reflecting both the strength and reliability of each feature's contribution.}
+
 While parameter uncertainty is effectively addressed, structural uncertainty is inherently more challenging to capture in a single tree.
@@
                 \]
-                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$
+                \State Set the observed impurity value: $\Omega_\text{obs} = \frac{\Omega(\mathbf{X}, \mathbf{Y}; \mathbf{w}, b)}{2}$ \Comment{Scaled to remain lower than current impurity and guide minimisation}
+
                 \State Condition on the observed impurity using:
@@
 By evaluating against a single \gls{spyct} tree, an ensemble of \glspl{spyct}, \gls{opct} tree, and an ensemble of \glspl{pct}, we aimed to assess how \gls{vspyct}'s predictive performance compares against a spectrum of models, from single decision trees to more complex ensembles and option trees.
+
+\subsection{Parameter Configuration Summary}
+\added[id=r2, comment=C12]{}
+
+Table~\ref{tab:parameter_ranges} summarises the parameter ranges tested in this study and their relevance to the optimisation of \gls{vspyct}.
+
+\begin{table}[htbp]
+\centering
+\begin{tabular}{lll}
+Parameter & Range tested & Description \\ \hline
+Monte Carlo samples ($M$) & 10 - 100 & Approximation of posterior distributions \\
+Variational iterations ($I_{vb}$) & 50 - 500 & Convergence of ELBO optimization \\
+Batch size ($\beta$) & 32 - 256 & Mini-batch size for optimization \\
+Learning rate ($\lambda$) & 0.001 - 0.01 & Step size for gradient updates \\
+\end{tabular}
+\caption{Investigated parameter configurations for VSPYCT.}
+\label{tab:parameter_ranges}
+\end{table}
+
+After evaluating performance and computational trade-offs, the final chosen parameter values were set to $M=100$, $I_{vb}=100$, $\beta=128$, and $\lambda=0.01$.
 
@@
 
+\added[id=r1, comment=C7]{While the present work marginalises only over split-parameter posteriors, future work will adopt Bayesian non-parametric priors over tree topologies to capture structural, as well as parameter, uncertainty.
+This would allow the model to reason not only about decision-boundary uncertainty, but also about the tree structure itself (i.e., how many splits to create and where to place them).
+Possible approaches include integrating structural priors, or developing ensemble-like variational methods that maintain a distribution over multiple plausible tree structures.
+Such extensions would provide a more comprehensive Bayesian treatment of decision trees and further enhance both predictive reliability and interpretability.}
+
+\added[id=r1, comment=C6]{An explicit empirical comparison of training time and scalability between \gls{vspyct} and deterministic approaches was beyond the scope of this study, due to implementation-specific and hardware-dependent factors.}
+
+\textit{Advantages of VSPYCT:}
+\added[id=r2, comment=C10]{}
+\begin{itemize}
+    \item Integrated interpretability and uncertainty quantification.
+    \item Competitive predictive performance compared to ensembles.
+    \item Effective handling of complex classification boundaries.
+\end{itemize}
+
+\textit{Disadvantages and Limitations:}
+\begin{itemize}
+    \item Higher computational complexity due to variational inference.
+    \item Potential scalability issues with extremely large or high-dimensional datasets.
+    \item Limited capability to represent structural uncertainty compared to ensembles.
+\end{itemize}
+
 \subsection{Analysis of VSPYCT Performance under Varying Dataset Properties}
@@
 
+\subsection{Practical Interpretation of Results}
+\added[id=r2, comment=C7]{}
+The empirical results presented earlier can be further understood by considering their practical implications and physical meaning in real-world contexts.
+For regression tasks, such as predicting concrete compressive strength, the improved accuracy of \gls{vspyct} indicates that probabilistic modeling of uncertainty in feature weights enables better handling of measurement variability or experimental errors common in physical systems.
+This modeling approach allows practitioners, for example civil engineers, to better anticipate the reliability of material properties predicted by the model.
+
+Similarly, classification results can be interpreted from a practical perspective by analysing the decision boundaries constructed by \gls{vspyct}.
+For instance, in medical diagnosis (such as chronic kidney disease), the high predictive performance reflects the model’s capacity to clearly delineate clinically meaningful groups, effectively capturing critical diagnostic thresholds that physicians rely on for clinical decision-making.
+
+Moreover, the results obtained for employment duration prediction offer a direct practical interpretation: demographic and occupational splits made by \gls{vspyct} align closely with known socio-economic factors affecting employment outcomes.
+Thus, the proposed model effectively incorporates and reveals practically meaningful dynamics from the data, enhancing both interpretability and applicability.
+
 \subsection{Interpretability}
@@
 
+\paragraph{Practical use for domain experts}
+
+\added[id=r1, comment=C9]{}
+After training, VSPYCT provides three complementary outputs that can be inspected with standard analysis tools:
+\begin{enumerate}
+  \item \textbf{Node summaries:} for every split we report the posterior mean, standard deviation, and $95\%$ credible interval of each coefficient, allowing analysts to verify which variables dominate within a specific branch.
+  \item \textbf{Tree file:} the complete tree is exported in a plain, machine-readable graph format so that users can open it in any network-visualisation environment and interactively traverse, fold, or unfold decision paths.
+  \item \textbf{Global relevance ranking:} the importance scores (see Figure~\ref{fig:feature_importance}) order the features by their expected effect on the prediction, helping experts prioritise factors for monitoring or intervention.
+\end{enumerate}
+
 \section{Conclusion}
 
+\begin{comment}
 We introduced the \gls{vspyct} model, an approach that combines the benefits of oblique splits, variational Bayes, and Bayesian inference within a single, interpretable predictive framework.
@@
 Future work will focus on further refining the model's performance, exploring the inclusion of soft splits to enable smoother decision boundaries and probabilistic routing, and extending its applications in diverse predictive tasks.
-
-\begin{comment}
+\end{comment}
+
+\added[id=r2, comment=C2C13]{}
+We introduced \glspl{vspyct}, a novel framework combining Bayesian inference with predictive clustering trees.
+\Glspl{vspyct} uniquely incorporate variational Bayes into oblique splits, providing built-in uncertainty quantification and preserving interpretability.
+
+Our experimental evaluations demonstrated that \glspl{vspyct} outperform ensemble methods in binary and multi-class classification tasks, while remaining competitive in multi-target and single-target regression scenarios.
+For example, \gls{vspyct} achieved the highest average rank in binary classification tasks and tied for best in multi-class classification, confirming its robustness and generalisation capabilities.
+
+The primary advantages of \gls{vspyct} include inherent interpretability, integrated uncertainty quantification, and competitive predictive performance without requiring ensembles.
+However, its main disadvantage is increased computational complexity due to variational inference optimisation and Monte Carlo sampling.
+
+Future research should explore methods to reduce computational complexity, integrate structural priors to handle structural uncertainty, and extend \gls{vspyct} to broader domains and larger datasets, improving scalability and general applicability.
+
 \section*{Acknowledgement}
-The authors acknowledge the research core funding No.\ P2-0001,  P2-0103 and P2-0383 financially supported by the Slovenian Research Agency.
-The authors also acknowledge the funding received from the European Union's Horizon 2020 research and innovation programme project HECAT under grant agreement No.\ 870702.
-\end{comment}
+The authors acknowledge the research core funding No.\ P2-0001, and V5-24020 financially supported by the Slovenian Research and Innovation Agency, and the Ministry of the Economy, Tourism and Sport of the Republic of Slovenia.
 
@@
 qsar234                         & 0.572577 & \textbf{0.415656} & 0.428724 & 0.428721 & 0.441020 \\
-puma8NH                         & 2.795463 & \textbf{2.632242} & 2.711678 & - & - \\
+puma8NH                         & 2.795463 & \textbf{2.632242} & 2.711678 & 2.734472 & 2.769841 \\
 yprop                           & 0.020126 & 0.019693 & 0.019969 & \textbf{0.019682} & 0.023687 \\

### 2025-06-19 src/models/model.py

@@ -11,7 +11,6 @@ def nanvar(tensor, dim=None, keepdim=False):
     output = (tensor - tensor_mean).square().nanmean(dim=dim, keepdim=keepdim)
     return output
 
-# Adjusted function to handle multiple target variables
 def multi_target_impurity(tensor, target_dim=1):
     """
     Compute impurity for multiple target variables.
@@ -19,14 +18,11 @@ def multi_target_impurity(tensor, target_dim=1):
     tensor: Input tensor where each row represents a sample and each column represents a different target variable.
     target_dim: The dimension along which the targets are stored (e.g., 1 if columns are different targets).
     """
-    # Remove NaN values and check for empty tensor
     if tensor.isnan().all():
         return float('inf')
     
-    # Compute variances for each target variable (along target_dim)
     variances = nanvar(tensor, dim=0 if target_dim == 1 else 1)
     
-    # Return the sum of variances across all target variables
     total_variance = torch.sum(variances)
 
     return total_variance

### 2025-07-17 reports/paper/vspyct_paper.tex
latest changes

--- 2025-06-19 reports/paper/markup_vspyct.tex
@@
 \Glspl{spyct}, in particular, have pioneered the use of oblique splits for \gls{sop}, allowing for more intricate decision boundaries beyond the capabilities of traditional axis-parallel splits.
-\added[id=r1, comment=C4]{In this work we use the term SOP to denote predictive tasks where each example’s label is itself a vector, set or hierarchy (e.g., multi-target regression, hierarchical multi-label, multi-output classification).}
+In this work we use the term SOP to denote predictive tasks where each example’s label is itself a vector, set or hierarchy (e.g., multi-target regression, hierarchical multi-label, multi-output classification).
 Despite their advantages, a significant challenge with \glspl{spyct} is their reliance on ensemble methods for optimal predictive performance.
@@
 
-\added[id=r2, comment=C4]{Despite the progress in predictive clustering methods, existing approaches either sacrifice interpretability by relying on ensembles or fail to directly incorporate uncertainty quantification.
+Despite the progress in predictive clustering methods, existing approaches either sacrifice interpretability by relying on ensembles or fail to directly incorporate uncertainty quantification.
 Recent works, such as \glspl{spyct}~\citep{Stepi_nik_2021} and \glspl{opct}~\citep{Stepisnik_2020}, have addressed some limitations but lack the integrated capability of uncertainty quantification, critical for real-world decision-making contexts.
-Hence, there is a clear research gap in developing methods that simultaneously ensure interpretability, accuracy, and robust uncertainty estimation.}
+Hence, there is a clear research gap in developing methods that simultaneously ensure interpretability, accuracy, and robust uncertainty estimation.
 
@@
 
-\added[id=r2, comment=C1C6]{In contrast to prior work, \glspl{vspyct} combine oblique splits from predictive clustering trees with a Bayesian inference framework using variational Bayes.
+In contrast to prior work, \glspl{vspyct} combine oblique splits from predictive clustering trees with a Bayesian inference framework using variational Bayes.
 Unlike traditional ensemble methods or single deterministic trees, \glspl{vspyct} explicitly model uncertainty in split parameters, providing inherent uncertainty quantification without sacrificing interpretability.
-This integration represents a significant novelty by directly embedding probabilistic modeling into oblique splits, ensuring a single interpretable model can achieve performance comparable to or surpassing state-of-the-art ensembles across various prediction tasks.}
+This integration represents a significant novelty by directly embedding probabilistic modeling into oblique splits, ensuring a single interpretable model can achieve performance comparable to or surpassing state-of-the-art ensembles across various prediction tasks.
 
@@
 
-\added[id=r1, comment=C8]{The complete implementation is available at \url{https://github.com/viksa98/variational_spyct}.}
+The complete implementation is available at \url{https://github.com/viksa98/variational_spyct}.
 
@@
 
-\added[id=r2, comment=C14]{Advanced modeling methods, including fractional calculus and artificial neural networks, have been applied extensively across diverse domains for solving complex modeling and prediction problems~\cite{Abbas_2024,Abbas_2025,Ramzan_2025,Abbas_2025_2, Andonovikj_2022}.}
+Advanced modeling methods, including fractional calculus and artificial neural networks, have been applied extensively across diverse domains for solving complex modeling and prediction problems~\cite{Abbas_2024,Abbas_2025,Ramzan_2025,Abbas_2025_2, Andonovikj_2022}.
 
@@
 
-\added[id=r1, comment=C3]{To provide a clearer overview of the related predictive clustering approaches discussed, we summarise and contrast their interpretability, capability to handle high-dimensional data, and uncertainty quantification properties in Table~\ref{tab:model_comparison}.}
+To provide a clearer overview of the related predictive clustering approaches discussed, we summarise and contrast their interpretability, capability to handle high-dimensional data, and uncertainty quantification properties in Table~\ref{tab:model_comparison}.
 
@@
 
-\added[id=r2, comment=C8C9]{We selected Gaussian priors for split parameters due to their computational tractability and effectiveness in Bayesian inference frameworks.
+We selected Gaussian priors for split parameters due to their computational tractability and effectiveness in Bayesian inference frameworks.
 The impurity function used for evaluating splits leverages the variance reduction weighted by selection probabilities, a common choice ensuring effective partitioning of data into homogeneous subsets.
-Feature importance is calculated using the normalised expected weight-to-variance ratio, reflecting both the strength and reliability of each feature's contribution.}
+Feature importance is calculated using the normalised expected weight-to-variance ratio, reflecting both the strength and reliability of each feature's contribution.
 
@@
 \subsection{Parameter Configuration Summary}
-\added[id=r2, comment=C12]{}
 
@@
 
-\added[id=r1, comment=C7]{While the present work marginalises only over split-parameter posteriors, future work will adopt Bayesian non-parametric priors over tree topologies to capture structural, as well as parameter, uncertainty.
+While the present work marginalises only over split-parameter posteriors, future work will adopt Bayesian non-parametric priors over tree topologies to capture structural, as well as parameter, uncertainty.
 This would allow the model to reason not only about decision-boundary uncertainty, but also about the tree structure itself (i.e., how many splits to create and where to place them).
 Possible approaches include integrating structural priors, or developing ensemble-like variational methods that maintain a distribution over multiple plausible tree structures.
-Such extensions would provide a more comprehensive Bayesian treatment of decision trees and further enhance both predictive reliability and interpretability.}
-
-\added[id=r1, comment=C6]{An explicit empirical comparison of training time and scalability between \gls{vspyct} and deterministic approaches was beyond the scope of this study, due to implementation-specific and hardware-dependent factors.}
+Such extensions would provide a more comprehensive Bayesian treatment of decision trees and further enhance both predictive reliability and interpretability.
+
+An explicit empirical comparison of training time and scalability between \gls{vspyct} and deterministic approaches was beyond the scope of this study, due to implementation-specific and hardware-dependent factors.
 
 \textit{Advantages of VSPYCT:}
-\added[id=r2, comment=C10]{}
 \begin{itemize}
@@
     \centering
-    \includegraphics{../figures/feat_importances.pdf}
+    \includegraphics{feat_importances.pdf}
     \caption{Feature importance scores for both the original and noise-augmented features.
@@
 \subsection{Practical Interpretation of Results}
-\added[id=r2, comment=C7]{}
 The empirical results presented earlier can be further understood by considering their practical implications and physical meaning in real-world contexts.
@@
 
-\added[id=r1, comment=C9]{}
 After training, VSPYCT provides three complementary outputs that can be inspected with standard analysis tools:
@@
 
-\added[id=r2, comment=C2C13]{}
 We introduced \glspl{vspyct}, a novel framework combining Bayesian inference with predictive clustering trees.
